\chapter{Parts of speech}
\label{sec:POS}

\todo{replace \hfill in this section to right align text in first line of examples}

In this chapter, I describe the parts of speech in Gyeli, also referred to as word classes. The presentation of Gyeli's parts of speech system relies on a grammatical rather than semantic classification into categories. Following \citet[1-2]{schachter2007}, I consider grammatical properties such as ``the word’s distribution, its range of syntactic functions, and the morphological or syntactic categories for which it is specifiable'' as determining criteria for parts of speech classification. 

I generally distinguish lexical and grammatical word classes as well as open and closed classes.
Gyeli has only two open word classes, namely the lexical classes of nouns and verbs. The lexical parts of speech of adjectives and adverbs, which are open classes in many other languages, are closed classes in Gyeli given their limited number of members. The semantic functions that they carry in languages with large adjective and adverb classes are taken over by nouns. In addition to these typical lexical word classes, Gyeli also has a lexical, closed class of ideophones. 

The frequency of lexical word classes' occurrence in the Gyeli corpus is displayed in Table \ref{Tab:POSnolex}. Lexical words constitute 46.9\% of the words in the corpus.\footnote{As described in \sectref{sec:Data}, comprises 3304 words in total. For the distribution of word class frequencies, only 3133 word were taken into account, omitting e.g.\ code-switching and repetitions.} Out of these lexical words, 87.8\% constitute open class words, namely nouns and verbs. Among the closed lexical word classes, adverbs are the most frequent ones in the corpus, followed by ideophones and finally adjectives.

\todo[14]{indent the non-bold numbers and percentages so that it's clearer that they constitute the sum of the bold numbers}
\begin{table} 
\centering
\begin{tabular}{ll|ll}
 \midrule
  \multicolumn{2}{l}{Word class} &  \multicolumn{2}{l}{Frequency}  \\ %0 in SG: 51,  in PL: 21
 \midrule
  {\bfseries Open} &  & {\bfseries 1289}  & {\bfseries (87.8\%)} \\
  & Nouns  & 630  &  (48.9\%) \\
 &   Verbs  &  659  &  (51.1\%)  \\ % really 659 before 568
 \midrule
  {\bfseries Closed} &  & {\bfseries 179} & {\bfseries (12.2\%)} \\
 & Adjectives &  9 &  (5\%)   \\
  & Adverbs   & 150 &   (83.8\%)   \\ % 2x 3/6, 2x 3/0
 & Ideophones    &  20  &  (11.2\%) \\ 
 \midrule
Total & & 1468 & (100\%) \\
 \midrule
\end{tabular}
\caption{Frequency of lexical word classes}
\label{Tab:POSnolex}
\end{table}


In comparison, grammatical words constitute more than half of the corpus with 53.1\%. Their various sub-categories are summarized in Table \ref{Tab:POSnogram}.\footnote{Anaphoric markers can occur both as free morphemes and suffixes that attach to demonstratives. The anaphoric markers in Table \ref{Tab:POSnogram} under `Noun adjuncts' only count the free morphemes; bound anaphoric markers are discussed in \sectref{sec:ANASuff}.} Following \citet{schachter2007} with slight modifications,\footnote{Differences between \citet{schachter2007} and my parts of speech classification concern the subclasses of major categories. While the authors (p.\ 35) only subsume role markers, quantifiers, classifiers, and articles under noun adjuncts, I basically treat every grammatical word class as an element of the noun phrase that can appear in the noun phrase.} I distinguish pronouns, other pro-forms, elements of the noun phrase, elements of the verb phrase, adpositions, conjunctions, and other minor word classes in Gyeli, each of which has some subclasses. Elements of the verb phrase constitute the most frequent grammatical word category with 33.3\%. Within this category, the subject-tense-aspect-mood-polarity (STAMP) marker is the most common with 430 occurrences.

\iffalse

%??? table too large for page
\begin{table} 
\centering
\small
\begin{tabular}{ll|ll}
 \midrule
  \multicolumn{2}{l}{Word class} &  \multicolumn{2}{l}{Frequency}  \\ %0 in SG: 51,  in PL: 21
 \midrule
 {\bfseries Pronouns}  &  & {\bfseries 272} & {\bfseries (15.3\%)} \\ % plus possessives 58
  & Subject pronouns  & 73 & (4.1\%) \\
  & Non-subject pronouns  & 122 & (6.8\%) \\
  & Interrogative pronouns  &  11 & (.6\%) \\
  & Possessor pronouns  & 58 & (3.2\%) \\
  & Reflexive pronoun {\itshape mɛ́dɛ́}  & 8 & (.4\%) \\
 {\bfseries Other pro-forms} &   & {\bfseries 506} & {\bfseries (28.4\%)} \\
  & STAMP marker  & 430 & (24.1\%) \\
  & Interrogative pro-forms  & 32  & (1.8\%) \\
  & Pro-adverb & 33 & (1.8\%) \\
  & Pro-clause &  5 & (.3\%) \\
  & Pro-sentence & 6 & (.3\%) \\
 {\bfseries Noun adjuncts} &  & {\bfseries 312} & {\bfseries (17.5\%)} \\
 & Limiting modifiers  &  15 & (.8\%) \\
 & Demonstratives  & 53  & (3\%) \\
 & Anaphoric marker (free)  & 2 & (.1\%) \\ % 11 anaphoric markers are bound
 & Agreeing quantifiers   & 36 & (2\%) \\
 & Invariable quantifiers   & 6 & (.3\%) \\
 & Attributive/genitive markers   & 114  & (6.4\%) \\
 & Plural markers  & 11 & (.6\%) \\
 & Adpositions   & 75 & (4.2\%) \\
 {\bfseries Verb adjuncts} &  & {\bfseries 206} & {\bfseries (11.6\%)} \\
 & Auxiliaries   &  176 & (9.9\%) \\
 & Verbal particles   & 30 & (1.7\%) \\
 {\bfseries Conjunctions} &  & {\bfseries 284} & {\bfseries (15.9\%)} \\
 & Coordinators   & 60 & (3.4\%) \\
 & Complementizer  & 109 & (6.1\%) \\
& Comitative marker  & 62  & (3.5\%) \\ % nee, is 58 und 41 CONJ
& Adverbializers  & 14 & (.8\%) \\ % plus ka if
  & Discourse structuring {\itshape yɔ́ɔ̀}  & 39 & (2.2\%) \\
 {\bfseries Others} &  & {\bfseries 200} & {\bfseries (11.2\%)} \\
 %& Similative marker  & 4 & (.2\%) \\
 & Copulas  & 70 & (3.9\%) \\
% & Infinitives  & 37 & (2\%) \\
 & Sentential modifiers  & 57 & (3.2\%) \\
 & Interjections  & 31 & (1.7\%) \\
 & Exclamations  & 42 & (2.3\%) \\
 \midrule
  Total & & 1780 & (100\%) \\
 \midrule
\end{tabular}
\caption{Frequency of grammatical word classes}
\label{Tab:POSnogram}
\end{table}

\fi

\todo[96]{indent the non-bold numbers and percentages so that it's clearer that they constitute the sum of the bold numbers}
\begin{table} 
\centering
\small
\begin{tabular}{ll|ll}
 \midrule
  \multicolumn{2}{l}{Word class} &  \multicolumn{2}{l}{Frequency}  \\ %0 in SG: 51,  in PL: 21
 \midrule
 {\bfseries Pronouns}  &  & {\bfseries 240} & {\bfseries (14.4\%)} \\ 
  & Subject pronouns  &   61 &    (25.4\%) \\ %before 73
  & Non-subject pronouns  &   103 &   (42.9\%) \\ %before 122
  & Interrogative pronouns  &    10 &   (4.2\%) \\
  & Possessor pronouns  &   59 &   (24.6\%) \\
  & Reflexive pronoun {\itshape mɛ́dɛ́}  &   7 &   (2.9\%) \\
 {\bfseries Other pro-forms} &   & {\bfseries 63} & {\bfseries (3.8\%)} \\
  & Interrogative pro-forms  &   19  &   (30.2\%) \\ %before 32
  & Pro-adverb &   33 &   (52.4\%) \\
  & Pro-clause &    5 &   (7.9\%) \\
  & Pro-sentence &   6 &   (9.5\%) \\
 {\bfseries Elements of the noun phrase} &  & {\bfseries 233} & {\bfseries (14\%)} \\
 & with agreement prefix  &    43 &   (18.5\%) \\
 & with plural agreement only  &   5  &   (2.1\%) \\
 & with agreeing free morpheme  &   178 &    (76.4\%) \\ 
 & prenominal invariable modifiers  &   0 &    \\
 & postnominal invariable modifiers   &   7 &   (3\%) \\
 {\bfseries Elements of verbal complex} &  & {\bfseries 555} & {\bfseries (33.3\%)} \\
  & STAMP marker  &   430 &   (77.5\%) \\
 & Auxiliaries   &   75  &   (13.5\%) \\ % only counted true, semi go with verbs % before 176
 & Verbal particles   &   50 &   (9\%) \\
 {\bfseries Adpositions} &  & {\bfseries 156} & {\bfseries (9.4\%)} \\
 & Prepositions  &   120  &   (76.9\%) \\
 & Postpositions   &   36 &   (23.1\%) \\
 {\bfseries Conjunctions} &  & {\bfseries 180} & {\bfseries (10.8\%)} \\
 & Coordinators   &   56 &   (31.1\%) \\
 & Subordinators   &   124 &   (68.9\%) \\
 {\bfseries Other minor classes} &  & {\bfseries 238} & {\bfseries (14.3\%)} \\
 & Copulas  &   55 &   (23.1\%) \\
 & Identificational marker  &    13 &   (5.5\%) \\
  & Discourse structuring {\itshape yɔ́ɔ̀}  &   39 &   (16.4\%) \\
 & Question markers  &    1 &   (.4\%) \\
% & Infinitives  & 37 & (2\%) \\
 & Sentential modifiers  &   57 &   (23.9\%) \\
 & Extrasentential modifiers  &    73 &   (30.7\%) \\
 \midrule
  Total & & 1665 & (100\%) \\ %116 less than before
 \midrule
\end{tabular}
\caption{Frequency of grammatical word classes}
\label{Tab:POSnogram}
\end{table}

%& Similative marker  ADD TO POS prepositions CHAPTER AND verbs with prep

%Pronouns and other pro-forms constitute over 45\% of grammatical words in the corpus. Among them, the subject-tense-aspect-mood-polarity (STAMP) marker, a clitic encoding subject agreement and grammatical information within the clause, such as tense, aspect, and negation, is the most common with 430 occurrences in the corpus. In comparison, noun adjuncts, i.e.\ grammatical parts of speech that can occur in the noun phrase, are comparatively rare. Non of them attain more than 6.4\% (attributive/genitive markers); the total of noun adjuncts amounts to 17.5\% of all word classes in the corpus. Verb adjuncts, namely auxiliaries and verbal particles, are less frequent than noun adjuncts, which might be reflected in the general noun/verb ratio with nouns being more frequent than verbs. Even conjunctions such as coordinators, complementizers, comitative markers, adverbializers, and discourse elements, are slightly more frequent than verb adjuncts.


With regard to open versus closed word classes, the latter are significantly more frequent in Gyeli. All grammatical parts of speech categories presented in Table \ref{Tab:POSnogram} 
are closed classes. In addition, also the lexical classes of adjectives, adverbs, and ideophones belong to the closed word classes, as explained above. Thus, closed classes constitute 58.9\% (1844 in total numbers) of the 3133 word corpus.
The relative dominance of closed word classes in Gyeli is remarkable since it correlates with a morphological type of language that is closer to the analytic end of the analytic---synthetic scale. As \citet[23]{schachter2007} point out,
\begin{quote} ``closed word classes tend to play a more prominent role in analytic languages than they do in synthetic languages. This is because much of the semantic and syntactic work done by the members of closed word classes in analytic languages is done instead by affixes in synthetic languages.''
\end{quote} 
Gyeli words maximally contain three segmental morphemes with the possibility to host additional tonal morphemes. This is discussed in detail in \chapref{sec:WordForm}. The restriction on word length is, however, not morphological in nature, but phonological, as outlined in \sectref{sec:SyllDist}, generally permitting only up to three syllables in a word.

I will describe each part of speech in the remainder of this chapter, providing defining properties for each category. I start with the open word classes of nouns and verbs, giving information on selected subclasses, for instance the mass/count distinction in nouns. I then proceed with the other lexical classes of adjectives, adverbs, and ideophones before discussing grammatical classes.







	
	

\section{Nouns}
\label{sec:N}


There has been much discussion in the literature as to what a {\itshape noun} is, a linguistic term that is often used intuitively. \citet[10]{rijkhoff2002} maintains that ``there is still no general consensus among typologists on what constitutes a noun''. There is not even an unanimous agreement as to whether every language has a noun category. \citet{gil2013} claims, for instance, that Riau Indonesian does not have a noun (nor a verb) word class. \citet[12]{rijkhoff2002} distinguishes between i) languages without a major word class of nouns, ii) languages where nouns cannot be distinguished from other word classes, and iii) those languages that do have a distinct noun word class. \citet[5]{schachter2007}, on the other hand, hold that ``[t]he distinction between nouns and verbs is one of the few apparently universal parts-of-speech distinctions.'' They further explain that alleged examples of languages which would fall in category i) or ii) according to Rijkhoff had been based on incomplete data and therefore cannot be considered as counter-examples against this universal word class distinction. In any case, scholars seem to agree that at least most languages of the world have nouns as a distinct word class (Koptjevskaja-Tamm 2006: 720). 

According to \citet[708]{evans2000}, linguists usually define nouns by three different types of criteria, namely semantic, morphological, and syntactic.
In terms of semantics, a common definition is given by \citet[5]{schachter2007} who consider nouns a ``class of words in which occur the names of most persons, places, and things''. (Similar definitions are provided by other authors, for example by \citet[720]{tamm2006} and \citet[710]{evans2000}.) All these scholars emphasize, however, that this is a traditional definition of convenience, but that membership of a word in a certain part of speech has to be established on other grounds. There may be nouns that refer to other entities than persons, places or things, while, on the other hand, there may be persons, places or things that belong to some other word class than nouns. 

Another way of viewing nouns is to distinguish them from other open word classes such as verbs, adverbs, and adjectives on the basis of different morphosyntactic properties  (see, for instance, \citet{bhat2000} and \citet{baker2003}). The advantage of this approach is that it emphasizes the specific structures within a parts-of-speech system of a given language rather than over-generalizing across languages. Nouns may be inflected for categories such as number, case, possession, and definiteness (Koptjevskaja-Tamm 2006: 722). They may trigger agreement of these categories as heads of a noun phrase. Syntactically, they may take a certain position within a noun phrase that serves as an argument or adjunct, while dependent word classes are arranged in specific ways around them.

As \citet[733]{lehmann2000}  put it concisely, ``Like any other grammatical category, the word class `noun' has no universal status {\itshape a priori}; rather, it is a language-specific category.'' I will discuss noun properties in Gyeli in detail in the following section. This will help to distinguish nouns from other parts of speech as well as to establish subcategories of nouns that share some nominal features, but not all of them.


\subsection{Noun properties}
\label{sec:Nprop}

I define Gyeli nouns most broadly by their structure, function, and distribution in a phrase, distinguishing them from other word classes. Structurally, nouns consist minimally of a stem and, depending on the noun type, can take noun class prefixes as well as similative and object linking H tone prefixes, as outlined in \sectref{sec:Prefix}. This sets them apart from verbs which cannot take prefixes. While agreement targets of nouns also consist of a stem plus prefix, agreement targets can only take one prefix and that prefix generally differs in its form from noun class prefixes.

On the clause level, most nouns in Gyeli serve as subjects, objects, and adjuncts, as discussed in detail in \sectref{sec:verbalC}, as well as copular complements, as outlined in \sectref{sec:COP}. Nominalized past participles are an exception to this and can only occur as nominal predicates in copula constructions. All nouns can generally occur as bare nouns in their positions.

On the phrase level, nouns function as the head of the construction where they appear in initial position, followed by both agreeing and invariable modifiers, as outlined in \chapref{sec:NP}. In more complex noun phrases such as attributive constructions, the first constituent is always a noun, followed by an attributive or genitive marker and then containing another word, e.g.\ a noun or verb (see \sectref{sec:CONC} for more information on attributive constructions).
With respect to their morphosyntactic behavior, nouns have a grammatical gender and trigger agreement at their agreement targets (see \sectref{sec:Gender}).\footnote{I view agreement phenomena as a major reason to posit the noun as the lexical head of the phrase rather than assuming a (covert) functional head. The noun as the agreement trigger determines the morphological shape of all agreement targets, including demonstratives that could serve as potential determiner heads.}


Phonologically, nouns allow syllabic and tonological patterns that are disallowed in verbs. For instance, noun root onsets may be complex with clusters of up to three consonants while this pattern is not found in verbs. Also, diphthongs can be found in monosyllabic noun stems and rarely in first and second syllables of bisyllabic nouns. In contrast, they are always restricted to monosyllabic stems in verbs. For more information, see \sectref{sec:SyllDist}.  
Tonologically, nouns show a greater variety of patterns, allowing, for instance, H tones on second and third TBUs. Verbs, however, have underlyingly toneless TBUs in second and third syllables which surface as L tones in isolation, as explained in \sectref{sec:Tinventory}.


\subsection{Noun types}
\label{sec:Ntyp}

Gyeli nouns do not constitute a unified class. Instead, they have further subclasses which show different morphosyntactic behavior. This is nothing unusual from a typological perspective; as \citet[8]{schachter2007} point out:
\begin{quote} ``In most languages some grammatical distinction is made between common nouns, which are used to refer to any member of a class of persons, etc.\ (e.g.\ girl, city, novel), and proper nouns, which are used to refer to specific persons, etc.\ (e.g.\ Mary, Boston, Ivanhoe).'' \end{quote}

\noindent Gyeli has three types of nouns: common nouns, proper nouns, and nominalized past participles. I discuss them one by one in the following sections.





\subsubsection{Common nouns}
\label{sec:commonN}

Common nouns differ from other noun types  in their morpho-phonological structure as well as their morphosyntactic behavior.
Structurally, common nouns in Gyeli consist minimally of a nominal stem and add maximally up to three prefixes, the first of which is tonal, as shown in the template in (\ref{Nstruc}). The different prefix types are described in \sectref{sec:Prefix}.\footnote{Further information as well as an explanation of terminological distinctions of `noun class', `agreement class' and `gender' are provided in \sectref{sec:Gender}.}

\begin{exe}
\ex \label{Nstruc}  object linking H tone - noun class - similative marker - stem
\end{exe}

Common nouns can thus take a larger variety and number of prefixes compared with other noun types: proper nouns can only take a similative prefix, as described in \sectref{sec:properN} and nominalized past participles can only take a nasal noun class prefix, as described in \sectref{sec:NounPart}.

Another difference between common and other nouns is the former's potential for number inflection. While most common nouns (with the exception of uncountable nouns) have a singular and plural counterpart, as reflected by their pairing of different agreement and noun classes, proper nouns and nominalized past participles do not inflect for number.

On a phrasal level, common and proper nouns differ as well. In nominal possessive constructions, common possessor nouns  require an attributive marker, as discussed in \sectref{sec:CONC}. In contrast, proper nouns take a distinct genitive marker instead, as described in \sectref{sec:GEN}. Nominalized past participles do not occur in possessive constructions.


In summary, a set of tests help to reliably identify whether a word is a common noun or not. A Gyeli common noun can:

\begin{enumerate}
\itshapeem serve as the subject of a clause
\itshapeem serve as  the  first constituent of a N + N construction
\itshapeem be modified by an agreeing demonstrative or possessive pronoun
\itshapeem possibly make a number distinction (even though not all nouns do so) 
\end{enumerate}

\noindent I discuss the number distinction in more detail in \sectref{sec:mass}.






\subsubsection{Proper nouns}
\label{sec:properN}

%??? Check AGR class of place names


Proper nouns, also called `proper names',  appear to be often viewed as one category and refer to names of people and places.  In Gyeli, however, proper names of persons and proper names of places form two distinct subcategories of one noun type that I broadly call `proper nouns'.  While the two subcategories share some features in which they differ from common nouns, they also differ in a range of aspects. Table \ref{Tab:PropName} lists the features that distinguish all proper names from common nouns as well as those in which person and place names differ from one another.


\begin{table} 
\centering
\begin{tabular}{l|cc}
 \midrule
Feature		& 					Person names	& Place names \\
 \midrule
No noun class marker		& 			\checkmark		& 		\checkmark	\\
No plural formation			& 			\checkmark		&		\checkmark	\\
No object linking H tone		& 			\checkmark		& 		\checkmark	\\
Restriction to a few agreement classes & 		\checkmark		& 		\checkmark	\\   
\hdashline[0.6pt/4pt]
Similative prefix 			& 			\checkmark		& 			\\
Vocative suffix				& 			\checkmark		& 			\\
Special genitive marking		& 			\checkmark		& 			\\
 \midrule
\end{tabular}
\caption{Features of proper names}
\label{Tab:PropName}
\end{table}


 \noindent I will elaborate on each of these aspects in the following. 

In contrast to common nouns, proper names of persons and places never take noun class prefixes nor do they have singular/plural pairings. Names of people can, however, take the associative plural (AP) marker {\itshape bà} which precedes the proper name, as in {\itshape bà Àdà}, referring to Ada and his family or relatives or, depending on the context, to people that share character traits with Ada (i.e.\ people like Ada). The associative plural marker {\itshape bà} is not restricted to proper nouns, but is also used with common nouns and pronouns, as discussed in \sectref{sec:AP}. As proper nouns do not take noun class prefixes, they do not provide any TBU to take an object linking H tone, as discussed in \sectref{sec:OBJTone}.

All proper names trigger agreement just like common nouns. In comparison to common nouns, they are very restricted in the agreement classes to which they are affiliated. 
All proper names of persons are a subcategory of class 1. In contrast, all proper names of places such as settlements, villages, towns, rivers, and countries are generally in class 7, with the exception of {\itshape kàmɛ̀rún} `Cameroon' which is also in class 1. Since many of the place names are derived from common nouns,\footnote{The source noun of place names does not necessarily have to come from Gyeli, but could come from another language in the area. Still, the original meaning is recognized and allows for other agreement classes than class 7. Also, even though there are some lexical differences, cognates across languages of the area are often recognizable to speakers and are found in the same gender.}  place names can also agree in gender with the noun they are derived from. For instance,  the village name {\itshape Ngòló} is derived from the Bulu word {\itshape nkôl} `hill'.\footnote{The Bulu name for the village is  {\itshape Nko'olong}.}  % (alexandre 1955: 217)
Since the cognate {\itshape nkùlɛ́ }`hill' in Gyeli belongs to gender 3/4, the village name can trigger agreement patterns both in class 7 and class 3.

Person names feature a range of characteristics that place names do not exhibit. Names of persons productively take the similative prefix {\itshape ná} in the derivation of female names, as discussed in \sectref{sec:SIM}. In contrast, I did not find any place name with this prefix. 
Person names can further take the vocative suffix -{\itshape o}. The vocative suffix never attaches to a place name, but to a few common nouns  of relation that can be used as an address, such as {\itshape nyá-ò} `mother' and {\itshape tá-ò} `father'. The use of the vocative is generally, however, not restricted to nouns, but the morpheme can also attach to other parts of speech, as discussed in \sectref{sec:VOCSuff}.

Finally, person and place names differ in their marking of N + N genitive constructions when the possessor is a proper name. As shown in (\ref{GenAtt}), place names behave like common nouns and take an attributive marker while person names take a special genitive marker.


\begin{exe}
\ex\label{GenAtt} 
\begin{xlist}
\ex\label{GenAtt1} person name
\gll  j-ínɔ̀ {\bfseries ngá} Nampoundi \\
	le5-name GEN $\emptyset$1.PN \\
 \trans `the name of [the woman] Nampoundi'
\ex\label{GenAtt2} place name
\gll j-ínɔ̀ {\bfseries lé} Ngòló \\
    le5-name 5:ATT $\emptyset$7.PN	\\
 \trans `the name of [the village] Ngolo'
\ex\label{GenAtt3} common noun
\gll  j-ínɔ̀ {\bfseries lé} síngì \\
	le5-name 5:ATT $\emptyset$7.cat \\
 \trans `the name of the cat'
\end{xlist}
\end{exe}

\noindent  More information on attributive and genitive markers is provided in \sectref{sec:ATT} and \sectref{sec:GEN}. 





\paragraph{Ethnographic note on naming strategies}

The Bagyeli have bipartite names, consisting of a vernacular name that is followed by a Christian French name.\footnote{The sample of proper names comprises 111 female and male names and covers all proper names from three Gyeli villages, namely Ngolo, Bomnapenda, and Bibira. It also includes some of the names from yet other villages such as Lebdjom (in the Basaa speaking area) and Ebobissé (within Kribi town).} Taking a Christian name seems to imitate the naming strategy of the Bantu farmers since Christiantity does not play a big role in most Gyeli villages. Unless a Gyeli village is in very close contact and on good terms with their farming neighbors, the Bagyeli tend not to go to church and I do not know of any Gyeli village that has their own church at the time of writing. Since the Christian religion is very strong among the Bantu farmers, however, claiming to be Christian in front of outsiders and having a Christian name seem to serve at reducing stigmatization and creating common ground between the Bagyeli and Bantu farmers. Also, the Bagyeli who attend school are more likely to use their Christian name, at least officially, since it is required for enrollment.  In practical terms, however, I have met a few Bagyeli who had forgotten their Christian name. This is not implausible given that the Bagyeli do not call each other by their Christian, but by their vernacular name, and that there is often no official documentation such as birth certificates or ID cards that would remind people of their names. 

The vernacular name is either considered typical Gyeli or a name that is found in other languages of the area as well. If a name occurs in other languages as well, it is most often shared with the Kwasio dialects Mabi and Ngumba, even if the person was born in, for instance, the Bulu contact region. If a name is shared by other languages than Mabi and Ngumba, such as Basaa, Bulu or Fang, it is almost certainly predictable that the person comes from that specific contact region.

Many of the vernacular names have a (derived) meaning, often from the plant world or animal kingdom. Also, many of them are not gender-specific, but can be used for men and women alike. For others, female names can be derived from some male names. The derivations of a female from a male name are numerous and seem largely unpredictable. Differences between a male and a female form of the same name encompass tone differences as in {\itshape Mimbe} (male: {\itshape Mìmbɛ̂}, female: {\itshape Mímbɛ̂}), different prefixation ({\itshape Mgbâ} (M) > {\itshape Mímgbâ} (F) and {\itshape Sàmɛ̀} > {\itshape Màsámɛ̀} (F)), as well as denasalization of a final vowel ({\itshape Mbɔ̀} (F) > {\itshape Mbɔ̃̀} (M)). The most productive derivation strategy is through the similative prefix {\itshape Na}- as in {\itshape Nanze} with its male counterpart {\itshape Nze} or {\itshape Nandtoungou} which is derived from {\itshape Ntoungou}.
Table \ref{Tab:PN} provides examples of vernacular names as found amongst my consultants and Bagyeli from other Gyeli villages. The table\footnote{A blank cell in the table means that no certain information is available. In contrast, a hyphen (in the Meaning column) means that speakers state that there is no associated meaning with a name.  Languages are abbreviated with their ISO code: gyi - Gyeli, nmg - Kwasio (mostly Mabi in this case), fan - Fang, bum - Bulu, bas - Basaa.} specifies whether a name is used for men and/or women,\footnote{The superscripted \textsuperscript{D} after the gender means that the name has a counterpart in the opposite sex: Mandzoué (F) > Mandzong (M), Mba (M) < Mimba (F),  Mímbɛ̂ (F) > Mìmbɛ̂ (M), Nanze (F) > Nze (M), Nandtoungou (F) > Toungou (M), Tsimbo (F) > Batsimbo (M).} its potential use in other languages of the area, and its meaning (if known).  

The orthography of names\footnote{The orthography is provided by different Mabi speakers since the Gyeli speakers are mostly illiterate.} is a mix between Bantu and French notation strategies which, in some parts, seem to lack a strict convention. For instance, the sound /u/ can be represented by either the French style {\itshape ou} or the Bantu notation {\itshape u}. A word-final /e/, as in {\itshape Mamende} or {\itshape Mabalé}, can either be written with plain {\itshape e} or with the French style {\itshape é}; accents in local orthography do not mark tone. Other versions seem to be admissible as well, for example varying between {\itshape Mabale}, {\itshape Mabalé}, {\itshape Mabali}, and potentially {\itshape Mabally}. This variation can be explained both by idiosyncratic preferences as well as dialectal variation in pronunciation.

\begin{table} 
\begin{tabular}{p{2.5cm}|lll}
Name  & Gender & Languages & Meaning	\\

 \midrule
Ada & M, F & gyi, nmg, fan, bum & -- 				 \\
Bibanga & F & gyi, nmg, fan & -- 			 \\
Bikanda & M, F & & 			\\
Biyang & M & gyi, nmg & remedy			 \\
Bouolpuma & M & gyi & rotten breadfruit 			\\
Bwedila%/Bouedjila
 & M & gyi, nmg 	& -- 			 \\
Kimpile & F & & 			\\
Luonga & F & gyi, nmg & group		 \\
Mabalé & M & gyi, nmg & 			 \\
Mambi & M & gyi & behavior 			\\
Mandzoué & F\textsuperscript{D} & & 			\\
Manligui & F & & 			\\
Mba & M\textsuperscript{D} & gyi & rank 				 \\
Mbiambo & F & gyi & plenty 			 \\
Mimbanji%/Mimbanguie 
& M & gyi & arbalest, crossbow 	 \\
Mímbê & F\textsuperscript{D} & gyi, bas, bum & -- 			 \\
Minlar & M & gyi & union 		 \\
Nalingui & F & gyi, nmg & -- 	\\
Nanze & F\textsuperscript{D} & gyi, nmg, bum, bas &	panther		\\
Nandtoungou & F\textsuperscript{D} & gyi, nmg & --		\\
Nashuong & F & gyi & young palm heart?	\\
Ngolo & F & gyi, nmg & --				\\
Ngo Minsem & F & gyi, bas & daughter of Minsem		\\
Nguiamba & M & &					\\
Ngusa & M & gyi, bas & -- 		\\
Nziwu & M & gyi, nmg & Great antelope			\\
Sedyua%/Segua 
& M & gyi, nmg & derived from civet?	\\
Tsimbo & F\textsuperscript{D} & gyi, nmg &	outcast	\\
 \midrule				
\end{tabular}
\caption{Gyeli proper names (in local orthography)}
\label{Tab:PN}
\end{table}



In addition to the vernacular and Christian name, many of my consultants, both men and women, have nicknames by which they are consistently called in everyday life. They acquire their nicknames either through their parents or peers or even sometimes come up with a nickname on their own. Usually, nicknames refer to something that a person has achieved or say something about the person's character. Nicknames also come from Western languages (French, English). Examples of nicknames used in Ngolo include {\itshape Bataillon} or {\itshape Délégué}. Also outsiders might receive a nickname; the project's cameraman was thus called {\itshape Freeboy}, presumably due to his nonchalant attitude towards kneeling in the mud while filming. There seems to be a tendency to pick nicknames originating from other languages, as is particularly obvious with Western words. Local languages also provide nicknames, for instance {\itshape ə̀və́lə̀ tíd} `red animal' from Bulu which was given to a woman for her bright color of skin.



\subsubsection{Nominalized past participles}
\label{sec:NounPart}

Nominalized past participles are defective nouns that are of the most deviant noun type. Their category label does not imply that there are non-nominalized participles. All nouns of this category are derived from verbs and function like a past participle, as illustrated in (\ref{NounPart1}). More information on the derivation process is provided in \sectref{sec:NOMPart}.

\begin{exe}
\ex\label{NounPart1} 
\glll yíì {\bfseries nkɛ̀lá} \\
yíì n-kɛ̀l-a \\
7.ID N-hang-NOM \\
\trans `It is hung up [lit.: a hung-up one].'
\end{exe}

Unlike full nouns, nominalized past participles never allow a plural form. Thus, while the nominal predicate in (\ref{NounPart2a}) takes the plural noun class marker {\itshape ba}-, agreeing in number with the subject, this is not the case for the nominalized past participle in (\ref{NounPart2b}).  

\begin{exe}
\ex\label{NounPart2} 
\begin{xlist}
\ex\label{NounPart2a} 
\glll Àdà nà Mambì báà {\bfseries bà}ngɛ̀lɛ́nɛ̀ \\
 Àdà nà Mambì báà ba-ngɛ̀lɛ́nɛ̀ \\
$\emptyset$1.PN COM $\emptyset$1.PN 2.COP ba2-teacher \\
\trans `Ada and Mambi are teachers.'
\ex\label{NounPart2b} 
\glll Àdà nà Mambì báà {\bfseries mbánâ} \\
 Àdà nà Mambì báà m-bán-a \\
$\emptyset$1.PN COM $\emptyset$1.PN 2.COP N-marry-NOM \\
\trans `Ada and Mambi are married [lit.: are married ones].'
\end{xlist}
\end{exe}


The occurrence of nominalized past participles is restricted to the predicate position of a STAMP copula construction (\sectref{sec:SCOP}), as shown in (\ref{NounPart1}) and (\ref{NounPart2}). Consequently, they do not serve as an argument or adjunct, unlike common and proper nouns. Given their distributional restriction, they never occur in a position where they would trigger agreement, for instance through the addition of agreement targets in the predicate NP. Likewise, speakers would not replace the nominalized past participle through a pronoun that could indicate the affiliation to an agreement class. 

Another hypothesis would be to consider these forms as verbs, given their verbal stem and translation. I am ruling this option out: despite significant differences from common and proper nouns,  I classify nominalized past participles as a defective noun type. Evidence for this comes from their prefixation and tonal behavior, and their distribution in sentences which distinguishes them from verbs.   Morphologically, verbs do not take prefixes, but only suffixes. The nominalized past participle, however, consistently takes a nasal prefix. Verbs only have tonal specifications for the first syllable while the potential second and third syllables are underlyingly toneless and thus surface as L in isolation, as explained in \sectref{sec:Tonology}. In contrast, nominalized past participles never surface L on the last syllables, but either H or HL.  Also in terms of their distribution in sentences, nominalized past participle forms cannot be verbs since verbs follow the subject-tense-aspect-mood-polarity (STAMP) marker, as described in \sectref{sec:SCOP}. These participles cannot combine with the STAMP marker. They only occur in STAMP copula constructions, as exemplified in (\ref{COPNom}) and discussed in more detail in \sectref{sec:COP}. There are several predication types for copula constructions, including nominal and adjectival copulas, but never verbs.

\begin{exe}
\ex\label{COPNom} 
\begin{xlist}
\ex\label{COPNoma} 
\glll ndáwɔ̀ nyíì mbúyâ (nà vìyɔ́) \\
	ndáwɔ̀ nyíì m-búy-a (nà vìyɔ́) \\
	9$\emptyset$.house 9.COP N-destroy-NOM COM 8$\emptyset$.fire \\
 \trans `The house is destroyed (by fire).'
\ex\label{COPNomb} 
\glll ndáwɔ̀ nyí búyá nà vìyɔ́\\
	ndáwɔ̀ nyi-H búy-a-H nà vìyɔ́\\
	9$\emptyset$.house 9-PRES destroy-PASS-R COM 8$\emptyset$.fire \\
 \trans `The house is being destroyed by fire.'
\end{xlist}
\end{exe}

The translation might also suggest a passive reading. There is a distinct passive form, which is discussed in \sectref{sec:PASS} and exemplified in (\ref{COPNomb}). While  the passive and the nominalized past participle are two distinct categories, both categories are, however, linked semantically and formally. In terms of semantics, their subjects are the undergoer of an action while the agent would appear in an adjunct or not at all. This is true for both categories, but since the nominalized past participle is more about the result, the agent is mentioned very rarely.

Formally, both categories take a suffix -{\itshape a}.  There are two possibilities to analyze -{\itshape a} with respect to the different categories. Either, one could posit that it is the same suffix which just takes different tonal patterns in different categories. Or one could assume two different suffixes -{\itshape a} which each come with their own tonal patterns for the passive and the nominalized past participle. I choose the second option, as reflected  in the glosses. The reason for this is not only the different tone patterns associated with the different suffixes, but also a (synchronically) insufficient link between the two categories. Thus, glossing both suffixes -{\itshape a} as passive (and assuming that nominalization is primarily encoded through the nasal prefix in the nominalized past participle) presupposes a derivation chain with passivization as a necessary step.  This assumption is, however, not justified since many verbs which have a nominalized past participle form lack a passive form: only 105 (27\%) verbs take a passive form, but 325 (86\%) have a nominalized past participle form.

 


\subsection{Mass nouns and countability}
\label{sec:mass}

Gyeli has a mass/count distinction like many languages in the world. Formally, one can distinguish nouns that occur both in a singular and a plural form, those that only come in a singular noun class, and those which only have a plural form. Nouns with a singular and a plural form are mostly countable. Typically, they describe material entities such as humans, animals, plants, tools and the like that come as individualizable objects. 

Mass nouns are most frequently and regularly found in the transnumeral gender 6. (More information on the gender and agreement system is provided in \sectref{sec:Gender}.) Semantically, they include liquids, as in (\ref{LMN}). 

\begin{exe}
\ex\label{LMN} Liquid mass nouns
\begin{xlist}
\ex ma-jíwɔ́ `water' 
\ex ma-vúdɔ́ `oil'
\ex ma-tàngò `palm wine'
\ex ma-vínó `pus'
\ex ma-nzálɛ̀ `urine'
\ex ma-dyúmù `sperm' 
\end{xlist}
\end{exe}

\noindent Also deverbal event nouns of gender 6, as in (\ref{DEN}), are uncountable. More information on their derivation process is provided in \sectref{sec:NOM6}.  

\begin{exe}
\ex\label{DEN} Deverbal event nouns
\begin{xlist}
\ex ma-nyû `drink (n.)' > nyùlɛ `drink (v.)'
\ex ma-bwã̂sà `thoughts' > bwã̂sa `think' 
\ex ma-bwàlɛ̀ `birth' > bwàlɛ `be born'
\ex ma-sâ `game (playing) > sâ `do'
\ex ma-tálá `beginning' > tálɛ `begin'
\ex ma-dìlá `funeral' > dìlɛ `bury'
\end{xlist}
\end{exe}

There are other mass nouns with only a plural form in other agreement classes, but they seem to be less frequent. They mostly belong to class 8 and comprise entities that usually occur as many, for instance {\itshape bè-sìngì} `spirits'. They also include deverbal nouns such as {\itshape bè-déwɔ̀} `food' which is derived from {\itshape dè} `eat'.

Then there are nouns that only have a singular form. Most often, they are abstract nouns of class 7, as illustrated in (\ref{AN}).

\begin{exe}
\ex\label{AN} Abstract nouns
\begin{xlist}
\ex dú `lie' 
\ex sɔ̀mɔ̀nɛ̀ `complaint'
\ex ngɔ̀ngɔ̀lɛ́ `sadness, compassion'
\ex pɔ́nɛ̀ `truth'
\ex sɔ́nɛ̀ `shame'
\ex mɛ̀vâ `pride'
\end{xlist}
\end{exe}

There are a few other singular nouns without a plural form in other classes. Semantically, they describe mass entities which have a rather unspecified shape and lack clear-cut boundaries such as {\itshape pfùdɛ́} `mold' (cl. 9) or {\itshape dùwɔ́} `sky' (cl. 5). {\itshape bíwɔ̀} `bad luck' (cl. 3) is another example of an abstract noun. 
Also a few nouns in agreement class 8 lack a plural form. This is remarkable since class 8 is generally a plural class. As explained in \sectref{sec:NC}, however, there are also singular nouns that trigger class 8 agreement, namely those that lack the CV- noun class prefix {\itshape be}-.  Examples of singular-only class 8 nouns include {\itshape vísɔ́} `sun' and {\itshape vìyɔ́} `fire'.  More examples of uncountable nouns are given in \sectref{sec:MinGen} on inquorate genders.

Finally, there are nouns which display characteristics of both mass and count nouns. They have a singular and a plural form, and semantically designate granular aggregates such as {\itshape nsɛ́/mì-nsɛ́} `sand' or {\itshape ndísì/mì-ndísì} `rice'. In their singular form, they behave like other uncountable nouns, for instance liquids, except that they occur in the singular. This becomes especially obvious when modified by quantifiers (see \sectref{sec:NomQUANT}). If used in the plural form, these  nouns get a reading of `different types of' or `different units of'. In these cases, they grammatically behave more like countable nouns.









\section{Verbs}
\label{sec:V}



Nouns and verbs constitute the two major word classes in possibly all languages in the world, as \citet[408]{viberg2006} points out. But what are verbs and how are they distinguished from nouns? \citet[9]{schachter2007} provide a general, semantically based definition, stating that
\begin{quote}
``{\itshape Verb} is the name given to the parts-of-speech class in which occur most of the words that express actions, processes, and the like.''
\end{quote}

\noindent Other properties that the authors highlight include, for instance, the verbs' foregrounding of temporal relations as well as their function as predicates. After all, characteristics of verbs (as any other word class) are language specific and therefore, it makes sense to distinguish them based on a given language's properties. In Gyeli, nouns and verbs are distinct in many ways. As shown in \chapref{sec:Phon}, they differ on phonological grounds, for example in their distribution of phonemes and tones, nouns allowing a larger degree of freedom while verbs restrict occurrences of consonants, vowels, and tones more. On a morphological level, nouns take prefixes which Gyeli verbs do not. In contrast, verbs take (extension) suffixes which is not the case for nouns. In terms of syntactic function, verbs serve canonically as predicates while nouns (or noun phrases) constitute arguments to a given predicate. These various formal differences show clearly that nouns and verbs in Gyeli belong to different word classes.

In the following, I will first describe the structure of the verb.  I then discuss different verb types, including main verbs and auxiliary verbs. 




\subsection{Verb structure}
\label{sec:StructVerb}

The Gyeli verb consists of a root that can take a valence-changing suffix, as shown in Table  \ref{Tab:Gyeliverb}.

\begin{table} 
\centering
\setlength{\tabcolsep}{3pt}
%\scalebox{0.9}{
\begin{tabular}{|l|p{16mm}p{16mm}p{16mm}p{16mm}lp{16mm}p{16mm}|}
 \midrule %\noalign{\smallskip}
Slot &  &  &  &  & {\bfseries Radical} & Prefinal &   \\ %\noalign{\smallskip}
 \midrule %\noalign{\smallskip}
Function &  &  &  &  & {\bfseries Root} & valence change &   \\
%\noalign{\smallskip} 
 \midrule
\end{tabular}
\caption{The Gyeli verb structure}
\label{Tab:Gyeliverb}
\end{table}

Table \ref{Tab:Gyeliverb} indicates the ``slot'' in which the root and valence changing suffix occur and is based on the morphological Bantu verb schema by \citet[184]{gueldemann2003}. While the general Bantu verb schema has eight slots, four before the root and three after the root, Gyeli has a more reduced verbal structure. For instance, subject concord and TAM information are not encoded on the verb, but in a subject-tense-aspect-mood-polarity (STAMP) portmanteau morpheme (\sectref{sec:SCOP}). 

I follow the Bantuist tradition (e.g.\ \citealt{guthrie71}, \citealt{hyman93}, and  \citealt{schadeberg2003}) in my terminological distinction between  {\itshape radical} and {\itshape stem}. The radical, also called {\itshape root}, is the ``irreducible core'' (\citealt[14]{guthrie71}) of the verb that cannot be parsed into further morphemes. In Gyeli, its phonological structure is typically C(C)VC-, but there are exceptions in surface forms pertaining to an additional vowel in some bisyllabic underived verbs (\sectref{sec:StemFinalV}) and the deletion of the root final consonant in monosyllabic verb forms (\sectref{sec:RFCV}). 

The root in Gyeli can function as an independent word without any further bound morphemes attached, as exemplified in (\ref{Vroot1}) for monosyllabic verb roots. All monosyllabic verbs consist of a root only. Under derivation, a root final consonant (or variants thereof) will surface, as described in \sectref{sec:RFCV}. This root final consonant is deleted in monosyllabic roots in order to adhere to an open syllable structure. 

\begin{exe}
\ex\label{Vroot1} Monosyllabic roots
\begin{xlist}
\ex dè `eat' 
\ex kwê `fall'
\ex bvúɔ̀ `break (tr.)'
\end{xlist}
\end{exe}

\noindent  Also some bisyllabic verb roots satisfy the criterion of an irreducible core, as in (\ref{Vroot2}).


\begin{exe}
\ex\label{Vroot2} Bisyllabic roots
\begin{xlist}
\ex bámɔ `scold' 
\ex púndi `polish'
\ex gyàga `buy'
\end{xlist}
\end{exe}

The root can take an extension or expansion derivation suffix that brings about a valence change. A list of all verbs in the database and their extension morphemes is given in Appendix I. %\ref{sec:AppendixI}
  The root and the potential suffix constitute the stem.\footnote{Traditionally, the stem additionally includes the {\itshape final vowel} that encodes tense-aspect-mood information in more agglutinative Bantu languages. In these languages, Bantuists use the term {\itshape base} to designate the root and potential derivation suffixes without the final vowel. In Gyeli, however, there is no final vowel. Therefore, this distinction is not necessary.}
There are also bisyllabic verbs that consist of a root plus extension suffix, as shown in (\ref{Vroot3}). Derivation with extension and expansion suffixes is described in \sectref{sec:VDeriv}.

\begin{exe}
\ex\label{Vroot3} Bisyllabic stems
\begin{xlist}
\ex  bè.n-a `be refused' (passive extension -{\itshape a}) 
\ex\label{Vroot3b} jì-bɔ `close sth.' (-{\itshape bɔ} expansion)
\ex\label{Vroot3c} vú-lɛ `get rid of sth.' (-{\itshape lɛ} expansion)
\end{xlist}
\end{exe}

Thus, whether a bisyllabic verb consists of a root only, as in (\ref{Vroot2}), or constitutes a stem with a root plus suffix, as in (\ref{Vroot3}), depends on the synchronic function of the second syllable.   In synchronic bisyllabic verb roots, the vowel of the second syllable is part of the lexeme since its shape is not predictable on morphophonological or morphosyntacic grounds. In contrast, in a bisyllabic stem, the second syllable functions as a valence changing suffix. A root vs.\ stem contrast can be found even with the same lexeme, as for instance with the root {\itshape bédɔ} `mount (tr.)' whose passive form {\itshape béd-a} `be mounted' is analyzed as a stem.
A more detailed discussion on the status of the final vowel as part of the root is given in \sectref{sec:StemFinalV}.

The number of transparent derivational suffixes a root can take is restricted to one.\footnote{As discussed in \sectref{sec:VDeriv}, two categories, e.g.\ applicative and passive, can be merged into one morpheme through vowel change of the applicative suffix in trisyllabic verbs.} Derivational extensions can, however, come as mono- or bisyllabic suffixes, allowing a maximum of three syllables in a stem, as shown in (\ref{Vroot4}). As indicated by the dot for syllable boundaries, syllable and morpheme boundaries do not always coincide. In fact, if the derivation suffix starts with a vowel, the root final consonant gets resyllabified as the onset of the extension vowel.

\begin{exe}
\ex\label{Vroot4} Trisyllabic stems
\begin{xlist}
\ex  gyá.mb-ɛ.lɛ `cook for sb.' (applicative extension -{\itshape ɛlɛ}) 
\ex lɛ̀.b-a.la `follow each other' (reciprocal extension -{\itshape ala})
\ex dyɛ́.g-ɔ.wɔ `get in a leaning position' (positional extension -{\itshape -ɔwɔ})
\end{xlist}
\end{exe}

In the following, I will discuss the shape of the verb root in more detail, focusing on two issues. First, I explore the status of the Gyeli stem final vowel, arguing that it does not occupy the ``final'' slot of \posscitet{gueldemann2003} morphological Bantu verb structure. I then describe root final consonants and their variants.




\subsubsection{Stem final vowel} 
\label{sec:StemFinalV}

Though the Gyeli verb structure is significantly different from \posscitet{gueldemann2003} morphological verb schema, one might wonder whether Gyeli does have a vowel in the ``final'' slot which is typically related to tense-aspect-mood. 
Due to a canonical CV syllable structure, Gyeli verbs always end in a vowel, but they are by no means comparable to the ``final vowel'' in the ``final'' slot found in eastern and southern Bantu languages where the final vowel has a grammatical function. In contrast, Gyeli root and stem final vowels are lexically specified.  As discussed in \sectref{sec:CardVowels}, vowel quality is restricted by the stem's syllable length. In monosyllabic verbs, any of the seven vowels, except for /o/, can occur in final position, while bisyllabic verbs only allow five vowels in this position, /i/, /o/, /ɛ/, /ɔ/, /a/. Trisyllabic verb stems only allow /ɛ/, /a/, and /ɔ/ as a final vowel.

Another argument not to consider Gyeli stem final vowels as occupying the final slot of \posscitet{gueldemann2003} Bantu verb structure comes from verb extensions. When Bantu languages such as Swahili add an extension morpheme in the prefinal slot, the final vowel is not necessarily affected by this. The Swahili stem {\itshape chek-a} `laugh', for instance, keeps the final vowel {\itshape -a} even if the stem is extended by a causative morpheme {\itshape -Ish-}: {\itshape chek-esh-a} `make laugh'. Extension morphemes in Gyeli, however, come with their own final vowels and override a bisyllabic root final vowel  as in {\itshape jílɔ} `be satisfied' $\rightarrow$ {\itshape jíl-ɛsɛ} `make satisfied'.

While all verb final vowels in Gyeli are lexically specified, they differ with regards to their morpheme affiliation. There are three types of verb final vowels. First, a verb final vowel is the nucleus of the verb root in monosyllabic verb forms. It is tonally specified and does not usually change in derived forms. The root vowel ends up in the final position because the final root consonant is deleted, as illustrated in (\ref{Vroot5}). The deleted root final consonants in parentheses only surface with derived forms of the verb, as with the passive forms in (\ref{Vroot5}). (More information on root final consonant deletion is provided in the next section.) 

\begin{exe}
\ex\label{Vroot5}
\begin{xlist}
\ex dy{\bfseries à}(y) `sing' < dyày-a `be sung'
\ex kw{\bfseries à}(g) `grind' < kwàg-a `be ground'
\ex nd{\bfseries à}(ng) `cross' < ndàng-a `be crossed'
\end{xlist}
\end{exe}

Second, in bisyllabic verb roots, the final consonant is followed by a lexicalized (underlyingly toneless) vowel. This vowel is synchronically part of the root since its quality is not predictable and does not have any grammatical function. In derived forms, this vowel is deleted, as shown in (\ref{Vroot6a}). The fact that these additional root vowels are not specified for tone, a property they share with verb extension and expansion suffixes, suggests that diachronically they were derivation suffixes as well.

\begin{exe}
\ex\label{Vroot6a}
\begin{xlist}
\ex fùl{\bfseries ɔ} `descend' < fùl-a `be descended'
\ex dyɔ̀d{\bfseries ɛ} `deceive' < dyɔ̀d-a `be deceived'
\ex gyáng{\bfseries a} `work' < gyáng-ɛsɛ `make sb.\ work'
\end{xlist}
\end{exe}

\noindent Final vowels of monosyllabic verb forms with a diphthong or long vowel as nucleus are treated the same way. As shown in (\ref{Vroot6b}), the second vowel of the diphthong gets deleted in derived forms.

\begin{exe}
\ex\label{Vroot6b}
\begin{xlist}
\ex bvú{\bfseries ɔ̀} `break (tr.)' < bvú.g-ɛ `make break'
\ex dyù{\bfseries ù} `kill' < dyù.w-a.la `kill each other'
\ex ní{\bfseries ɛ̀} `be beautiful' < ní.ng-ɛ.sɛ `make beautiful'
\end{xlist}
\end{exe}

\noindent Historically, these verbs were likely bisyllabic, as the examples in (\ref{Vroot6a}). This would have involved a process in which first the root final consonant got deleted and then the vowel of the second syllable was merged with the first syllable's nucleus. Synchronically, the second vowel of the diphthong is clearly part of the root vowel since it is specified for tone.

The third type of stem final vowel is specified through the derivation suffix a root can take, as shown in (\ref{Vroot7}). 

\begin{exe}
\ex\label{Vroot7}
\begin{xlist}
\ex dyúw-ɛl{\bfseries ɛ} `listen to'
\ex ntɛ́g-al{\bfseries a} `bother each other'
\ex pwàs-ɔw{\bfseries ɔ} `stretch oneself out'
\end{xlist}
\end{exe}

\noindent The segments of derivation suffixes do not change in different tense-aspect-mood categories, but their tonal patterns do (\sectref{sec:GramTM}).






\subsubsection{Suppletive root vowels}
\label{sec:SRV}

Gyeli has a few verbs which change their root vowel in (some) derived forms. I view these as lexically specified exceptions since they do not follow any predictable pattern and are generally rare. All suppletive root vowel forms are given in Table \ref{Tab:SupplRV}.


\begin{table} 
\centering
\begin{tabular}{ll|lll|l}
\multicolumn{2}{l|}{Underived form}  & Reciprocal         & Passive & Causative  & Variants \\
 \midrule
lùà 	& `curse' &  lɔ̀g-ala &  lɔ̀g-a & lɔ̀g-ɛsɛ	&  		ua/ɔ \\
lṹã̀		& `whistle' &  lɔ́ng-ala &  lɔ́ng-a & 	&  	ua/ɔ \\
túà		& `move places' & tɔ́g-ala		&  & tɔ́g-ɛsɛ &  ua/ɔ \\ 		
bwà 		& `become big' &  bɔ̀g-ala &  & 		& 			wa/ɔ	 \\
bwádɔ 	&  `wear' &  bɔ́d-ala &   & bɔ́d-ɛsɛ 	&  		wa/ɔ	\\
bwɛ̀dɔwɔ	& `be tasty' &   & 		& bɔ́d-ɛsɛ	& 		we/ɔ  \\ \hdashline[0.5pt/5pt]
b{\bfseries wè}		& `catch' & b{\bfseries è}y-ala & 	b{\bfseries ù}l-ɛ	& 		&   we/u  \\
k{\bfseries wê} 		& `fall' 	&  k{\bfseries wé}y-ala & 		&  k{\bfseries ù}-ɛsɛ  & 	 we/u \\
lâ		& `harvest' & léy-ala		& léy-a &  &  a/e \\ 
lága		& `contaminate' & lég-ala	 &  & lég-ɛsɛ  &  a/e \\ 
bô 		& `lie down'                 &          & búg-a 	       & 	& 		 o/u \\
yíɛ̀ 		& `dodge' 	& yé-ala &  &   & 		iɛ/e \\
d{\bfseries è}		& `eat' &  d{\bfseries í}y-ala & d{\bfseries í}b-a	& 	d{\bfseries í}l-ɛsɛ	&    e/i  \\
 \midrule
\end{tabular}
\caption{Root final consonant variants (monosyllabic verbs)}
\label{Tab:SupplRV}
\end{table} 

Ten out of the thirteen suppletive root vowels are regular in the sense that all derived forms have the same suppletive vowel. For instance, {\itshape lùà} `curse' takes {\itshape ɔ̀} as root vowel in its reciprocal, passive, and causative forms. Also, the suppletive vowels retain the same tonal pattern as in the underived form, namely H for underived verbs which have a HL pattern and L for L underived verbs. There are a few more irregular cases, however, which have different suppletive vowels for different derived forms and/or tonal changes on the suppletive vowel. {\itshape bwè} `catch', for example, retains /e/ in the reciprocal form {\itshape bèyala}, but loses the glide /w/, while it has a suppletive vowel /u/ in the passive form {\itshape bùlɛ}. All root vowels remain L. In contrast, {\itshape kwê} `fall' has a regular reciprocal form {\itshape kwéyala}, both in terms of the vowel and its tone, but an irregular causative form {\itshape kùɛsɛ} with both a suppletive vowel and a tonal change from H to L. Finally, {\itshape dè} `eat' has the same suppletive vowel /i/ for all derived forms, but all derived forms have a H instead of a L tone.

Most verbs with suppletive root vowels have monosyllabic stems containing the diphthong /ua/ or the glide /w/ which is changed to /ɔ/ in derived forms. The verb of the underived form is, however, not predictive of a necessary vowel change in derived forms since verbs generally keep their glides and vowels in derived forms. (\ref{SRV1}) gives an opposition between a regular and an irregular form.

\begin{exe}
\ex\label{SRV1}
\begin{xlist}
\ex bwà `give birth' → bwàl-ɛsɛ (CAUS)
\ex bwà `become big' → b{\bfseries ɔ̀}d-ala (RECIP)
\end{xlist}
\end{exe}

\noindent Other suppletive forms, for instance from /a/ to /e/ in {\itshape lâ} `harvest' or /e/ to /i/ in {\itshape dè} `eat' seem even more exceptional.



\subsubsection{Root final consonant variants}
\label{sec:RFCV}

Generally all verb roots (with a few exceptions) have a final consonant which is lexically specified and only surfaces when a vowel initial derivation suffix attaches. In monosyllabic stems (\ref{Vroot1}) and with derivation suffixes that are consonant initial such as -{\itshape lɛ} or -{\itshape bɔ} in (\ref{Vroot3}), the root final consonant is deleted. In turn, when deriving a monosyllabic verb, the question is which root final consonant will it have. 

As shown in Table \ref{Tab:EpenC}, the majority of monosyllabic stems have the same root final consonants in all their derived forms.\footnote{This is based on 86 monosyllabic verb stems. As discussed in \sectref{sec:SyllV}, there are 88 monosyllabic verb stems in my database. Yet, not all of them undergo derivation. {\itshape dɔ̀} `negotiate' and {\itshape kɛ̀}  `go' do not have any derived forms and therefore the underlying root final consonant never surfaces.} The types of consonant that can consistently appear root finally are limited to seven: /ŋg/, /g/, and /y/ are the most frequent ones while /l/, /s/, /n/, and /w/ are rare. There are two exceptions to this general pattern. First, eleven monosyllabic verb stems have different root final consonants with different verb extensions, and second, there are six verbs which consistently take no root final consonants in any of their forms.

\todo[704]{indent the consonants, also, the dotted lines don't look so good although it would be nice to have a visually "weaker" line}
\begin{table} 
\centering
\small
\begin{tabular}{lll|llll}
Root ending & \multicolumn{2}{l|}{Frequency}  & \multicolumn{3}{l}{Example} \\  \midrule
{\bfseries Consonant} & 	{\bfseries 69} & {\bfseries (80.1\%)} &  &  &   \\
  /ng/ & 	23 & & sã̂ `vomit' & → &  sá{\bfseries ng}ala `vomit together' \\
  /g/   & 	22 & & dvùɔ̀ `hurt' & → & dvù{\bfseries g}ɛsɛ `make hurt' \\
  /y/   & 	17 & & bà `smoke' & → & bà{\bfseries y}aga `smoke (by itself)' \\
  /l/    &        3  & & vɔ̂ `be calm' & → & vɔ́{\bfseries l}ɛsɛ `make calm' \\ 
  /s/  & 	2  &  & sɔ́ɔ̀ `continue' & → & sɔ́{\bfseries s}ɛlɛ `continue with sth.' \\ 
  /n/  & 	1  & & nyɛ̂ `see' & → & nyɛ́{\bfseries n}ala `see one another' \\
   /w/  & 	1  &  & dyû `kill' & → & dyú{\bfseries w}ala `kill one another' \\ \hdashline[0.5pt/5pt]
{\bfseries Variants}   & {\bfseries 16} &  {\bfseries (18.6\%)} & see Table \ref{Tab:RFCVmono} &  &  \\ \hdashline[0.5pt/5pt]
{\bfseries No consonant}     & 	{\bfseries 1}  & {\bfseries (1.2\%)}   & dyâ `lie down' & → & dyáala `lie down together' \\
 \midrule
\end{tabular}
\caption{Root final consonants in the derivation of monosyllabic verbs}
\label{Tab:EpenC}
\end{table} 

\noindent  I do not include bisyllabic verb roots in this analysis since there is never a question which root final consonant will surface as there are no forms in which the consonant gets deleted: all bisyllabic verb roots end in a vowel. Comparatively, bisyllabic verb roots allow for a larger variety of root final consonants, including frequently also /m/, /d/, /mb/, /nd/, and /ŋg/.

The diversity of root final consonants surfacing in derived verb forms likely has a historical explanation. Some monosyllabic verb stems may originate from a diachronic extension that got reduced and merged with the monosyllabic root. In the process, the onset consonant of the second syllable--the historical extension suffix--got lost in monosyllabic forms and the suffix vowel got merged with the root vowel.  This reduction is synchronically reflected in monosyllabic verb stems with diphthongs and long vowels, as discussed in \sectref{sec:Diph} and \sectref{sec:VLength}. The original consonants still surface in some derived forms. This scenario would explain why only a limited number of consonants can now serve as root final consonants: they are related to a limited number of suffixes, some of which do not exist anymore.

The quality of the root final consonant that will surface in the derivation of monosyllabic verbs is not (entirely) predictable on  phonological grounds, as the oppositions in (\ref{RFC0}) to (\ref{RFC2}) show.

\begin{exe}
\ex\label{RFC0}
\begin{xlist}
\ex bwà `give birth'  → bwà{\bfseries l}-ɛsɛ `make give birth'
\ex bwà `become big' → bɔ̀{\bfseries g}-ala `become big together'
\end{xlist}
\end{exe}

\begin{exe}
\ex\label{RFC1}
\begin{xlist}
\ex bâ `marry'  → bá{\bfseries n}-ala `marry each other'
\ex bà `smoke' → bà{\bfseries y}-ala `smoke together'
\end{xlist}
\end{exe}

\begin{exe}
\ex\label{RFC2}
\begin{xlist}
\ex nyâ `suckle'  → nyá{\bfseries ng}-ɛsɛ `breast-feed'
\ex nyàà `defecate' → nyà{\bfseries g}-ɛsɛ `make defecate'
\end{xlist}
\end{exe}

\noindent There are, however, some tendencies to predict the underlying root final consonant based on the phonological shape of the monosyllabic verb stem.  
Monosyllabic stems ending in nasal vowels, for instance, almost exclusively have /ŋg/ as root final consonant, as exemplified in (\ref{epenng}). This ties in with the scenario of a historical extension suffix that has been lost: /ŋg/ may have been the onset of the suffix that was lost, while nasality survived on the root vowel. 

\begin{exe} \ex \label{epenng}
\begin{tabular}{lllll}
lã̂ & `pass' & → & là{\bfseries ng}ɛlɛ & `let pass, spend time' \\
kẽ̀ & `shave' & → & kè{\bfseries ng}ala & `shave one another' \\
sã̂ & `vomit' & → & sá{\bfseries ng}ɛsɛ & `make vomit' \\
dyũ̂ & `be hot' & → & dyú{\bfseries ng}ɛlɛ & `heat sth.' \\
\end{tabular}
\end{exe} 

Another tendency is found with monosyllabic verb stems containing a diphthong. Their final root consonant is almost exclusively /g/, as shown in (\ref{epeng}), with a few exceptions concerning the diphthong /iɛ/ which sometimes may also take /y/ as in {\itshape tsíyala} `cut each other', derived from {\itshape tsíɛ̀} `cut'.


\begin{exe} \ex \label{epeng}
\begin{tabular}{lllll}
dvùɔ̀ & `hurt (intr.)' & → & dvù{\bfseries g}ala & `hurt one another' \\
lùà & `curse' & → & lɔ̀{\bfseries g}a & `be cursed' \\
tɔ̀à & `boil (intr.)' & → & tɔ̀{\bfseries g}ala & `boil together' \\
líɛ̀ & `cede, let' & → & lí{\bfseries g}ala & `let to one another' \\
\end{tabular}
\end{exe} 

\noindent All other root final consonants seem not to be predictable on phonological grounds.


There are two exceptions to the general pattern described so far. First, in a few cases, the same underived monosyllabic verb stem has different root final consonants with different extension morphemes.  Table \ref{Tab:RFCVmono} gives an exhaustive list of all final root consonant variants that occur in the database. While there are usually only two variants for the same lexical root, {\itshape dè} `eat' shows that there can be even three variants. 

\begin{table} 
\centering
\small
\begin{tabular}{ll|llll|l}
\multicolumn{2}{l|}{Underived form}  & Reciprocal         & Passive & Causative & Applicative & Variants \\
 \midrule
bâ 			 & `marry' &  bá{\bfseries n}-ala & 	       & bá{\bfseries l}-ese	& 		& n/l \\		
bwè 			& `catch' &  bè{\bfseries y}-ala & bù{\bfseries l}-ɛ & 		& 		& 		y/l \\
vû 			&  `leave' &  vú{\bfseries y}-ala &  vú{\bfseries m}-a & 	&  		& 	y/m \\
sĩ́ĩ̀ 			& `approach' & sí{\bfseries ng}-ala & 		& 		&  sí{\bfseries s}-ɛlɛ & ng/s \\
níyɛ			& `be beautiful' & ní{\bfseries ndy}-ala  & 		& ní{\bfseries ng}-ɛsɛ	& 		& ng/ndy  \\
vè'è 			& `try on clothes' &  vè{\bfseries g}-ala & 	& 		&  vè{\bfseries '}ɛlɛ  & 	g/'  \\
dyà			& `sing'    & dyà-ala		& dyà{\bfseries y}-a & 	& 		& 		y/none \\
kwê 			& `fall' 	&  kwé{\bfseries y}-ala & 		&  kù-ɛsɛ  & 		& 	y/none \\
dã̂ 			& `draw water' &  dà{\bfseries ng}-ala &  dã̀-ã̀la & 	&  dã̀-ã̀lɛ	& 	ng/none \\
dyɔ̀			& `laugh'		& dyɔ̀-ala		& dyɔ̀{\bfseries l}as-a & dyɔ̀{\bfseries l}-ɛsɛ &  & l/none \\ 
dè 			& `eat' 	& dì{\bfseries y}-ala & dí{\bfseries b}-a &  dí{\bfseries l}-ɛsɛ & 	& 	y/b/l \\
 \midrule
\end{tabular}
\caption{Root final consonant variants (monosyllabic verbs)}
\label{Tab:RFCVmono}
\end{table} 

Root final consonant variants likely occur for the same reason that root final consonants take different shapes generally. Gyeli probably had more derivation suffixes diachronically and possibly allowed more suffixes than the synchronic limit of three syllables. Different final root consonants may reflect remnants of former extension suffixes or diachronic stacking of derivation suffixes. For instance, /l/ could be related to the expansion suffix -{\itshape lɛ}, as discussed in \sectref{sec:DiaEx}. /s/ in {\itshape sís-ɛlɛ} `approach sb.' could be related to the causative suffix -{\itshape ɛsɛ}. 

Other variant forms may rather reflect an ongoing reduction of segmental material, as in {\itshape vè'è} `try on clothes' which has retained a probably older final consonant /g/ in its reciprocal form {\itshape vèg-ala} that got reduced to a glottal stop in the monosyllabic and applicative forms. The next step on the continuum of segmental reduction is the complete loss of the final root consonant.

Final root consonant variants also occur with bisyllabic verb roots, but they are less frequent. Table \ref{Tab:RFCVbi} shows all their occurrences found in the database.

\begin{table} 
\centering
\small
\begin{tabular}{ll|llll|l}
\multicolumn{2}{l|}{Underived form}  & Reciprocal         & Passive & Causative & Applicative & Variants \\
 \midrule
dyú{\bfseries w}ɔ & `hear' &  dyù{\bfseries w}-ala & 	&   dyú{\bfseries g}-ɛsɛ & dyú{\bfseries w}-ɛlɛ & w/g \\
lí{\bfseries y}ɛ & `leave' &  lí{\bfseries g}-ala & 	& 	& 	& y/g \\
vò{\bfseries w}a & `wake up' &  vò{\bfseries w}-ala	& 	&  vò{\bfseries l}-ɛsɛ 	& 	& w/l \\
tì{\bfseries n}ɔ &  `harvest tubers' &  tì{\bfseries n}-ala &  tì{\bfseries l}-ɛ & 	& 	& n/l \\
bí{\bfseries y}ɔ &  `hit' &  bí{\bfseries n}-ala &  bí{\bfseries l}-a &  bí{\bfseries l}-ese &  bí{\bfseries y}-ɛlɛ & y/n/l \\
 \midrule
\end{tabular}
\caption{Root final consonant variants (bisyllabic verbs)}
\label{Tab:RFCVbi}
\end{table} 

The second exception concerns the lack of a root final consonant in which case adjacent vowels are allowed. Only one verb is known that has a derived form with a zero final root consonant, but no variant consonant in another derived form: {\itshape bvû} `think' whose reciprocal form is {\itshape bvúala}.
In all the other cases of zero root final consonants, there is another consonant variant in another derived form. The variants of zero-consonant and root final consonant in derived verb forms are listed in Table \ref{Tab:RFCVmono}. Other variants of zero-consonants do not show in derived verbs, but in the nominalized past participle (NPP) forms which are discussed in \sectref{sec:NOMPart}. All instances  of variants showing up only in the nominalized past participle are given in Table \ref{Tab:zeroRFC}.


\begin{table} 
\centering
\small
\begin{tabular}{ll|ll|l}
\multicolumn{2}{l|}{Underived form}  & Reciprocal         & NPP  & Variants \\
 \midrule
dyâ			& `lie down'    &  dyá-ala      	& ndyá{\bfseries y}-â     & none/y \\
sâ			& `do'              &  sá-ala      	& nsá{\bfseries y}-â     & none/y \\
yíɛ̀			& `avoid'         &  yé-ala      	& nyé{\bfseries y}-â     & none/y \\
kã̂ 			& `wrap' 	       &  kã́-ala              & nkã́{\bfseries l}-â	     &   none/l \\
láà			& `tell'		& lá-ala		& nlá{\bfseries w}-â        & none/w \\ 
 \midrule
\end{tabular}
\caption{Zero root final consonant variants in nominalized past participles}
\label{Tab:zeroRFC}
\end{table} 

As described in \sectref{sec:SyllV}, there is some variation in the production of vowel sequences in verb stems. While synchronically vowel sequences are found in verb stems, these have  alternate forms with a glottal stop, as illustrated by the two variants in (\ref{RFCglottal}).   

\begin{exe}
\ex\label{RFCglottal}
\begin{xlist}
\ex múɛlɛ `nibble'  → mú-ala (RECIP) → mú-ɛsɛ (CAUS)
\ex mú{\bfseries '}ɛlɛ `nibble'  → mú{\bfseries '}-ala (RECIP) → mú{\bfseries '}-ɛsɛ (CAUS)
\end{xlist}
\end{exe}

\noindent The exact distribution of one variant in comparison to the other is not known. There is variation across speakers as well as within the same speaker. The tendency seems, however, to go towards the loss of segmental material.

 






\subsection{Verb types}
\label{sec:StructTypes}

I distinguish three different verb types in Gyeli, based on their morphosyntactic behavior: main verbs, auxiliary verbs, and light verbs. I define and describe each of these and their potential sub-types in turn.


\subsubsection{Main verbs}
\label{sec:MainVerbs}

I view the main verb as the lexical verb in a phrase which, according to \citet[796]{anderson2011a}, contributes lexical content to an expression. I use both terms, main verb and lexical verb, interchangeably.  The main verb in Gyeli always serves as the semantic head of a clause, but is only the syntactic, finite head in simplex predicate constructions. In complex predicate constructions, the syntactic head is an auxiliary or semi-auxiliary (\sectref{sec:AUX}), while the main verb appears in its non-finite form.  In contrast to true auxiliaries, main verbs can occur on their own in a simplex predicate construction. 


In the simple predicate construction in (\ref{simPred}), the main verb {\itshape gyɛ́sɔ́} `look for' is the syntactic and semantic head of the clause.

\begin{exe} 
\ex\label{simPred}
  \glll  mùdã̂ á {\bfseries gyɛ́sɔ́} bédéwɔ̀ \\
        m-ùdã̂ a-H gyɛ́sɔ-H H-be-déwɔ̀  \\
           N1-woman 1-PRES look.for-R OBJ.LINK-be8-food  \\
    \trans `The woman looks for food.'
\end{exe}

\noindent As the syntactic head, the main verb is inflected for its tense-mood category, as described in \sectref{sec:GramTM}. In this case, {\itshape gyɛ́sɔ́} `look for' is a finite form, carrying a realis marking H tone.

In contrast, in a complex predicate construction, the main verb is the semantic head of the clause.  An auxiliary or light verb serves as the syntactic head, as exemplified in (\ref{simPredb}) with the negative subjunctive auxiliary verb {\itshape dúù}. In this example, the auxiliary is the finite verb encoding the tense-mood category it belongs to. The main verb takes its non-finite form, namely with an underlyingly toneless final vowel, as described in \sectref{sec:toneless}. 

\begin{exe} 
\ex\label{simPredb}
  \glll  mùdã̂ á {\bfseries dúù} {\bfseries gyɛ́sɔ̀} bédéwɔ̀ \\
        m-ùdã̂ a-H dúù gyɛ́sɔ H-be-déwɔ̀  \\
           N1-woman 1-PRES NEG.SBJV look.for-R OBJ.LINK-be8-food  \\
    \trans `The woman must not look for food.'
\end{exe}

%\paragraph{Infinitives}

The non-finite form in (\ref{simPredb}) is, at the same time, the {\bfseries infinitive} form. Infinitive forms in Gyeli do not receive any special morphological or tonal marking, but are identical to their citation form. As shown in \sectref{sec:Tinventory}, second and third syllables are underlyingly toneless, surfacing with a L tone. Infinitive forms are found in complex predicates (\ref{simPredb}) as well as two types of subordinate clauses. First, they occur in subordinate infinitival clauses (\sectref{sec:InfSub}), as in (\ref{INFPred1}).

\begin{exe}
\ex\label{INFPred1}
  \glll    [{\bfseries pámɔ̀} tísɔ̀nì]\textsubscript{INF} á súmɛ́lɛ́ bùdì \\
            pámɔ tísɔ̀nì a-H súmɛlɛ-H b-ùdì \\
             arrive $\emptyset$7.town 1-PRES greet-R ba2-people   \\
    \trans `Having arrived in town, he greets the people.'
\end{exe}

\noindent These subordinate infinitival clauses can also be negated with the negative auxiliary {\itshape tí}, as in (\ref{INFPred2}).

\begin{exe} 
\ex\label{INFPred2} 
  \glll  à múà nà bábɛ̀ [tí {\bfseries wúmbɛ̀} wɛ̀] \\
          a múà nà bábɛ̀ tí wúmbɛ wɛ̀   \\
         1 be COM $\emptyset$7.illness NEG want-R die \\
    \trans `He was sick, without wanting to die.'
\end{exe}


\noindent And second, the main verb of certain attributive clauses with the complementizer {\itshape nâ} appears in its infinitival form, as shown in (\ref{INFPred3}) and explained in more detail in \sectref{sec:COMPINF}. 

\begin{exe} 
\ex\label{INFPred3}
  \glll mùdã̂ à lɔ́ sìsɛ̀lɛ̀ nɔ́nɛ́gá [nâ nyɛ̂ nà {\bfseries kɔ́sɛ̀}] \\
m-ùdã̂ a lɔ́ sìs-ɛlɛ n-ɔ́nɛ́gá nâ nyɛ̂ nà kɔ́sɛ \\
N1-woman 1.PST RETRO scare-APPL 1-other COMP 1 COM cough     \\                        
  \trans `The woman scared the other by her coughing.'
\end{exe}


Infinitives are also found in non-verbal clauses where the infinitive is linked with the STAMP copula {\itshape yíì} of agreement class 7 to its predicate, as shown in (\ref{INFcop}). This construction is further described in \sectref{sec:COP}

\begin{exe} 
\ex\label{INFcop}
  \glll  jíwɔ̀ yíì bíwɔ̀ \\
        jíwɔ yíì bíwɔ̀ \\
           steal 7.COP bad  \\
    \trans `To steal is bad.'
\end{exe}

\noindent Verbal clauses are discussed in \sectref{sec:verbalC} and complex predicates are explained in more detail in \sectref{sec:CompPred}.

In contrast to other types of verbs, lexical verbs take a range of different valencies (intransitive, transitive, ditransitive), as illustrated in (\ref{Val}). 

\begin{exe}
\ex\label{Val}
\begin{xlist}
\ex \label{Vala}
   \glll   Màmbì à {\bfseries kɛ́}   \hfill{intransitive}\\
          Màmbì a kɛ̀-H \\
            $\emptyset$1.PN 1.PST1 fall-PST  \\
    \trans 'Mambi fell.'
\ex\label{Valb}
\glll  Màmbì à {\bfseries bɛ́} lé   \hfill{transitive} \\
       Màmbì a bɛ̀-H lé \\
           $\emptyset$1.PN 1.pst plant-R  $\emptyset$7.tree   \\
    \trans `Mambi planted a tree.'
\ex\label{Valb}
\glll  Màmbì à {\bfseries vɛ́} Bìyã́ màntúà \hfill{ditransitive} \\
       Màmbì a vɛ̀-H  Bìyã́ ma-ntúà \\
           $\emptyset$1.PN 1.PST give-R  $\emptyset$1.PN ma6-mango \\
    \trans `Mambi gave Biyang mangoes.'
 \end{xlist}
\end{exe}

The valency of a verb is lexically specified, but can also be changed through verb extensions, which are explained in \sectref{sec:VDeriv}. Valency change and verb extensions also relate to different voices a main verb can express, such as active, middle voice, and passive voice. Examples of each are shown in (\ref{Vale}).

\begin{exe}
\ex\label{Vale}
\begin{xlist}
\ex \label{Valea}
   \glll  Màmbì à {\bfseries vìdɛ́} màtúà  \\
          Màmbì a vìdɛ-H màtúà \\
          $\emptyset$1.PN 1.PST turn-R $\emptyset$1.car   \\
    \trans `Mambi turned the car.'
\ex\label{Valeb}
\glll  màtúà à {\bfseries vìdɛ́gá} \\
      màtúà a vìd-ɛga-H  \\
       $\emptyset$1.car 1.PST turn-AUTOCAUS-PST     \\
    \trans `The car turned (around).'
\ex\label{Valec}
\glll  màtúà à {\bfseries vìdá} (nà Màmbì) \\
      màtúà a vìd-a-H nà Màmbì \\
        $\emptyset$1.car 1.PST turn-PASS-PST COM $\emptyset$1.PN    \\
    \trans `The car was turned (by Mambi).'
 \end{xlist}
\end{exe}

\noindent In the following, I will discuss two subtypes of main verbs, namely main verbs that require a preposition with their object and main verbs that require a cognate object.

\paragraph{Main verbs requiring a preposition with their object}

A few main verbs require a preposition with their object argument. This seems to generally be a rare case since this only includes 14 verbs (3.7\%) of the 377 verb database. In most cases, the comitative preposition {\itshape nà} is required. All twelve cases are listed in Table \ref{Tab:Verbsna}.

\begin{table} 
\centering
\begin{tabular}{ll}
 \midrule
báàla nà &	 `repeat sth.' \\
bága nà & 	`stop sth.' \\
bísi nà & 	`pay attention to' \\
bvúda nà & 	`quarrel' \\
gyíka nà & 	`resemble sb./sth.' \\
kàmbɔ nà	 & `defend sth.' \\
làdo nà & 	`meet sb.' \\
náàta nà	&  `stick to' \\
njì nà & 	`bring, come for' \\
tã́ã̀la nà & 	`judge sb.' \\
túwanɛ nà	 & `meet with sb.' \\
vúba nà & 	`hug sb.' \\
 \midrule
\end{tabular}
\caption{Main verbs requiring the comitative {\itshape nà}}
\label{Tab:Verbsna}
\end{table}

The other preposition that links an argument is the directional {\itshape bà}. It occurs only in two verbs of the database, namely {\itshape sĩ́ĩ̀ bà} `approach sth.' and {\itshape sísɔ bà} `approach sth.'
Obviously, the prepositions {\itshape nà} and {\itshape bà} occur more frequently in the text corpus, but they are usually found in adjunct noun phrases.

\paragraph{Main verbs with cognate objects}

Gyeli has a few verbs that take a cognate object as argument, as in (\ref{Vcognate}) where the verb is marked in bold.

\begin{exe}
\ex\label{Vcognate}
\begin{xlist}
\ex {\bfseries gyá} gyà `sing (a song)'  
\ex {\bfseries sá} sálɛ́ `work (a work)'
\ex {\bfseries kɛ́} kɛ̀ndɛ̀ `walk (a walk)'
\end{xlist}
\end{exe}

All these verbs can also take a different lexeme as an object, as for instance, in (\ref{Cogna}). They cannot appear without an object, as (\ref{Cognb}) shows.

\begin{exe}
\ex\label{Cogn}
\begin{xlist}
\ex \label{Cogna}
   \glll  mɛ́ kɛ́ tísɔ̀nì  \\
          mɛ-H kɛ̀-H tísɔ̀nì \\
         1\textsc{sg}-PRES go-R $\emptyset$7.town    \\
    \trans `I go to town.'
\ex\label{Cognb}
\glll  *mɛ́ kɛ̀ \\
      mɛ-H kɛ̀  \\
       1\textsc{sg}-PRES go    \\
    \trans `I walk.'
 \end{xlist}
\end{exe}

At the same time, the cognate objects can also appear with other verbs, as shown in (\ref{CognO}).

\begin{exe} 
\ex\label{CognO} 
  \glll yɔ́ɔ̀ bá {\bfseries téé} {\bfseries kɛ̀ndɛ̀} \\
        yɔ́ɔ̀ ba-H téè-H kɛ̀ndɛ̀ \\
         so 2-PRES start.walking-R $\emptyset$7.walk   \\
    \trans `So they go on a walk.'
\end{exe}



\subsubsection{Auxiliaries and semi-auxiliaries}
\label{sec:AUX}

A set of verbs in Gyeli occur as the finite verbal element in a complex predicate construction without (fully) contributing to its lexical content. (Complex predicate constructions are discussed in \sectref{sec:CompPred}.) I call these verbs ``auxiliaries'', which I subdivide into true auxiliaries and semi-auxiliaries.  They both precede the lexical verb. (\ref{G114}) illustrates the contrast between a complex predicate with a semi-auxiliary (the modal {\itshape yánɛ} `must') in (\ref{G114a}) and its simplex predicate counterpart in (\ref{G114b}). In the complex predicate construction, the semi-auxiliary {\itshape yánɛ} is inflected for tense-mood (see \sectref{sec:GramTM}), while the lexical verb {\itshape dyâ} `lie down' appears in its non-finite form. In the simplex predicate construction, the lexical verb receives the tense-mood marking H tone.

\begin{exe} 
\ex\label{G114} 
\begin{xlist}
\ex\label{G114a} 
  \glll  mɛ̀ {\bfseries yánɛ́} {\bfseries dyâ} vâ kùgúù dẽ̀ màfú mábáà \\
     mɛ yánɛ-H dyâ vâ kùgúù dẽ̀ ma-fú má-báà \\
          1\textsc{sg} must-R lie.down here $\emptyset$7.evening today ma6-day 6-two  \\
    \trans `I had to sleep here the evening two days ago [= from today].'
\ex\label{G114b} 
  \glll  mɛ̀ {\bfseries dyá} vâ kùgúù dẽ̀ màfú mábáà \\
     mɛ  dyâ-H vâ kùgúù dẽ̀ ma-fú má-báà \\
          1\textsc{sg}.PST lie.down-R here $\emptyset$7.evening today ma6-day 6-two  \\
    \trans `I slept here the evening two days ago [= from today].'
\end{xlist}
\end{exe}

True auxiliaries and semi-auxiliaries can be distinguished along two parameters, as shown in Table \ref{Tab:AUX}: i) full conjugation potential across different tense-mood categories vs.\ restrictions thereof and ii) full lexical meaning vs.\ no lexical meaning. True auxiliaries are restricted in the tense-mood category they can appear in, as detailed in \sectref{sec:ComplAUX}, as well as in the verbal predicate type they occur in: true auxiliaries can never appear on their own in a simple predicate construction, but require the addition of a lexical verb. Semi-auxiliaries, in contrast, have full lexical meaning and the potential to serve as the finite element in a simple predicate construction. They have full conjugation potential across all tense-mood categories in both simple and complex predicate constructions. 


\begin{table}[!h]
\centering
%\scalebox{0.9}{
\begin{tabular}{l|l|l}
			& 	Full inflection 	& Inflectional restrictions \\
 \midrule
No lexical  & 	& {\bfseries True auxiliaries} \\
	meaning	& 		        & {\itshape nzíí} PROG.PRES, {\itshape nzí} PROG.PST, \\
				& 			& {\itshape nzɛ́ɛ́} PROG.SUBORD, {\itshape lɔ̀} RETRO, \\
				& 			& {\itshape sàlɛ́} NEG.PST, {\itshape pálɛ́} NEG.PST, \\
				& 			& 	{\itshape kálɛ̀} NEG.FUT, {\itshape tí} NEG.IMP  \\   \midrule
Lexical  & {\bfseries Semi-auxiliaries} & \\
meaning	& 	{\itshape kɛ̀} `go', {\itshape lã̀} `pass'                 &  {\itshape bwàá} `have', {\itshape múà} `be almost',    \\
			&   {\itshape njì} `come', {\itshape lígɛ} `stay',	& 	{\itshape dúù} `must not' \\
			&  {\itshape sílɛ} `finish', {\itshape pã̂} `do first',   & 	\\
			&  {\itshape táalɛ} `begin', {\itshape bàga nà} `stop' &    \\
                           & {\itshape lèmbɔ} `know', {\itshape kwàlɛ} `like',  & 	\\
			& {\itshape wúmbɛ} `want', ({\itshape yánɛ} `must')    & 	\\
\end{tabular}
\caption{Auxiliary types}
\label{Tab:AUX}
\end{table}


Since tense-mood categories are only marked tonally, but true auxiliaries are restricted to specific categories, it cannot be proven that they take tonal inflection instead of having a fixed tonal pattern, as there are no contrastive pairs. There are several reasons, however, to classify true auxiliaries as finite verbal elements. First, their tonal patterns coincide with the tonal patterns of their respective tense-mood category. Second, they occur in the same position as semi-auxiliaries that clearly inflect for tense-mood tonal marking. Third, they are followed by a non-finite lexical verb.

Semi-auxiliaries and true auxiliaries can be thought of as distributed towards opposite ends of a grammaticalization scale. Semi-auxiliaries are closest to main verbs while true auxiliaries are highly grammaticalized. 
While most (semi-) auxiliaries fall neatly in either one of the auxiliary types, there are nevertheless some exceptions which behave slightly differently, which reflects their different stages on the grammaticalization path. This is the case for {\itshape dúù} `must not' which is restricted to present and subjunctive clauses and cannot appear as the finite verb in a simple predicate but, unlike true auxiliaries, it has a lexical meaning. The same is true for {\itshape bwàá} `have' with its restriction to the two past categories, and {\itshape múà} `be almost' with its restriction to the future.
Another outlier within the semi-auxiliaries is the deontic modal {\itshape yánɛ} `must' which is the only one that cannot appear in a simple predicate construction. In this respect, it patterns with true auxiliaries, but has a lexical meaning like semi-auxiliaries. Since it has no tense-mood category restrictions, I classify it as a semi-auxiliary.

Both true auxiliaries and semi-auxiliaries encode elements of various functional domains, i.e.\ there is no one-to-one mapping from their form to one specific function.  True auxiliaries comprise some aspect markers and all negation auxiliaries. Semi-auxiliaries also encode some aspect markers as well as modality and motion/posture verbs.




%Light verbs: Language-internal criteria (distributional differences) to distinguish light verbs from main and auxiliary verbs.
%light verbs contribute to a semantically complex but syntactically monoclausal predication and [...] form a syntactically distinct class. \citet[64]{butt2010}

%go, come, finish, give a scream

%\begin{quote}``Even though the light verbs clearly do not have the same predicational content as their full/main verb counterparts, they are always exactly form identical to a full verb and inflect exactly like that full verb. This characteristic sets light verbs apart from auxiliaries in terms of historical change, as auxiliaries may be form identical to a full verb at the initial stages of reanalysis from verb to auxiliary, but then quickly tend to develop away from the original form of the full verb.'' \citet[53]{butt2010} \end{quote}












\section{Adjectives}
\label{sec:QUAL}


Gyeli has a small set of adjectives, as listed in Table \ref{Tab:QUAL}. They constitute a closed class in Gyeli and denote properties of the noun such as value, dimension, and color. 

\begin{table} 
\centering
\begin{tabular}{m|p{3cm}p{3.5cm}}
 \midrule
% \multicolumn{2}{l}{quality} \\
\multirow{2}{*}{\bfseries quality} & mpà &  `good' \\
 & bíwɔ̀ &  `bad' \\
 & díyɛ̀  & `expensive' \\
 & pówàlà & `calm' \\
 & nátĩ̂ & `straight' \\
 \midrule
% \multicolumn{2}{l}{size} \\
\multirow{2}{*}{\bfseries size}  & píyɔ̀ &  `small' \\
 & nɛ́nɛ̀ &  `big' \\
 \midrule
% \multicolumn{2}{l}{color} \\
\multirow{5}{*}{\bfseries color} & námbàmbàlà & `white' \\
 & návyûvyû & `black' \\
 & nábèbè & `red' \\
 & nápfûpfû & `darkened color' \\
 & náyɛ̂yɛ̂ & `lightened color' \\
 \midrule
\end{tabular}
\caption{Adjectives}
\label{Tab:QUAL}
\end{table}





Morphosyntactically, adjectives can be clearly delimited from other parts of speech such as nouns and verbs. Adjectives do not exhibit any verbal qualities such as combining with a STAMP marker or an aspect marker. They can also be clearly distinguished from nouns as they do not exhibit (most) typical nominal behavior. First, they do not take a singular and/or plural form. Second, they do not have the possibility of being modified by other elements of a noun phrase such as demonstratives or possessor pronouns. They can, however, serve as the head of an attributive construction, as further explained below.

This word class in Gyeli meets the broad criteria for adjectives given in the typological literature (which often mixes semantic and morphosyntactic criteria), for instance, following \citet[16]{bhat94} in terms of ``(i) their belonging, prototypically, to the semantic class of properties, and (ii) their having modification (of a noun) as the primary (categorial) function.'' 
\citet{dixon2004}, who postulates that every language has a class of adjectives which is distinct from nouns and verbs, adds to this list predicative use of adjectives, for example as a copula complement. 

Besides these broad criteria, however, adjectives form a vastly diverse class cross-linguistically, as for instance pointed out by \citet{segerer2008} for adjectives in African languages. Gyeli adjectives are unusual from a Bantu perspective in that they do not take any agreement prefixes, but are invariable in their form, both in attributive and predicative use.

In attributive use, adjectives modify nouns in two different default constructions: either the adjective directly follows the head noun or it appears as second constituent in an attributive construction where the attributive marker agrees with the head noun.
\todo[1194]{create vertical spaxe between two types or other representation altogether}
\begin{center}
1) [N ADJ]  2) [N ATT ADJ]
\end{center}

\noindent Examples of both construction types are given in (\ref{qualia}) and (\ref{qualib}), respectively.

\begin{exe}
\ex\label{qualia}
\begin{xlist}
\ex\label{qualia1}
 \gll  nkɔ́lɔ̀ mpà  \\
          $\emptyset$3.watch good  \\
    \trans `a/the good watch'
\ex \label{qualia2}
  \gll   nkɔ́lɔ̀ nɛ́nɛ̀ \\
          $\emptyset$3.watch big \\
    \trans `a/the big watch'
\ex \label{qualia3}
  \gll    nkɔ́lɔ̀ nábèbè \\
          $\emptyset$3.watch red \\
    \trans `a/the red watch'
\end {xlist}
\end {exe}

\begin{exe}
\ex\label{qualib}
\begin{xlist}
\ex\label{qualib1}
 \gll  nkɔ́lɔ̀ wá mpà  \\
          $\emptyset$3.watch 3:ATT good  \\
    \trans `good watch'
\ex \label{qualib2}
  \gll   nkɔ́lɔ̀ wá nɛ́nɛ̀ \\
          $\emptyset$3.watch 3:ATT big \\
    \trans `big watch'
\ex \label{qualib3}
  \gll    nkɔ́lɔ̀ wá nábèbè \\
          $\emptyset$3.watch 3:ATT red \\
    \trans `red watch'
\end {xlist}
\end {exe}

\noindent Constructions that either take or optionally omit the attributive marker are discussed in \sectref{sec:CONC}.

The order of adjective and noun can also be reversed, as a  more marked form. The adjective can either precede the noun directly or can appear as the head of an attributive construction in which case the attributive marker takes the default agreement form of class 7. Choices between construction forms
usually entail a change in meaning, as shown in (\ref{qualif1}). 

\begin{exe}
\ex\label{qualif1}
\begin{xlist}
\ex\label{qualif1a}
 \gll  sɔ́ wà nɛ́nɛ̀  \\
          $\emptyset$1.friend 1:ATT big  \\
    \trans `big friend'
\ex \label{qualif1b}
  \gll   sɔ́ nɛ́nɛ̀\\
          $\emptyset$1.friend big \\
    \trans `important friend'
\ex \label{qualif1c}
  \gll    nɛ́nɛ̀ yá sɔ́ \\
          big 7:ATT friend \\
    \trans `big size of the friend'
\end {xlist}
\end {exe}

It is difficult to detect the exact meaning contrast present. It depends on the lexical semantics of the adjective and noun in question and the construction they stand in. Another example of meaning contrast across different construction types is given in (\ref{qualif2}). While the use of the attributive marker is optional in both constructions, it is preferred in (\ref{qualif2a}) and dispreferred in (\ref{qualif2b})

\begin{exe}
\ex\label{qualif2}
\begin{xlist}
\ex\label{qualif2a}
 \gll  m-wánɔ̀ (wà) bíwɔ̀  \\
          N1-child 1:ATT bad  \\
    \trans `bad child [bad character traits]'
\ex \label{qualif2b}
  \gll   bíwɔ̀ (yá) m-wánɔ̀\\
          bad 7:ATT N1-child \\
    \trans `ugly child'
\end {xlist}
\end {exe}

There are also examples where a switch of constituents does not seem to change the meaning as speakers state that both mean exactly the same, as in (\ref{qualif3}) and (\ref{qualif4}), although in these cases both constituents are clearly nouns which have a plural form and which can be modified by demonstratives and possessor pronouns.

\begin{exe}
\ex\label{qualif3}
\begin{xlist}
\ex\label{qualif3a}
 \gll  nkwɛ̌ (wá) nkpámá  \\
          $\emptyset$3.basket 3:ATT $\emptyset$3.newness  \\
    \trans `new basket'
\ex \label{qualif3b}
  \gll   nkpámá (wá) nkwɛ̌\\
          $\emptyset$3.newness 3:ATT $\emptyset$3.basket \\
    \trans `new basket'
\end {xlist}
\end {exe}

\begin{exe}
\ex\label{qualif4}
\begin{xlist}
\ex\label{qualif4a}
 \gll  m-ùdì (wà) nkángɛ̀\\
          N1-person 1:ATT $\emptyset$3.courage  \\
    \trans `courageous person'
\ex \label{qualif4b}
  \gll   nkángɛ̀ (wá) m-ùdì\\
          $\emptyset$3.courage 3:ATT N1-person \\
    \trans `courageous person'
\end {xlist}
\end {exe}


In predicative use, the adjective serves as the copula complement as shown in (\ref{qualif5}). 

\begin{exe}
\ex\label{qualif5}
\gll m-àmbɔ̀ máà {\bfseries mpà} \\
	ma6-thing 6:COP good\\
    \trans `Things are good.'
\end {exe}

\noindent The adjective clearly shows no agreement morphology, although this would be expected with all plural classes. The same is true for an adjectival complement in a negative non-verbal construction, as in (\ref{qualif6}).

\begin{exe}
\ex\label{qualif6}
\glll  mìnsáyá mí bèyá sâ mí bɛ́lɛ́ {\bfseries mpà}\\
      mi-nsáyá mí bèya-H sâ mi-H bɛ́-lɛ́ mpà\\
	mi4-thing 4:ATT 2\textsc{pl}-PRES do 4-PRES be-NEG good  \\
    \trans `The things that you do are not good.'
\end {exe}


Adjectives can be used as parameters of comparison in comparison constructions, as described in \chapref{sec:CC}.  They are, however, not marked morphologically in these constructions. Finally, they can also be used adverbially to modifiy a verb, as discussed in \sectref{sec:G3ADV}.




Some special remarks are in order for color adjectives. As shown in Table \ref{Tab:QUAL}, all color term adjectives (and the quality adjective {\itshape nátĩ̂}) `straight') have in common that they start with the similative marker {\itshape ná-}, as described in \sectref{sec:NOMSIM}. There is evidence that,  historically, color terms in at least some related languages of this area were verbs. These verbs used for color descriptions then developed into other parts of speech. For instance, in Bulu the basic color terms are synchronically nouns: {\itshape évìndì} `black', {\itshape évèlè} `red', and {\itshape éfùmùlù} `white'.\footnote{\citet{bates04} gives the verbal color forms for Bulu as follows: {\itshape vé} ‘be/get red’, {\itshape vìn} ‘be/get black’, and {\itshape fùm} ‘be white’ without mentioning any nominal color forms. \citet[44]{alexandre55} explains that these verbs can take a causative suffix {\itshape vìn} `be black' $\rightarrow$ {\itshape vìn-ì} `make black'. These causative verbs were then nominalized and assigned to noun class 5 with the prefix {\itshape é-}. \citet[68]{alexandre55} states that this class usually hosts deverbal nouns derived from stative verbs.} In Gyeli, it is likely that such color verbs were grammaticalized, together with the {\itshape ná} similative marker, into a synchronic uninflected element of the noun phrase.

Another argument that color adjectives are grammaticalized verbs including a similative marker comes from the atypical terms {\itshape nápfûpfû} `darkened color' and {\itshape náyɛ̂yɛ} `lightened color' which describe a change of color as opposed to a specific hue. When asked for the meaning of these atypical colors, speakers give a verbal explanation, namely that a more prototypical color such as `black', `white', or `red' has changed by either having become darker ({\itshape nápfûpfû}) or lighter, being `bleached out' ({\itshape náyɛ̂yɛ}). In contrast, other colors are referred to by French adjectives in explanations.

According to traditional color theories, these two special color terms are unusual in that they do not fit into basic color words that have been investigated cross-linguistically (see, for instance, \citealt{berlin69}). 
Nevertheless, I classify {\itshape nápfûpfû} `darkened color' and {\itshape náyɛ̂yɛ} `lightened color' as color terms since they only show up in discourse when talking about colors and they were systematically used by speakers in the color booklet task \citep{majid2007a}.\footnote{Gyeli has more color terms than the adjectives listed in Table \ref{Tab:QUAL}. Other color terms include, for instance, {\itshape mpùlɛ́} `yellow', which is derived from the name of a tree with yellow bark ({\itshape Enantia chlorantha}), or {\itshape màká} `green' which is a noun which also means `leaves'. Those other color terms are, however, recently acquired and differ in their morphosyntactic status in that they are nouns rather than adjectives, as further explained in \citet{grimm2014}.}







\section{Adverbs}
\label{sec:ADV}

%GIVE SHORT VERSION OF ADVERB SECTION AND THINK ABOUT HOW TO DISTRIBUTE INFO IN SENSIBLE WAY

%(add lii)

%ná still, again, anymore: change from sentential modifier to adverb??? 




Adverbs, along with nouns, verbs, and adjectives, constitute an open part-of-speech class. According to \citet[20]{schachter2007}, adverbs may have various subclasses, such as directional adverbs (`down'), degree adverbs (`extremely'), manner adverbs (`quickly'), time adverbs (`today'), or sentence adverbs (`unfortunately'). These subclasses show that adverbs do not necessarily modify verbs, but may also modify adjectives or other adverbs or even whole sentences. \citet[20]{schachter2007} thus provide a broad definition of adverbs as elements which ``function as modifiers of constituents other than nouns.'' 

In general, the class of adverbs in Gyeli is rather restricted in diversity, just as in many other Bantu languages. Thus, in the Gyeli text corpus, as described in \sectref{sec:Data},
 fewer than 20 different adverbs occurred. One reason for this is that, according to  \citet[126]{creissels2008}, in many African languages, ``the possibility of deriving manner adverbs from other categories or to use adjectives as verb modifiers, is very limited.''  This is also true for Gyeli where the meaning of typical English manner adverbs is instead expressed by ideophones, as will be discussed in \sectref{sec:IDEO}, or by nouns in complement position, as in (\ref{maNCOMP}).


\begin{exe} 
\ex\label{maNCOMP}
  \glll     màlɛ́ndí máà vɛ̀ɛ̀ kwè mípìndí  \\
	ma-lɛ́ndí máà vɛ̀ɛ̀ kwè H-mi-pìndí \\
              ma6-palm.nut 6.DEM.PROX only fall OBJ.LINK-mi4-non.ripe   \\
    \trans `These palm nuts only fall non-ripe.'
\end{exe}

\noindent Despite this restricted diversity, Gyeli adverbs occur pervasively in all types of text genres (dialogues, folktales, autobiographic narratives). Almost a quarter of all intonation phrases in the Gyeli text corpus (123 (23\%) of 540 intonation phrases) include an adverb.

Gyeli adverbs are invariable and do not receive any specific morphological marking, e.g.\ through suffixes, like the English -{\itshape ly} or French -{\itshape ment}. Subclasses of adverbs can be distinguished through several morphosyntactic properties and/or a combination of them. I will consider the following three subclasses as described by their most salient characteristics:

%\begin{table} 
\begin{center}
\begin{tabular}{ll}
Group 1: & adverbs optionally combining with LOC preposition {\itshape ɛ́} \\
Group 2: & adverbs that can occur in noun + attributive constructions \\
Group 3: & adverbial lexemes that can act as nominal modifiers in NPs  \\
%Group 4: & adverbial lexemes that occur as nouns in NPs \\
%Group 5: & adverb that does not exhibit any of the above mentioned properties \\
\end{tabular}
\end{center}
%\caption{Main characteristics of adverb groups}
%\label{Tab:ADV1}
%\end{table}

Subclassification of adverbs in the literature is typically done on a semantic basis, such as manner, temporal or locative adverbs. The choice of semantic categories may, however, be arbitrary and may not match the morphosyntactic categories of a language. In Gyeli, morphosyntactic classes map onto semantic categories, as shown in Table \ref{Tab:ADV2}.  Group 1 consists entirely of deictic adverbs which include locative and manner deictics.  Group 2 hosts temporal adverbs and group 3 contains manner adverbs.%, and group 4 locative and directional adverbs. Group 5 only has one member, namely an anaphoric adverb. 

Nevertheless, the defining criteria for adverbial subclasses in Gyeli are four morphosyntactic properties as listed in the column names of Table \ref{Tab:ADV2}: i) the potential combination with the locative {\itshape ɛ́}, ii) use of a lexeme as both adverb modifying a verb and adjective/quantifier modifying a noun, iii) occurrence in noun + attributive marker construction, and iv) occurrence in phrase final position only. The last column also provides information on the derivational source of the adverbs. Yet, since this is not a morphosyntactic property, it does not determine adverbial classification.


\begin{table} 
\centering
\scalebox{0.9}{
\begin{tabular}{ll|cccc||l}
Group & Semantic & LOC {\itshape ɛ́} & ADJ / & ATT constr. & final posi- & derivational   \\ 
           & core         &                   & QUANT &                   & tion only    & source         \\     \midrule
       1  &      deictic  &      x           & ---          &      (x)          &  ---         & underived    \\
       2a  &      temporal &   ---          & ---         &     x            & ---          & underived   \\
       2b  &      temporal &   ---          & ---         &     x            & ---          & denominal  \\
       3  &      manner &    ---           &  x          &      ---         & x           &  ADJ/QUANT\\ 
\end{tabular}}
\caption{Criteria for adverb classification}
\label{Tab:ADV2}
\end{table}

The distinctive characteristic of group 1 adverbs is their potential combination with the locative preposition {\itshape ɛ́} which no other adverbial subclass allows for. Also, some (but not all) group 1 adverbs can be used in noun + attributive marker constructions. This property is defining for group 2 adverbs. Group 3 adverbs are the only ones to be restricted to a phrase final position only while all other adverbs can also occur at the beginning of a phrase. Lexemes occurring in group 3 can also be used as adjectives or quantifiers to modify nouns. % In contrast, lexemes in group 4 cannot be used in nominal modification, but they can be used as nouns in noun phrases. Finally, the anaphoric adverb of group 5 is defined by the absence of all four morphosyntactic properties. 

%In the following, I will describe each adverb subclass in more detail.
%In order to be consistent with the structure of this grammar, I will only discuss adverbs that modify verbs in this section. Note that I treat words such as `also', `still', and `only' separately in \sectref{sec:SentMod} since they behave as modifiers on a clausal level and, as such, show greater positional variability.


\subsection{Group 1 adverbs: deictic} 
\label{sec:G1ADV}

Adverbs of group 1 are all deictic in nature, including both locative and manner deixis. They are the most frequent ones occurring in natural text out of all adverb types. 
Deictic adverbs, as any deictic elements, are often accompanied by showing gestures or assume  common knowledge of the specific place under discussion. Table \ref{Tab:DeicAdv} provides a summary of deictic adverbs in Gyeli as well as their numeric frequency in the Gyeli text corpus.\footnote{Obviously, this is a very limited corpus, but it shows some tendency as to which adverb gets used more frequently.} The deictic elements represented in the table mostly function as adverbs, namely when they occur with verbs, but as the last column shows, almost all of them may also occur in the nominal domain modifying nouns.  \sectref{sec:LOCe} provides more information on the locative {\itshape ɛ́}.

\begin{table} 
\centering
\begin{tabular}{llll}
Deictic element & Gloss & \multicolumn{2}{c}{Frequency} \\ 
                        &            &        with verb  & with noun \\  \midrule
(ɛ́) vâ & `here' & 41 & 2 \\
(ɛ́) pɛ̀ & `over there' & 21 & 0 \\
(ɛ́) wû & `there' & 12 & 3 \\ 
% wɛ̂ & `there' \\ put into noun deictic, maybe with tè
(ɛ́) tè & `there' & 8 & 13 \\
%(ɛ́) mpù  & `like this' &  14   & 0    \\
%bà & `to, {\itshape chez}' & 2 \\
\end{tabular}
\caption{Deictic adverbs}
\label{Tab:DeicAdv}
\end{table} 



\paragraph{Formal commonalities}

I view deictic adverbs as a category, based on formal similarity and their potential co-occurrence with the locative marker {\itshape ɛ́}, which distinguishes them from other adverb subclasses. 
All deictic adverbs are monosyllabic. They do not seem to be derived from another part of speech, in contrast to, for instance, group 3 adverbs. Some of them may, however, also be used to modify nouns rather than verbs, namely as the second constituent in noun + attributive marker constructions, as discussed in \sectref{sec:CONC}.  The distribution of deictic adverbs as modifying verbs as opposed to nouns is illustrated in Table \ref{Tab:DeicAdv} under `Frequency'. (\ref{DeicNom}) gives an example of a deictic element as nominal modifier while the examples in the remainder of this section show deictic adverbs modifying verbs.


\begin{exe} 
\ex\label{DeicNom} 
  \glll mɛ̀gà mɛ́ɛ̀ dyúwɔ́ nzã́ã̀ [dúwɔ̀ lé {\bfseries tè}] \\
        mɛ-gà mɛ́ɛ̀ dyúwɔ-H nzã́ã̀ d-úwɔ̀ lé tè \\
         1\textsc{sg}-CONTR 1\textsc{sg}.PST2 feel-R $\emptyset$7.appetite le5-day 5:ATT there \\
    \trans `As for me, I felt appetite the day there.'
\end{exe}

\noindent Contrasting deictics as verbal versus nominal modifiers, there is a tendency that the more frequently a (locative) deictic element occurs as verbal modifier, the less frequently it is found as a nominal modifier. This is the case, for instance, with {\itshape vâ} `here'. Within the Gyeli text corpus, {\itshape vâ} is found 41 times as a verbal, but only twice as a nominal modifier.  Conversely, the less frequently a deictic adverb modifies verbs, the more often it occurs as nominal modifier as with {\itshape tè} `there' which occurs only 8 times with verbs, but 13 times with nouns. 

%
\iffalse
The manner deictic {\itshape (ɛ́) mpù} never occurs as nominal modifier.  It generally serves to introduce gestures and ideophones, as for instance in (\ref{ADVmpu1}). In this example, the first occurrence of {\itshape mpù} frames the ideophone while the second refers to a gesture that may actually be made or not, i.e.\ the gesture is most likely implied, but not necessarily made.


\begin{exe} 
\ex\label{ADVmpu1}
  \glll yɔ́ɔ̀ Nzàmbí njí {\bfseries mpù} bã̂ã̂ã̂ã̂ njì dígɛ̀ {\bfseries mpù} \\
       yɔ́ɔ̀ Nzàmbí njî-H mpù bã̂ã̂ã̂ã̂ njì dígɛ̀ mpù \\
        so $\emptyset$1.PN.PRES come-R like.this IDEO:walking.far come look like.this \\
    \trans `So Nzambi comes like this [depiction of walking a long distance], comes looking like this.'
\end{exe}

\noindent   Also, {\itshape mpù} is used in comparison constructions as in (\ref{ADVmpu2}). In these cases, {\itshape mpù} is translated as `like, than' rather than `like this'.

\begin{exe} 
\ex\label{ADVmpu2}
  \glll Màmá à ndáà gyà ntɛ̀ {\bfseries mpù} Màmbì \\
       Màmá a ndáà gyà ntɛ̀ mpù Màmbì.  \\
        $\emptyset$1.PN  1.have also $\emptyset$7.length $\emptyset$3.size like $\emptyset$1.PN \\
    \trans `Mama is as tall as Mambi.'
\end{exe}
\fi %

\paragraph{Phrase position} Another distinctive morphosyntactic property in adverbial subclasses is the phrase position in which adverbs can occur. As a default position, all adverb classes occur phrase finally. This is also true for group 1 adverbs, as shown in (\ref{ADV1finala}) and (\ref{ADV1finalb}).

\begin{exe} 
\ex\label{ADV1finala}
  \glll    mɛ́ bvú nâ nkwálá wúù tfùndɛ́ mɛ̀ {\bfseries vâ} \\
           mɛ-H bvû-H nâ nkwálá wúù tfùndɛ-H mɛ̀ vâ \\
              1\textsc{sg}-PRES think-R COMP $\emptyset$3.machete 3.PST2 miss-R 1\textsc{sg} here \\
    \trans `I think that the machete missed [= injured] me here.'
\end{exe}

\begin{exe} 
\ex\label{ADV1finalb}
  \glll mɛ́ pã́ ná kɛ̀ dígɛ̀ mùdì wà nû {\bfseries ɛ́} {\bfseries pɛ́} \\
        mɛ-H pã̂-H ná kɛ̀ dígɛ m-ùdì wà nû ɛ́ pɛ́ \\
        1\textsc{sg}.PRES try-R again go see N1-person 1:ATT 1.DEM.PROX LOC there \\
    \trans `I try again and go see that person there.'
\end{exe}


In contrast to group 3, group 1 adverbs also pervasively appear in phrase initial positions, as in (\ref{ADV1initiala}) and (\ref{ADV1initialb}). This position is clearly correlated with information structure, moving the deictic adverb into a focus position.\footnote{See \sectref{sec:IS} on information structure for a more detailed discussion.} While also group 2 (temporal) adverbs can occur in this initial focus position, deictic adverbs are significantly more frequently focussed in the Gyeli text corpus.

\begin{exe} 
\ex\label{ADV1initiala}
  \glll  {\bfseries ɛ́} {\bfseries vâ} mɛ̀ dyùwɔ́ nâ {\bfseries ɛ́} {\bfseries vâ} yíì sílɛ̀ njì búlɛ̀  \\   
         ɛ́ vâ mɛ dyùwɔ-H nâ ɛ́ vâ yíì sílɛ̀ njì búlɛ  \\
         LOC here 1\textsc{sg}.PST1 hear-R COMP LOC here 7.FUT finish.FUT come destroy  \\
    \trans `Here I heard that here it will all come to be destroyed.'
\end{exe}

\begin{exe} 
\ex\label{ADV1initialb} 
  \glll  {\bfseries ɛ́} {\bfseries pɛ́-ɛ́} mɛ̀ɛ̀ lwɔ̃̂ nyà ndáwɔ̀ \\
        ɛ́ pɛ́-ɛ́ mɛ̀ɛ̀ lwɔ̃̂ nyà ndáwɔ̀  \\
          LOC there-DIST 1\textsc{sg}.FUT build real $\emptyset$9.house \\
    \trans `Over there, I will build a real house.'
\end{exe}



\noindent If a deictic adverb occurs in the initial focus position, it is often repeated again at the end of the phrase in its default position, as shown in (\ref{ADV1botha}) and (\ref{ADV1bothb}).


\begin{exe} 
\ex\label{ADV1botha} 
  \glll {\bfseries ɛ́} {\bfseries pɛ̀} bà sílɛ́ bî lwɔ̃̂ mándáwɔ̀ {\bfseries ɛ́} {\bfseries pɛ̀}  \\
        ɛ́ pɛ̀ ba sílɛ-H bî lwɔ̃̂ H-ma-ndáwɔ̀ ɛ́ pɛ̀  \\
        LOC there 2.PST1 finish-R 1\textsc{pl}.NSBJ build OBJ.LINK-ma6-house LOC there  \\
    \trans `There, they have finished to build us houses there.'
\end{exe}


\begin{exe} 
\ex\label{ADV1bothb}
  \glll {\bfseries ɛ́} {\bfseries wû} bèyá lwɔ̃́ kwádɔ́ yã̂ {\bfseries ɛ́} {\bfseries wû}  \\
         ɛ́ wû bèyá lwɔ̃̂-H kwádɔ́ y-ã̂ ɛ́ wû  \\
         LOC there 2\textsc{pl}[Kwasio] build-R $\emptyset$7.village 7-POSS.1\textsc{sg} LOC there\\
    \trans `Over there you (pl) build my village over there.'
\end{exe}


The use of the locative {\itshape ɛ́} is more frequent when the adverb occurs phrase initially while post-verbal and  phrase final occurrences allow for a higher degree of optionality whether to use the locative or not. 
The higher degree of locative {\itshape ɛ́} omission when the deictic adverb occurs phrase finally might be phonologically conditioned. Phrase finally, the locative {\itshape ɛ́} usually follows a vowel either from a preceding verb or noun and may undergo deletion in fast speech. When asked, speakers state that the use of the locative {\itshape ɛ́} is possible in both phrase initial and phrase final positions. It is less clear at this point whether the co-occurrence of the locative {\itshape ɛ́} with a deictic adverb is generally optional, comparable to the optional use or omission of the attributive marker as discussed in \sectref{sec:CONOM} or whether the locative {\itshape ɛ́} is always underlyingly present with deictic adverbs and its omission in the surface form is purely phonological. 



\paragraph{Distinctions within the locative deictic system} 
Gyeli uses a range of deictic elements to refer to places or locations in varying distance to the speaker. Since most of these elements would be translated as `there' in English, the system merits a more thorough explanation. In general, distances in Gyeli are relative rather than absolute in that `here', for instance,  can denote a place within a hand-reach of the speaker, but could also talk about a whole village. On the other hand, `over there' can then be a distant place or, in other cases,  a place even within  the village, depending on the discourse topic. 

Semantically, the clearest distinction is between {\itshape vâ} `here', which refers to the relative immediate surroundings of the speaker, and {\itshape pɛ̀} `over there', which denotes the place furthest away. In French, {\itshape pɛ̀} gets translated as {\itshape là-bas}. {\itshape wû} and {\itshape tè} would both be translated as `there', or {\itshape là} in French, which makes it more difficult to grasp their semantic distinctions. Differences in their morphosyntactic behavior can help to disentangle their meaning contrast.

In the default case, it seems that {\itshape wû} denotes a medial distance between {\itshape vâ} `here' and {\itshape pɛ̀} `over there' and occurs mainly in the verbal domain. In contrast, {\itshape tè} is mostly used with nouns rather than with verbs where {\itshape tè} seems to be related more to specificity and/or anaphora than to actual location. In that sense, {\itshape tè} may be less part of the distance-related deictic system, as (\ref{Advte1}) illustrates. In this example, {\itshape tè} is more existential than about distance.


\begin{exe} 
\ex\label{Advte1}
  \glll     bã̂ yɔ́ɔ̀ yíì {\bfseries tè} \\
	bã̂ y-ɔ́ɔ̀ yíì tè \\
              $\emptyset$7.word 7-2\textsc{sg}.POSS  7.ID there  \\
    \trans `Your word is there [= you are understood].'
\end{exe}

\noindent Also in (\ref{Advte2}), the use of {\itshape tè} is not primarily locative, but more anaphoric to the circumstances of earning only 250 Francs CFA.

\begin{exe} 
\ex\label{Advte2}
  \glll ká bá kɛ́ wɛ̂ vɛ̀ bé-bwúyà  bébáà nà mà-wú mátánɛ̀ wɛ́ sá {\bfseries tè} ná  \\
        ká ba-H kɛ̀-H wɛ̂ vɛ̀ H-be-bwúyà  bé-báà nà ma-wú má-tánɛ̀ wɛ-H sâ-H tè ná  \\
        if 2-PRES go-R 2\textsc{sg}.NSBJ give OBJ.LINK-be8-hundred 8-two COM ma6-ten 6-five 2\textsc{sg}-PRES do-R there how  \\
    \trans `If they go give you 250 (Francs), how do you manage there? [because it's very little money]'
\end{exe}

In other cases, however, as in (\ref{Advte3}), {\itshape tè} is place-denoting just like the other deictic adverbs. Speakers state that, in this example, {\itshape tè} can also be replaced by {\itshape pɛ̀} or {\itshape wû}. 

\begin{exe} 
\ex\label{Advte3}
  \glll     {\bfseries tè} mɛ̀ɛ̀ jíbì kɛ̀ lwɔ̃̂ {\bfseries tè} \\
	tè mɛ̀ɛ̀ jíbi kɛ̀ lwɔ̃̂ tè \\
             there 1\textsc{sg}.FUT start go build there  \\
    \trans `There, I will first go to build there.'
\end{exe}

Further, distance cannot be the only distinctive criterion within the locative deictic system: An increased sense of distance can be added phonologically by lengthening the final vowel of the adverb and a H tone, as shown in (\ref{pEE}) and in (\ref{ADV1initialb}) above.


\begin{exe} 
\ex\label{pEE}
  \glll     lèkfúdɛ̀ à nzí bíyɔ̀ nlô {\bfseries pɛ́ɛ́}  \\
	le-kfúdɛ̀ a nzí bíyɔ nlô pɛ́-ɛ́ \\
              le5-idiot 1 PROG.PST hit $\emptyset$3.head over.there-DIST  \\
    \trans `The idiot was hitting his head far over there.'
\end{exe}

\noindent This way of expressing further distance by vowel lengthening and H tones is possible with both {\itshape pɛ̀} and {\itshape wû}. An example for the latter is given in (\ref{wUU}). In contrast, this does not seem to be possible with {\itshape tè} which indicates again that {\itshape tè} behaves differently from the other more purely locative deictic elements.\footnote{{\itshape vâ} `here' also does not allow for final vowel lengthening and a H tone, but that is clearly a semantic restriction since it denotes a place that is close to the speaker.}

\begin{exe} 
\ex\label{wUU}
  \glll  báà tfùbɔ̀ báà tfùbɔ̀ mpàgó wá nùmbà {\bfseries wúú}\\
         báà tfùbɔ̀ báà tfùbɔ̀ mpàgó wá nùmbà wú-ú\\
          2.FUT pierce 2.FUT pierce $\emptyset$3.road 3:ATT $\emptyset$1.logger there-DIST \\
    \trans `They will cut, they will cut. The road of the loggers there.'
\end{exe}

Another difference between {\itshape wû} and {\itshape tè} concerns the combination with a vocative morpheme -{\itshape o} which, at the same time, can further take a H tone to indicate distance between the speaker and the addressee. This vocative morpheme can be used with {\itshape wû}, as shown in (\ref{wuo}), but not with {\itshape tè} nor any other deictic element.

\begin{exe} 
\ex\label{wuo} 
  \glll mùdì kí tàtɔ̀ {\bfseries wúó} \\
        m-ùdì kí tàtɔ wú-o-H \\
        N1-person NEG scream there-VOC-DIST \\
    \trans `Nobody scream over there!'
\end{exe}

In summary, it seems that {\itshape vâ} `here', {\itshape wû} `there' and {\itshape pɛ̀} `over there' form the core locative deictic system while {\itshape tè} `there' takes over other functions (specificity, anaphora) as a default, but can also act as a deictic element within the locative system. The different properties of the various locative deictics as discussed above are summarized in Table \ref{Tab:LOCdeic}.


\begin{table} 
\centering
\scalebox{0.9}{
\begin{tabular}{ll|clcc}
Deictic & Gloss & LOC {\itshape ɛ́} & mostly modifying & DIST marking &  Vocative -{\itshape o}  \\  \midrule
vâ       & `here'  &        x         &     verbal              &         ---           &      --- \\
wû      & `there' &        x         &      verbal             &           x           &       x  \\
pɛ̀       & `over there'  &  x      &      verbal            &            x           &      --- \\
tè        & `there'  &       x         &      nominal          &           ---          &     --- \\
\end{tabular}}
\caption{Morphosyntactic properties of locative deictics}
\label{Tab:LOCdeic}
\end{table} 





\subsection{Group 2 adverbs: temporal}
\label{sec:G2ADV}

Adverbs of group 2 have four members which are all temporal and listed in Table \ref{Tab:ADVGroup2}. While group 2 adverbs form a unitary morphosyntactic category, they differ in their derivational source. While {\itshape tɛ́ɛ̀} `now' and {\itshape dẽ̂} `today' seem to be underived lexemes, the other two adverbs in the group are clearly derived from nouns: {\itshape nàkùgúù} `yesterday' is derived from {\itshape kùgúù} `evening' and {\itshape nàmɛ́nɔ́} `tomorrow' from {\itshape mɛ́nɔ́} `morning'. The {\itshape nà}- prefix in these adverbs is a derivational similative marker, as described in \sectref{sec:SIM}.


\begin{table} 
\centering
\begin{tabular}{ll|l}
Adverb & Gloss & Derivational source \\  \midrule
tɛ́ɛ̀ & `now' & underived \\
dẽ̂ & `today'  & underived\\
nàkùgúù & `yesterday' & denominal \\
nàmɛ́nɔ́ & `tomorrow' & denominal \\
\end{tabular}
\caption{Group 2 adverbs}
\label{Tab:ADVGroup2}
\end{table} 


The defining property of group 2 temporal adverbs is that they can all also occur in nominal modification as second constituent in a noun + attributive marker construction, as in (\ref{G2nom}).

\begin{exe}
\ex\label{G2nom}
\begin{xlist}
\ex \label{G2nom1}
  \glll   bèdéwɔ̀ bé {\bfseries dẽ̂} \\
	be-déwɔ̀ bé dẽ̂  \\
           be8-food 8:ATT today    \\
    \trans `food of today.'
\ex\label{G2nom2}
 \glll  nlã̂ wá {\bfseries nàkùgúù} \\
	nlã̂ wá nàkùgúù \\
         $\emptyset$3.story 3:ATT yesterday  \\
    \trans `yesterday's story.'
\end{xlist}
\end{exe}

\noindent While some group 1 adverbs exhibit the same property, deictic adverbs also combine with the locative {\itshape ɛ́}, unlike group 2 temporal adverbs.


All group 2 adverbs occur phrase finally as a default position. Examples are given in (\ref{ADV2a}) through (\ref{ADV2c}).

\begin{exe} 
\ex\label{ADV2a} 
  \glll wɛ́ làwɔ́ {\bfseries tɛ́ɛ̀}  \\
        wɛ-H làwɔ-H tɛ́ɛ̀  \\
         2\textsc{sg}-PRES talk-R now   \\
    \trans `You speak now.'
\end{exe}

\begin{exe} 
\ex\label{ADV2b}
  \glll  nyɛ̀ náà à múà wɛ̂ bíyɔ̀ {\bfseries dẽ̂} \\
        nyɛ náà à múà wɛ̀ bíyɔ̀ dẽ̂\\
           1 COMP 1 PROSP 2\textsc{sg}.NSBJ hit today \\
    \trans `He [says] that he is about to beat you today.'
\end{exe}

\begin{exe} 
\ex\label{ADV2c}
  \glll mɛ̀ nzí kɛ̀ jí {\bfseries nàkùgúù} \\
       mɛ nzí kɛ̀ jí nàkùgúù \\
      1\textsc{sg} PROG.PST go $\emptyset$7.forest yesterday \\
    \trans `I was going to the forest yesterday.'
\end{exe}

 
They can all also occur phrase initially, as shown in (\ref{ADV2d}). In these cases, they are in focus, as discussed for group 1 adverbs and in \sectref{sec:IS} on information structure. In (\ref{ADV2d}), the narrator stresses that the mice will only eat the skulls the next day, as contrastive focus to the possibility that they might eat them right away.


\begin{exe} 
\ex\label{ADV2d}
  \glll àà {\bfseries nàmɛ́nɔ́} bwáà dè {\bfseries nàmɛ́nɔ́} \\
       àà nàmɛ́nɔ́ bwáà dè nàmɛ́nɔ́ \\
       EXCL tomorrow 2\textsc{pl}.FUT eat tomorrow  \\
    \trans `Ah, tomorrow you will eat, tomorrow.'
\end{exe}

\noindent In comparison to group 1 adverbs, which occur frequently in this focus position, group 2 adverbs are rarely found in this position in natural text.



%kɔ́ɔ̀ only, just % 3.23
 %kɔ́ɔ̀ m-ùdì m-vúdũ̂ `seulement une personne'
% kɔ́ɔ̀ b-ùdì bà-vúdû `toujours les même personnes'

%m-ùdì kɔ́ɔ̀ m-vúdũ̂ `seulement une personne'

%nâ more, still, again %3.53

%where does "really" belong to?


\subsection{Group 3 adverbs: manner} 
\label{sec:G3ADV}

Group 3 adverbs are defined by their lexemes' double affiliation to the part of speech of adjectives (\sectref{sec:QUAL}) or nominal modifiers (\sectref{sec:MODAgrPre}). Semantically, they map onto manner adverbs.
Manner adverbs are rare in Gyeli, both in terms of number and occurrence. Table \ref{Tab:ManAdv} gives an exhaustive list of all manner adverbs found in the Gyeli text corpus as well as text stemming from questionnaire elicitation. Each of these manner adverbs occurs only a couple of times in the corpus, thus their natural frequency seems to be generally low. Gyeli seems rather to have a preference to express the manner of an action or event by ideophones, as will be discussed in \sectref{sec:IDEO}.

\begin{table} 
\centering
\begin{tabular}{lll}
Manner adverb & Gloss & Affiliation to other POS \\  \midrule
mpà & 	good & invariable adjective \\
bíwɔ̀   & 	bad & invariable adjective \\
fí   & 	        different  & deictic modifier  (→ short form of -fúsì) \\
bvùbvù    &       a lot  & invariable quantifier \\ 
\end{tabular}
\caption{Manner adverbs and their affiliated parts-of-speech}
\label{Tab:ManAdv}
\end{table} 

%All of these manner adverbs are also found as nominal modifiers where they differ though in their behavior, as shown in \sectref{sec:MOD} and \ref{sec:INV}. Most of them such as {\itshape mpà} `good', {\itshape bíwɔ̀} `bad', and {\itshape bvùbvù} `much' are invariable also in noun phrases. Only -{\itshape fí} `different', the short form of -{\itshape fúsì} used as a deictic modifier, agrees with its head noun. In the verbal domain, however, all of them are invariable.

In terms of their position, manner adverbs exclusively occur (intonation) phrase finally. Thus, the adverb may follow the verb if there is no object, as demonstrated in (\ref{Advmpa}) and (\ref{Advfi}).

\begin{exe} 
\ex\label{Advmpa}
  \glll     wɛ̀ nzíí bàlɛ̀ {\bfseries mpà} \\
	wɛ nzíì-H bàlɛ mpà \\
              2\textsc{sg} PROG.PRES-R keep good   \\
    \trans `You are keeping [the words] well.'
\end{exe}


\begin{exe} 
\ex\label{Advfi}
  \glll     wɛ́ ná báàla nà nyɛ́ {\bfseries fí} nà wɛ́ ndyándyá ná sálɛ́ ɛ́ pɛ̂  \\
	wɛ-H ná báàla-H nà nyɛ̂-H fī́ nà wɛ-H ndyándya-H ná sálɛ-H ɛ́ pɛ̂ \\
              2\textsc{sg}-PRES again repeat-R COM see-R different COM 2\textsc{sg}-PRES work-R again $\emptyset$7.work LOC there   \\
    \trans `You repeat again and see differently [try something else] and you do again work there.'
\end{exe}

\noindent If the clause has an object, the manner adverb will follow the object instead of the verb, as shown in (\ref{Advbiwo}) and (\ref{Advbvu}).

\begin{exe} 
\ex\label{Advbiwo}
  \glll   á sìmbɔ́ màtúà {\bfseries bíwɔ̀}   \\
	a-H sìmbɔ-H màtúà bíwɔ̀ \\
              1-PRES drive-R $\emptyset$1.car bad   \\
    \trans `He drives the car poorly.'
\end{exe}

\begin{exe} 
\ex\label{Advbvu}
  \glll     mɛ̀ɛ́ jí-lɛ́ wɛ̂ {\bfseries bvùbvù} \\
	mɛ̀ɛ́ jí-lɛ́ wɛ̂ bvùbvù \\
              1\textsc{sg}.PRES.NEG ask-NEG 2\textsc{sg} much   \\
    \trans `I don't ask you [for] much.'
\end{exe}

\noindent In contrast to adverb groups 1 and 2, manner adverbs cannot be used in a phrase initial focus position.

\iffalse 
\subsection{Group 4 adverbs: locative/directional} 
\label{sec:G4ADV}

These are not adverbs but postpositions or also adverbs?? WHAT about V + sí? noun or adverb? If adverb, change in zero-derivation section***


Just like group 3, group 4 adverbs also have a double affiliation to two parts of speech. In contrast to group 3, however, group 4 adverbs do not serve as nominal modifiers, but are nouns themselves when they occur in a noun phrase. In fact, these adverbs are zero-derived from their corresponding referential nouns. Table \ref{Tab:DirAdv} shows the lexemes and their meaning both in adverbial and in nominal use. 

\begin{table} 
\centering
\begin{tabular}{lll}
Lexeme & Adverbial use & Nominal use \\  \midrule
sí & `under, down' & `ground' \\
 dyúwɔ̀ &  `up, about' & `sky' \\
tɛ́mɔ́ &  `between' & `middle' \\
písɛ̀ &   `behind' & `behind, back (n.)' \\
sɔ̂ & `in front, before' & `front (n.)' \\
\end{tabular}
\caption{Locative/directional adverbs and their source nouns}
\label{Tab:DirAdv}
\end{table}

\noindent Semantically, group 4 adverbs map onto locative and directional adverbs. The same adverb can be used both in a locative and directional way, as for instance {\itshape sí} which can mean both `under' and `down'. 

Given that these lexemes also have a use as nouns in noun phrases, one could argue that group 4 adverbs should not be analyzed as adverbs at all, but rather as oblique nouns. Syntactic and distributional differences show, however, that in Gyeli, group 4 adverbs belong to a different part of speech than their related nouns. First, when referential nouns are used as locations in a phrase, they are preceded by the locative preposition {\itshape ɛ́}, as shown in (\ref{ADV4a}). 

\begin{exe} 
\ex\label{ADV4a}
\glll mɛ̀ lɛ̀bɛ́lɛ́ wɛ̂ {\bfseries ɛ́}  {\bfseries tísònì} nàkùgúù \\
      mɛ lɛ̀bɛlɛ-H wɛ̀ ɛ́ tísònì nàkùgúù \\
     1\textsc{sg}.PST follow-R 2\textsc{sg}.NSBJ LOC $\emptyset$7.town yesterday \\
    \trans `I followed you to town yesterday.'
\end{exe}

\noindent In contrast, locative adverbs never co-occur with the locative {\itshape ɛ́}, as exemplified in (\ref{ADV4b}).

\begin{exe} 
\ex\label{ADV4b}
\glll mɛ̀ búgɛ́ wɛ̂ {\bfseries sí} nàkùgúù  \\
      mɛ búgɛ-H wɛ̀ sí nàkùgúù \\
     1\textsc{sg}.PST put.down-R 2\textsc{sg}.NSBJ down yesterday \\
    \trans `I put you down yesterday.'
\end{exe}

Second, locatives in adverbial use do not exhibit typical noun characteristics such as triggering agreement. The use of locative adverbs is invariable in (\ref{ADV4c}), i.e.\  the adverbial cannot be extended by, for instance, an attributive construction.

\begin{exe} 
\ex\label{ADV4c}
\glll nsɔ̃́ wɔ́ɔ̀ mɛ̂ {\bfseries sí/písɛ̀/sɔ̂}. \\
      nsɔ̃́ wɔ́ɔ̀ mɛ̀ sí/písɛ̀/sɔ̂ \\
      $\emptyset$3.worm 3:ID 1\textsc{sg}.NSBJ under/behind/in.front \\
\trans `The worm is under/behind/in front of me.'
\end{exe}

\noindent The minimal pair to this phrase in (\ref{ADV4d}) shows that also nouns can be used in this phrase, but in that case, the syntactic structure changes. The locative noun is followed by an attributive marker which agrees with the noun and thus clearly marks it as a noun. Further, the object {\itshape mɛ̀} `1\textsc{sg}' that precedes the adverbial locative follows the nominal locative. 

\begin{exe} 
\ex\label{ADV4d}
\glll nsɔ̃́ wɔ́ɔ̀ {\bfseries sí/písɛ̀/sɔ̂} {\bfseries yá} mɛ̂. \\
      nsɔ̃́ wɔ́ɔ̀ sí/písɛ̀/sɔ̂ yá mɛ̀ \\
      $\emptyset$3.worm 3:ID under/behind/in.front 7:ATT 1\textsc{sg}.NSBJ \\
\trans `The worm is under/behind/in front of me.'
\end{exe}

Comparable to group 3 manner adverbs, group 4 adverbs occur only phrase finally, as shown in (\ref{ADV4b}), (\ref{ADV4c}), and (\ref{ADV4e}). They do not appear in the phrase initial focus position.

\begin{exe} 
\ex\label{ADV4e} 
  \glll à kɛ̃́ɛ̃̀ nyî pɛ̀ {\bfseries dyúwɔ̀}, à dígɛ́ɛ̀ à díg-â dígɛ́ɛ̀. \\
       a kɛ̃́ɛ̃̀ nyî pɛ̀ dyúwɔ̀ a dígɛ́ɛ̀ a díg-â dígɛ́ɛ̀. \\
       1.PST1 go.PRF enter there on.top 1.PST1 watch.PRF 1.PST1 watch-1.PST1 watch.PRF \\
    \trans `He had gone and entered there on top, and watched and watched and watched.'
\end{exe}

Finally, group 4 adverbs can also modify phrases rather than verbs, as shown in (\ref{ADV4f}). %[THIS SHOULD GO IN CLAUSE CHAPTER, BUT WHERE?]

\begin{exe} 
\ex\label{ADV4f}
  \glll mbúmbù wã̂ wɛ́ kúmbɔ́ mɛ̂ sá mpù  ɛ́ mwánɔ̀ wɑ̃̂ {\bfseries dyúwɔ̀}\\
        mbúmbù w-ã̂ wɛ-H kúmbɔ-H mɛ̂ sá mpù ɛ́ m-wánɔ̀ wɑ̃̂ dyúwɔ̀\\
        $\emptyset$1.namesake 1-POSS.1\textsc{sg} 2\textsc{sg}-PRES arrange-R 1\textsc{sg}.NSBJ $\emptyset$7.thing like.this EXCL N1-child 1-POSS.1\textsc{sg} on \\
    \trans `My namesake, you do a thing to me like this. Hey, about [concerning] my child!'
\end{exe}





\subsection{Group 5 adverbs: anaphoric} 
\label{sec:G5ADV}

The fifth adverbial sublass is exceptional in that it is basically defined by negative values concerning the defining morphosyntactic features: it does not combine with the locative {\itshape ɛ́}, it does not have a double affiliation to another part of speech nor is it used as nominal modifier, and it is not restricted to a phrase final position. Group 5 only has one member which is {\itshape ndɛ̀náà} `like this'.

{\itshape ndɛ̀náà} `like this' is derived from the anaphoric marker {\itshape ndɛ̀} plus the interrogative {\itshape ná} `how'. Phonologically and following speaker intuition, this adverb is considered as one word. The inclusion of the anaphoric marker hints at the difference to the manner deictic {\itshape mpù} of adverb group 1 which receives the same translation. Instead of framing gestures and ideophones like {\itshape mpù}, {\itshape ndɛ̀náà} mainly serves as an anaphoric adverb, summarizing in some sense the previous discourse.\footnote{{\itshape mpù} is significantly more frequent in natural texts with 23 occurrences in the Gyeli corpus, contrasting with only 4 occurrences of {\itshape ndɛ̀náà}.} This is illustrated in (\ref{ADV5a}) through (\ref{ADV5c}). For example, in (\ref{ADV5a}), {\itshape ndɛ̀náà} refers to the event chain of collecting money and giving it to the person the money is owed to. 

\begin{exe} 
\ex\label{ADV5a}
  \glll  yɔ́ɔ̀ mɛ́ tɔ́kɛ́ mɔ̀nɛ́ wɛ̀ vɛ̀ nyɛ̂ nâ {\bfseries ndɛ̀náà}  yíì mpà\\
        yɔ́ɔ̀ mɛ-H tɔ́kɛ-H mɔ̀nɛ́ w-ɛ̀ vɛ̀ nyɛ̂ nâ ndɛ̀náà  yíì mpà  \\
         so 1\textsc{sg}-PRES collect-R $\emptyset$1.money 1-POSS.3\textsc{sg} give 1.NSBJ COMP like.this 7.ID good \\
    \trans 'So I collect her money [and] give [it to] her, so that like this it be good.'
\end{exe}

\noindent In (\ref{ADV5b}), the anaphoric adverb summarizes the previous events of leaving her child to the person {\itshape Nzambi} in exchange for food and thus returning home and arriving at the river bank `like this', i.e.\ without the child. 

\begin{exe} 
\ex\label{ADV5b}
  \glll  {\bfseries ndɛ̀náà} pámò lé-bũ̂ àá gyì. \\
        ndɛ̀náà pámo H-le-bũ̂ àá gyì. \\
        like.this arrive OBJ.LINK-le5-river.bank 1.INCH cry  \\
    \trans 'Having arrived like this [= without the child] at the river bank she is at the beginning of crying.'
\end{exe}

\noindent {\itshape ndɛ̀náà} can also occur relatively independently in its own intonation phrase, as in (\ref{ADV5c}), referring to the previous discourse.

\begin{exe} 
\ex\label{ADV5c}
  \glll   m̀m̀ {\bfseries ndɛ̀náà} lèkɛ́lɛ̀ lɛ́ndɛ̀ lɛ́ɛ̀ nâ... \\
        m̀m̀ ndɛ́náà le-kɛ́lɛ̀ lɛ́-ndɛ̀ lɛ́ɛ̀ nâ... \\
         EXCL like.this  le5-word 5-ANA 5.ID COMP \\
    \trans `Yes, like this. The word is that...'
\end{exe}

\fi


\subsection{Discussion: multiple adverbs} 
\label{sec:MultiADV}

When discussing the syntactic position of adverbs, I so far referred to phrase initial versus final positions. This, however, only holds if there is only one adverb in the phrase. In phrases with multiple adverbs, adverbs generally still occur after the verb and potentially after an object, but of course not each adverb can occur phrase finally. I therefore suggest that there is a general phrase final slot for adverbs which can be filled with multiple adverbs.

There seem to be some ordering principles within this adverb slot, i.e.\ some adverbs seem to be closer to the center of the phrase than others. Since multiple adverbs do not occur very frequently in natural speech, it is not possible at this point to give a full account of adverb order in multiple adverb constructions. The present examples, however, suggest that group 1 adverbs are closest to the center, i.e.\ verb and following object, as shown in (\ref{ADVM1}) and (\ref{ADVM2}).


\begin{exe} 
\ex\label{ADVM1}
  \gll pílì bèyá lɔ́ njì {\bfseries ɛ̀} {\bfseries vá} {\bfseries tɛ́ɛ̀} {\bfseries dé}  \\
         when 2\textsc{pl} RETRO come LOC here now today  \\
    \trans `When you just arrived here now today,'
\end{exe}

\begin{exe} 
\ex\label{ADVM2} 
  \glll  mɛ̀ nzí dyá {\bfseries vâ} {\bfseries kùgúù} [{\bfseries dẽ̂} màfú mábáà] \\
        mɛ̀ nzí dyá vâ kùgúù dẽ̂ ma-fú má-báà. \\
          1\textsc{sg} PROG.PST1 lie.down here $\emptyset$7.evening today ma6-day 6-two  \\
    \trans `I was here the evening two days ago.'
\end{exe}

\noindent Other generalizations as to whether any of the other adverb subclasses are closer to the center or the periphery of the clause require more investigation. This is most likely also correlated with information structure factors.























\section{Ideophones}
\label{sec:IDEO}


Ideophones are widely attested in the literature on African languages (see, for instance, \citet{doke1935}, who coined the term, \citet{westermann07} on Ewe, \citet{dumestre98} on Bambara, \citet{alexandre66} on Bulu, or  \citet{newman2001} on Hausa) and also found in Gyeli.  In defining the term `ideophone', I refer to \citet[25]{dingemanse2011} who views ideophones as ``marked words that depict sensory imagery'', a definition that deserves some further explanation. First, according to the author, ideophones are often marked by phonological peculiarities and/or stand out from other words by means of ``special word forms, expressive morphology, relative syntactic independence and foregrounded prosody'' (p. 26). Second, the fact that ideophones are words implies that they are ``conventionalized minimal free forms with specifiable meanings.'' Gyeli speakers use ideophones in a conventionalized way able to describe the meaning of single ideophones consistently.\footnote{Ideophones that are identical or similar in their form and meaning seem to be consistently used in the languages of the area either through genealogical affiliation or language contact. In any case, they are easily recognized and understood by speakers of neighboring languages such as Mabi and Bulu.} Third, \citet[27]{dingemanse2011} makes the point that ideophones rather {\itshape depict} than describe their referents. This is similarly explained by \citet[280]{guldemann2008} who notes that ``Metaphorically, one can characterize ideophones as a performance or a gesture in disguise of a word.''
Finally, Dingemanse restricts ideophones to a semantic domain depicting sensory imagery which he views as ``perceptual knowledge that derives from sensory perception of the environment and the body'' (p. 28).
The author argues that this semantic-functional definition makes sense for cross-linguistic comparison while grammatical-structural features of ideophones have to be considered language specifically.  

Gyeli ideophones\footnote{There are 19 occurrences of ideophones in the corpus, comprising 16 different ideophones.} modify verbs in some cases, namely when they behave like adverbs. Even when they are syntactically more independent or occur in  complement clauses, they depict the way an event happens.
Generally, Gyeli ideophones structurally stand out from other words in terms of their phonological shape and their syntactic integration into a phrase.

\subsection{Phonological shape of ideophones} 
\label{sec:IDEOphon}

Ideophones in Gyeli are phonologically  marked by various means, including reduplication or a repetitive character, final vowel lengthening, and special syllable structure such as closed syllables or syllables consisting of a consonant only. These three properties usually do not all occur in the same ideophone, but are partially mutually exclusive. For instance, final vowel lengthening excludes the possibility of a closed syllable. Also, reduplication does not usually  occur with final vowel lengthening while closed syllable ideophones may also be reduplicated. 

\paragraph{Reduplication/repetitive character}

Many Gyeli ideophones involve reduplication or repetition, where a word is minimally reduplicated. In most cases, however, the word gets repeated multiple times, i.e.\ more than twice, usually three to five or six times, depending on the ideophone and the dramatic effect aimed at in the discourse. 
For all repetitive ideophones it holds that the number of repeated syllables is not necessarily conventionalized. Each ideophone seems to have a preference for the number of repetitions as represented in the following examples, but the number is not fixed.

Repetitive ideophones can be divided into those that have the same tone on each repeated syllable and those that change their tonal melody across repeated syllables. In (\ref{IDEOredup1}), for instance, the ideophones involve repeated monosyllabic words each carrying the same tone. 


\begin{exe} \ex \label{IDEOredup1}
\begin{tabular}{lp{8.5cm}}
ʃyɛ̂ ʃyɛ̂ & `depiction of sneaking' \\
tʃɔ̀p tʃɔ̀p tʃɔ̀p & `depiction of dripping sound or sound walking in mud'  \\
mtʃà mtʃà mtʃà & `depiction of picky eating (only taking certain items off a plate)' \\
kɛ́ kɛ́ kɛ́ kɛ́ kɛ́ & `depiction of placing objects in a row' \\
tsùk tsùk tsùk tsùk & `depiction of noise that mice make' \\
\end{tabular}
\end{exe} 

In contrast, the ideophones in (\ref{IDEOredup2}) show an alternating tonal pattern with repeated monosyllabic words alternating between H and L tones. One could argue that two syllables, a H plus a L, actually  constitute one unit that gets repeated rather than the single syllable. The fact that these ideophones are often used with an uneven number of syllables, however, indicates that also for tonally alternating ideophones the repeated unit is usually the monosyllabic word.


\begin{exe} \ex \label{IDEOredup2}
\begin{tabular}{lp{8.5cm}}
gbĩ́  gbĩ̀ gbĩ́  gbĩ̀  gbĩ́& `depiction of small objects moving in space (e.g.\ bacteria roaming in a body)' \\
wùù wúú wùù wúú & `depiction of sound of bees'  \\
\end{tabular}
\end{exe}

There are a few instances, however, where the word is bisyllabic and again, it is the word that gets reduplicated, as shown in (\ref{IDEOredup3}). In contrast to monosyllabic ideophone words, bisyllabic ones are only subject to reduplication, but usually do not get repeated more than twice.


\begin{exe} \ex \label{IDEOredup3}
\begin{tabular}{lp{8.5cm}}
kpúdùm kpúdùm & `depiction of drumming' \\ 
kpàdà kpàdà & `depiction of drumming on bamboo pipes' \\
mátʃà màtʃà & `depiction of eating in little bits'  \\
\end{tabular}
\end{exe}


Semantically, ideophones that involve reduplication or repetition often depict iterative events, for example repeated motion such as drumming or dripping water or recurring sounds such as noise of mice.

\paragraph{Final lengthening}

A large group of Gyeli ideophones systematically employs final vowel lengthening, as shown in (\ref{IDEOlength}). All of them occur as monosyllabic words only. 


\begin{exe} \ex \label{IDEOlength}
\begin{tabular}{ll}
ndɛ̃́ɛ̃́ɛ̃́ɛ̃́ & `depiction of staring' \\
wɔ́ɔ́ɔ́ɔ́ɔ́ & `depiction of moving by foot or motorbike'  \\
bã̂ã̂ã̂ã̂ã̂ & `depiction of walking a long distance fast' \\
wùùùù & `depiction of pouring liquids or granulars'  \\
pfáááá & `depiction of flinging a long object or slinging' \\
tèèèè & `depiction of waiting' \\
\end{tabular}
\end{exe} 

In comparison to iterative, repetitive ideophones, this group depicts events that either persist in time, for instance staring or waiting, or depict distances, as it is the case with flinging an object (into some distance) or moving into distance.

As mentioned above, this group of ideophones that receives its special marking in the sense of \posscitet{dingemanse2011} definition by vowel lengthening usually does not combine with reduplication. There are a few exceptions, however. For instance, {\itshape wùùùù} `depiction of pouring liquids or granulars' was found to be used in a reduplicated form, depicting the situation when the main character in the Nzambi story (see Appendix II.2) repeatedly pours fuel onto a house. 


\paragraph{Special syllable structure} Some ideophones in Gyeli are further phonologically marked by a closed final syllable structure. As such, ideophones form an exception to a general rule of open syllables in the language (\sectref{sec:Syllable}).
Closed syllables in ideophones frequently end in /m/, but also voiceless obstruents such as /f/ or /k/. Most of them are monosyllabic, as in (\ref{IDEOsyll1}).


\begin{exe} \ex \label{IDEOsyll1}
\begin{tabular}{ll}
wɔ̀m & `depiction of (sudden) silence' \\
ùf & `depiction of sound when something catches fire suddenly'  \\
gbìm & `depiction of putting or falling down of a person or object' \\
bààm & `depiction of closing or finishing something'  \\ 
\end{tabular}
\end{exe}

\noindent There are also bisyllabic ideophones whose second syllable is closed, ending in the nasal /m/, as shown in (\ref{IDEOsyll2}).

\begin{exe} \ex \label{IDEOsyll2}
\begin{tabular}{ll}
pfùtùm & `depiction of sound when jumping into water' \\
pùdùm & `depiction of falling into mud or throwing stone into water'  \\
ntɔ̀ndɔ̀m & `depiction of monkeys jumping in trees' \\
\end{tabular}
\end{exe}


Most of these closed syllable ideophones occur without reduplication. In these cases, they typically depict some sort of suddenness  (sudden silence, suddenly catching fire) or an endpoint of an event (falling, closing, hitting water). There are, however, also a few examples of closed syllable ideophones which involve reduplication such as {\itshape wùf wùf} `depiction of walking mice'.

The other unusual syllable type found in ideophones is that of a consonantal nucleus. Examples are given in (\ref{IDEOsyll3}). The voiceless bilabial in {\itshape p p p p} `depiction of smoking pipe' is produced with an ingressive airstream, imitating the inhaling when smoking.


\begin{exe} \ex \label{IDEOsyll3}
\begin{tabular}{ll}
ḿ m̀ m̀ m̀ ḿ & `depiction of someone mumbling to himself' \\
p p p p & `depiction of smoking pipe'  \\
\end{tabular}
\end{exe}






\subsection{Morphosyntactic properties of ideophones}
\label{sec:IDEOsyn}

In terms of word class, ideophones have been assigned to different parts of speech in the literature, depending on the language. \citet[173]{dwyer2003} provide examples from different African languages where ideophones are categorized, for instance, as verbs, adjectives, interjectionals, special classes, but most commonly as adverbs. They further specify that ideophones
\begin{quote}
``often differ syntactically from the rest of the grammar. 1) usually occur either before or after a sentence; 2) often don't fit into any of the standard categories for parts of speech.''
(p. 174)
\end{quote}
These generalizations also apply in Gyeli. Gyeli ideophones constitute a word class on their own as characterized by their syntactic independence, i.e.\ outside of the syntactic phrase. Possible positions where ideophones are found are i) at the end of an intonation phrase, ii) independently, i.e.\ outside of an intonation phrase, and iii) as complements in complement clauses.

\paragraph{Ideophones at the end of intonation phrases}

Ideophones in Gyeli frequently occur at the end of an intonation phrase as in (\ref{IDEOadv1}) and (\ref{IDEOadv2}). In these cases, ideophones are similar to adverbs in their position and their function, namely depicting the manner in which an action or event happens.


\begin{exe} 
\ex\label{IDEOadv1}
  \glll yɔ́ɔ̀ mùdã̂ dígɛ́ mísì {\bfseries ndẽ́ẽ́ẽ́} \\
       yɔ́ɔ̀ m-ùdã̂ dígɛ-H m-ísì ndẽ́ẽ́ẽ́ \\
        so N1-woman watch-R ma6-eye IDEO:staring \\
    \trans `So the woman looks with her eyes [depiction of staring].'
\end{exe}


\begin{exe} 
\ex\label{IDEOadv2} 
  \glll  bá kɛ́ ndáà nà tɛ́lɛ́ mákùndù má kùrã̂  {\bfseries kɛ́}-{\bfseries kɛ́}-{\bfseries kɛ́}-{\bfseries kɛ́}-{\bfseries kɛ́}\\
         ba-H kɛ-H ndáà nà tɛ́lɛ-H H-ma-kùndù má kùrã̂  kɛ́-kɛ́-kɛ́-kɛ́-kɛ́ \\
        2-PRES go-R also COM put-R OBJ.LINK-ma6-clay.house 6:ATT $\emptyset$7.electricity IDEO:repeated.placement \\
    \trans `They also go and put clay houses with electricity, [depiction of putting the electricity poles along the road].'
\end{exe}

In contrast to adverbs, ideophones also occur in constructions with the deictic element {\itshape mpù} `like this', as shown in (\ref{IDEOadv3}).


\begin{exe} 
\ex\label{IDEOadv3}
  \glll yɔ́ɔ̀ Nzàmbí njí mpù {\bfseries bã̂ã̂ã̂ã̂} njì dígɛ̀ mpù \\
        yɔ́ɔ̀ Nzàmbí nji-H mpù bã̂ã̂ã̂ã̂ njì dígɛ̀ mpù \\
        so $\emptyset$1.PN come-R like.this IDEO:walking.far come look like.this \\
    \trans `So Nzambi comes like this [depiction of walking a long distance], comes looking like this.'
\end{exe}

\noindent The use of deictic elements such as {\itshape mpù} `like this' makes perfect sense in that it signals  the verbal depiction.


\paragraph{Ideophones as {\itshape nâ} complements} Similarly, the same sort of signaling happens when ideophones are used as complements in {\itshape nâ} clauses, as illustrated in (\ref{IDEOna}).

%[??? PUT MPU AND NA CONSTRUCTIONS IN ONE PARAGRAPH AS DEICTIC FRAMING? IS SYNTACTIC ANALYSIS AS COMPLEMENT CORRECT AT ALL?]


\begin{exe} 
\ex\label{IDEOna}
  \glll Nzàmbí, màbɔ́ɔ̀ nkwɛ́ɛ̀ dé nâ {\bfseries vɔ́sì} \\
    Nzàmbí ma-bɔ́ɔ̀ nkwɛ́ɛ̀ dé nâ vɔ́sì \\
          $\emptyset$1.PN ma6-bread.fruit $\emptyset$3.basket LOC COMP IDEO:pouring \\
    \trans `Nzambi pours the bread fruits into the basket.'
\end{exe}

\noindent This type of construction is parallel to reported speech, as discussed in \citet{guldemann2008}. For more information on Gyeli complement constructions and reported speech, see \sectref{sec:CompC}.

\paragraph{Syntactic independence of ideophones}
Gyeli ideophones occur independently from an intonation phrase, rather forming an intonation phrase on their own. In this, they differ from adverbs which cannot occur as independent intonation phrases. In (\ref{IDEOfree1}), the ideophone occurs before the intonation phrase it refers to in the discourse. The ideophone is separated from the following intonation by a short pause.

\begin{exe} 
\ex\label{IDEOfree1} 
  \glll  {\bfseries gbĩ́} {\bfseries gbĩ̀} {\bfseries gbĩ́} {\bfseries gbĩ̀} {\bfseries gbĩ́}   à múà nà bábɛ̀ tí wúmbɛ́ wɛ̀ \\
            gbĩ́-gbĩ̀-gbĩ́-gbĩ̀-gbĩ́  a múà nà bábɛ̀ tí wúmbɛ-H wɛ̀   \\
         IDEO:roaming 1\textsc{sg}.PST1 PROSP COM $\emptyset$7.illness NEG want-R die \\
    \trans `[depiction of disease roaming in his body] He was about to be sick without wanting to die.'
\end{exe}

\noindent Intonationally independent ideophones can also follow the intonation phrase they are semantically linked to in the discourse, as shown in (\ref{IDEOfree2}). 

\begin{exe} 
\ex\label{IDEOfree2}
  \glll wɛ́ dyúwɔ́ mpù bàmìntùlɛ̀ bɔ́gá bá tsígɛ̀ {\bfseries tsùk} {\bfseries tsùk} {\bfseries tsùk} \\
        wɛ-H dyúwɔ-H mpù ba-mìntùlɛ̀ bɔ́-gá ba-H tsígɛ̀ tsùk-tsùk-tsùk \\
        2\textsc{sg}-PRES hear-R like.this ba2-mouse 2-other 2-PRES take.off IDEO:rustling \\
    \trans `You hear like this the other mice take off [depiction of noise made by mice].'
\end{exe}

\noindent In addition to intonational breaks, the end of an intonation phrase can be indicated by the tonal melody. In (\ref{IDEOfree2}), it is the L tone on {\itshape tsígɛ̀} `take' off which shows the end on the intonation phrase. If the ideophone was part of the same intonation phrase, the final tone on {\itshape tsígɛ̀} would be H.











\section{Pronouns}
\label{sec:PRO}

Gyeli has different types of pronouns, i.e.\ grammatical free morphemes that can replace a noun. The different pronoun form paradigms arise from the pronouns' differing syntactic functions and distributions. I distinguish subject pronouns from non-subject pronouns. The latter are used in object and adjunct position. Gyeli has further interrogative pronouns, possessor pronouns, and a reflexive pronoun. 

Generally, agreement class 2 pronouns are also used for impersonal reference. For instance, active clauses with the impersonal {\itshape ba} pronoun are preferred over passive constructions (\sectref{sec:PASS}). This pronoun can also be used in impersonal relative clauses, expressing `who' in the subordinate clause even if the referent of the main clause is expressed by a different agreement/person class (\sectref{sec:Relativeclauses}).




\subsection{Subject pronouns}
\label{sec:SBJPRO}

Subject pronouns are rarely used in Gyeli, with only 17 occurrences in the corpus, since subject noun phrases are mostly expressed by a noun or entirely dropped, leaving only the STAMP marker (\sectref{sec:SCOP}) as portmanteau morpheme that expresses subject agreement on the predicate. Subject pronouns are used for subject focus of, mostly, speech act participants. Non-speech act participants are focused through other information structure strategies (\sectref{sec:IS}).

Table \ref{Tab:SBJpro} provides the subject pronoun forms for both speech and non-speech act participants. All subject pronouns are specified for tone (unlike the STAMP markers which take their tonal marking from the tense-mood category they encode). Most persons have a H tone pronoun, with the exceptions of the first and second person singular and the pronouns of agreement classes 1 and 9.

\begin{table} 
\centering
\begin{tabular}{l|l|l}
 \midrule
 & Singular & Plural \\

Speech Act Participants & 1\textsc{sg} {\itshape mɛ̀} & 1\textsc{pl} {\itshape bí} \\
 & 2\textsc{sg} {\itshape wɛ̀} & 2\textsc{pl} {\itshape bé} \\
 \midrule
Non-Speech Act Participants & cl.1 {\itshape nyɛ̀} & cl.2 {\itshape bá} \\
(3\textsuperscript{rd} person) & cl.3 {\itshape wú} & cl.4  {\itshape mí} \\
& cl.5 {\itshape lí} & cl.6 {\itshape má} \\
 & cl.7 {\itshape yí} & cl.8 {\itshape bé} \\
&  cl.9 {\itshape nyì} & \\
 \midrule
\end{tabular}
\caption{Subject pronouns}
\label{Tab:SBJpro}
\end{table}

While many subject pronouns are segmentally identical to the STAMP markers of their person/class, there are a few exceptions which clearly show that subject pronouns form a distinct paradigm.  These exceptions include the first and second person plural,  and the pronoun of agreement class 1. To indicate this distinction in the glosses, I mark subject pronouns with  `SBJ', while the STAMP marker is only marked for its agreement class/person, as in (\ref{SBJpro1}) where subject pronoun (in bold) and STAMP marker differ in their form.

\begin{exe} 
\ex\label{SBJpro1} 
  \glll  donc {\bfseries bí} yá táálɛ́ bê yàlànɛ̀ àà \\
        donc bí ya-H táálɛ-H bê yàlanɛ àà \\
       so[French] 1\textsc{pl}.SBJ  1\textsc{pl}-PRES begin-R 2\textsc{pl} respond[Bulu] EXCL   \\
    \trans `So we start to respond to you, mhm.'
\end{exe}

Other subject pronouns are form identical to their STAMP marker and might only differ tonally, depending on the tense-mood category, as in (\ref{SBJpro2}).

\begin{exe} 
\ex\label{SBJpro2} 
  \glll ah mbúmbù {\bfseries wɛ̀} wɛ́ tɛ́lɛ́ núndɛ̀ \\
      ah mbúmbù wɛ wɛ-H tɛ́lɛ-H nú-ndɛ̀ \\
        EXCL $\emptyset$1.namesake 2\textsc{sg}.SBJ 2\textsc{sg}-PRES stand-R 1.DEM-ANA \\
    \trans `Ah namesake, is it you who is standing there?'
\end{exe}

\noindent Thus, although the agreement class 2 subject pronoun {\itshape bá} is segmentally identical to its STAMP marker, the two forms differ due to the \textsc{future} marking on the STAMP marker in (\ref{SBJpro3}).

\begin{exe} 
\ex\label{SBJpro3}
  \glll {\bfseries bá} báà bù mpàgó \\
       bá báà bù mpàgó \\
        2.SBJ 2.FUT break $\emptyset$3.road  \\
    \trans `They will build a road.'
\end{exe}

The subject pronoun always occurs in subject position and always precedes the STAMP marker. If the subject is preceded by a fronted object, as for instance an interrogative pronoun in (\ref{SBJpro4}), the object pronoun will precede both the subject pronoun and STAMP marker.

\begin{exe} 
\ex\label{SBJpro4}
  \glll     gyí bí yá tfúgà yá tfúgá nà gyí\\
             gyí bí ya-H tfúga ya-H tfúga-H nà gyí\\
              what 1\textsc{pl}.SBJ 1\textsc{pl}-PRES suffer 1\textsc{pl}-PRES suffer-R COM what \\
    \trans `What do we suffer, we suffer from what?.'
\end{exe}

There are certain words that can enter between the subject pronoun and the STAMP marker. These are, for instance, the contrastive marker -{\itshape gà}  (\sectref{sec:CONTRS}) that attaches to the subject pronoun, as in (\ref{SBJpro5}).

\begin{exe} 
\ex\label{SBJpro5}
  \glll  yɔ́ɔ̀ Nzàmbí {\bfseries nyɛ̀gà} à kɛ̃́ɛ̃̀ dígɛ̀ mísì \\
         yɔ́ɔ̀ Nzàmbí nyɛ-gà a kɛ̃́ɛ̃̀ dígɛ m-ísì \\
           so $\emptyset$1.PN 1.SBJ-CONTR 1.PST1 go.COMPL watch ma6-eye  \\
    \trans `So this Nzambi has gone and watched with his eyes [= was thinking].'
\end{exe}

\noindent Other nominal modifiers, such as {\itshape bɔ́ɔ̀} `other' in (\ref{SBJpro6}) or {\itshape ndáà}  `also' in (\ref{SBJpro7}) occur between the subject pronoun and the STAMP marker.

\begin{exe} 
\ex\label{SBJpro6}
  \glll   {\bfseries bí} {\bfseries bɔ́ɔ̀} yá bígɛ́ mpá'à wá vɛ́ \\
           bí b-ɔ́ɔ̀ ya-H bígɛ-H mpá'à wá vɛ́ \\
           1\textsc{pl}.SBJ 2-other 1\textsc{pl}-PRES develop-R $\emptyset$3.side 3:ATT which  \\
    \trans `How will we others develop?'
\end{exe}

\begin{exe} 
\ex\label{SBJpro7} 
  \glll  ɛ̀sɛ́ béé ndáà bèyá làwɔ́ fàlà \\
       ɛ̀sɛ́ béé ndáà bèya-H làwɔ-H fàlà \\
        is.it[French] 2\textsc{pl}.SBJ also 2\textsc{pl}[Kwasio]-PRES speak-R $\emptyset$1.French  \\
    \trans `Isn't it, you (pl.), you also speak French.'
\end{exe}






\subsection{Non-subject pronouns}
\label{sec:OBJPRO}

%*** fronted object pronouns seem to surface with a L tone, only in-situ with HL

Gyeli has a paradigm of non-subject pronouns which are used for object and oblique noun phrases. They are glossed as `NSBJ'. They are significantly more frequent in the corpus than subject pronouns, counting 99 occurrences.

As shown in Table \ref{Tab:ProOBJ}, all speech act participants as well as the agreement class 1 pronoun are segmentally identical to their subject pronoun counterparts. All the other non-subject pronouns, namely agreement classes 2 through 9, differ structurally in that they have a non-subject pronoun root -{\itshape ɔ̂} that takes an agreement prefix. All non-subject pronouns are specified for a HL tone.

\begin{table} 
\centering
\begin{tabular}{l|l|l}
 \midrule
 & Singular & Plural \\

Speech Act Participants & 1\textsc{sg} {\itshape mɛ̂} & 1\textsc{pl} {\itshape bî} \\
 & 2\textsc{sg} {\itshape wɛ} & 2\textsc{pl} {\itshape bê} \\
 \midrule
Non-Speech Act Participants & cl.1 {\itshape nyɛ̂} & cl.2 {\itshape b-ɔ̂} \\
 & cl.3 {\itshape w-ɔ̂} & cl.4  {\itshape my-ɔ̂} \\
& cl.5 {\itshape l-ɔ̂} & cl.6 {\itshape m-ɔ̂} \\
 & cl.7 {\itshape y-ɔ̂} & cl.8 {\itshape by-ɔ̂} \\
&  cl.9 {\itshape ny-ɔ̂} & \\
 \midrule
\end{tabular}
\caption{Non-subject pronouns}
\label{Tab:ProOBJ}
\end{table}

 Non-subject pronouns that serve as objects occur in all object positions discussed in \sectref{sec:verbalC} and \sectref{sec:IS}.  The basic position is after the verb, as in (\ref{NSBJ1x}) and (\ref{NSBJ3}). 

\begin{exe} 
\ex\label{NSBJ1x}
  \glll  bwáá lã́ {\bfseries bɔ̂}  \\
         bwáa-H lã-H b-ɔ̂ \\
            2\textsc{pl}-PRES tell-R 2-NSBJ     \\
    \trans `You tell them!'
\end{exe}

\begin{exe} 
\ex\label{NSBJ3}
  \glll byɔ̂ bé vɛ́ {\bfseries bíì} màpè'è \\
        byɔ̂ be-H vɛ̀-H bíì ma-pè'è \\
        8.NSBJ 8-PRES give-R 1\textsc{pl}.NSBJ ma6-wisdom  \\
    \trans `They give us wisdom.'
\end{exe}

Non-subject pronouns serving as objects can also be dislocated to the left edge of the clause, as in (\ref{NSBJ1}).  In this marked position (\ref{NSBJ3}) as well in the in-situ focus position in (\ref{NSBJ1}), the pronoun is optionally lengthened for emphasis. 

\begin{exe} 
\ex\label{NSBJ1} 
  \glll  {\bfseries yɔ́ɔ̀} mɛ́ wúmbɛ́ wû \\
        yɔ́ɔ̀ mɛ-H wúmbɛ-H wû \\
         7.NSBJ 1\textsc{sg}-PRES want-R there \\
    \trans `That is what I want there.'
\end{exe}

The first person plural often occurs with the special form {\itshape bíyɛ̀} in the corpus, as in (\ref{NSBJ2}). This seems even more emphatic than the lengthened form {\itshape bíì}. The data is not sufficient, however, to pinpoint the exact distribution and functional difference between the two emphatic forms. The first person plural is the only person category that has such a suppletive emphatic form.

\begin{exe} 
\ex\label{NSBJ2} 
  \glll bvúlɛ̀ bá ntɛ́gɛ́lɛ́ ndáà bíyɛ̀ \\
       bvúlɛ̀ ba-H ntɛ́gɛlɛ-H ndáà bíyɛ̀ \\
         ba2.Bulu 2-PRES bother-R also 1\textsc{pl}.NSBJ   \\
    \trans `The Bulu bother us, too.'
\end{exe}

\noindent Non-subject pronouns also occur in obliques, as in (\ref{NSBJ2x}).

\begin{exe} 
\ex\label{NSBJ2x} 
  \glll á nyùlɛ́nyúlɛ́  kɔ̀fí nà yɔ̂ \\
       a-H nyùlɛ-nyulɛ-H  kɔ̀fí nà y-ɔ̂  \\
         1-PRES drink-HAB-R $\emptyset$7.coffee COM 7-NSBJ   \\
    \trans `He usually drinks coffee with it [= {\itshape ngùɔ́} `sugar']'
\end{exe}

 




\subsection{Interrogative pronouns}
\label{sec:INTERRPRO}

In addition to subject and non-subject pronouns, Gyeli also has two interrogative pronouns: {\itshape nzá} `who' for animate/human referents and {\itshape gyí} `what' for inanimate referents.
{\itshape nzá} `who/whom' and {\itshape gyí} `what' replace a nominal NP which is shown in (\ref{nza}) and (\ref{gyi}), respectively. In (\ref{nza}), the interrogative replaces the subject NP {\itshape m-ùdũ̂} `man' while, in (\ref{gyi}), the interrogative {\itshape gyí} replaces the object NP {\itshape má-jíwɔ́} `water'. In that sense, they behave like personal pronouns. Both interrogatives are used in all noun phrase environments, namely as subjects, objects, and obliques. 

\begin{exe}
\ex\label{nza}
\begin{xlist}
\ex \label{nza1}
  \glll     [mùdũ̂] à nyɛ́ mùdã̂ \\
              m-ùdũ̂ a nyɛ̂-H m-ùdã̂\\
                N1-man 1.PST1 see-R N1-woman \\
    \trans `The/a man saw the/a woman.'
\ex\label{nza2}
 \glll    {\bfseries nzá}  à nyɛ́ mùdã̂  \\
           nzá  a nyɛ̂-H m-ùdã̂ \\
          who 1.PST1 see-R N1-woman \\
    \trans `Who saw the/a woman?'
\end {xlist}
\end {exe}

\begin{exe}
\ex\label{gyi}
\begin{xlist}
\ex \label{gyi1}
  \glll     mùdũ̂ á nyùlɛ́ [májíwɔ́] \\
             m-ùdũ̂ a-H nyùlɛ-H H-ma-jíwɔ́ \\
              N1-man 1-PRES drink-R OBJ.LINK-ma6-water \\
    \trans `The/a man drinks water.'
\ex\label{gyi2}
 \glll     {\bfseries gyí} mùdũ̂ á nyùlɛ̀  \\
           gyí m-ùdũ̂ a-H nyùlɛ \\
             what N1-man 1-PRES drink \\
    \trans `What does the man drink?'
\end {xlist}
\end {exe}

Interrogative pronouns in oblique phrases are shown with the comitative marker {\itshape nà} in (\ref{nanzá}) and (\ref{nagyí}).

\begin{exe}
\ex\label{nanzá}
\begin{xlist}
\ex \label{nanzá1}
  \glll    mùdũ̂ à kɛ́ màkítì [nà Àdà] \\
           m-ùdũ̂ a kɛ̀-H m-àkítì nà Àdà \\
              N1-man 1.PST1 go-R ma6-market COM $\emptyset$1.PN  \\
    \trans `The/a man went to the market with Ada.'
\ex\label{nanzá2}
 \glll      {\bfseries nà} {\bfseries nzá} mùdũ̂ à kɛ́ màkítì  \\
       nà nzá m-ùdũ̂ a kɛ̀-H m-àkítì  \\
           COM who N1-man 1.PST1 go-R ma6-market\\
    \trans `With whom did the man go to the market?'
\end {xlist}
\end {exe}

\begin{exe}
\ex\label{nagyí}
\begin{xlist}
\ex \label{nagyí1}
  \glll    mùdũ̂ à kɛ́ màkítì [nà  tṹũ̀]\\
                 m-ùdũ̂ a kɛ̀-H m-àkítì nà  tṹũ̀ \\
              N1-man 1.PST1 go-R ma6-market COM $\emptyset$7.axe \\
    \trans `The/a man went to the market with an axe.'
\ex\label{nagyí2}
 \glll    {\bfseries nà} {\bfseries gyí} mùdũ̂ à kɛ́ màkítì \\
         nà gyí m-ùdũ̂ a kɛ̀-H m-àkítì  \\
             COM what N1-man 1.PST1 go-R ma6-market \\
    \trans `With what did the man go to the market?'
\end {xlist}
\end {exe}


{\itshape nà nzá} `with whom' is interesting in that {\itshape nzá} seems to take a plural marker if the expected answer is more than one person, as shown in (\ref{banzá}). Since the prefix {\itshape bà}- comes with a L tone, it seems to behave like either a noun class or agreement prefix. Since {\itshape nzá} only occurs with humans, the prefix (if it should be analyzed as such) is invariably class 2 {\itshape bà}-, therefore it is difficult to test whether the supposed prefix belongs to a noun or a modifier.

\begin{exe}
\ex\label{banzá}
\begin{xlist}
\ex \label{banzá1}
  \glll    mùdũ̂ à kɛ́ màkítì [nà Àdà nà Màmbì] \\
           m-ùdũ̂ a kɛ̀-H m-àkítì nà Àdà nà Màmbì \\
              N1-man 1.PST1 go-R ma6-market COM $\emptyset$1.PN COM $\emptyset$1.PN \\
    \trans `The/a man went to the market with Ada and Mambi.'
\ex\label{banzá2}
 \glll      {\bfseries nà} {\bfseries bànzá} mùdũ̂ à kɛ́ màkítì \\
          nà bà-nzá m-ùdũ̂ a kɛ̀-H m-àkítì \\
           COM 2-who ba1-man 1.PST1 go-R ma6-market\\
    \trans `With whom did the man go to the market?'
\end {xlist}
\end {exe}


\subsection{Possessor pronouns}
\label{sec:POSS}

\noindent Possessor pronouns in Gyeli consist of a root indicating the possessor and a prefix that agrees with the possessee, as shown in (\ref{POSS}).

\begin{exe}
\ex\label{POSS}
\begin{xlist}
\ex \label{POSS1}
  \gll     m-ùdì w-ɔ̂ \\
                N1-man 1-POSS.2\textsc{sg} \\
    \trans `your (SG) man'
\ex\label{POSS2}
 \gll     mì-nkwɛ́ my-áwɔ́ \\
                mi4-basket 4-POSS.3\textsc{pl} \\
    \trans `their baskets'
\end {xlist}
\end {exe}

\paragraph{Possessor roots}  Table \ref{Tab:SegPoss} shows the possessor roots. While most possessor roots are used for all agreement classes, there are both segmental and tonal changes depending on the phonological shape of agreement prefixes and the agreement class affiliation respectively.

\begin{table} 
\centering
\begin{tabular}{r|p{3cm}|p{3cm}}
 \midrule
 & Singular & Plural \\
  \midrule
 1 & \itshape{-ã} & \itshape{-isi} (\itshape{-usi}) \\
 2 & \itshape{-ɔ} & \itshape{-inɛ} (\itshape{-unɛ}) \\
 3 & \itshape{-ɛ} & \itshape{-awɔ} \\
  \midrule
\end{tabular}
\caption{Basic possessor roots}
\label{Tab:SegPoss}
\end{table}

 \noindent  Some possessor roots are influenced in their segmental form by the shape of the possessee agreement prefix. The first and second person plural are subject to variation if the possessee belongs to class 1 or 3. Then, the first high front vowel used in all other agreement classes turns into a high back vowel as an assimilation to the agreement prefix {\itshape w-} in class 1 and 3. The contrast between the two root shapes is illustrated in (\ref{w}). 

\begin{exe}
\ex\label{w}
\begin{xlist}
\ex \label{w1}
  \gll     gyà y-{\bfseries í}sí \\
                7.music 7-POSS.1\textsc{pl} \\
    \trans `our music'
\ex\label{w2}
 \gll     m-wánɔ̀ w-{\bfseries ù}sí \\
                N1-child 1-POSS.1\textsc{pl} \\
    \trans `our child'
\end {xlist}
\end {exe}

The agreement class that the possessor root takes also determines the tonal pattern of the root. The tonal pattern of the first and second person singular are the same in every agreement class, as shown in Table \ref{Tab:TonePoss}. The vast majority of agreement classes takes a H tone in the third person singular and a H H pattern for the plural possessor roots. Classes 1 and 9, however, are different: the third person singular has a falling HL tone and the plural persons are L H.

\begin{table} 
\centering
\begin{tabular}{|c|p{2cm}p{3cm}|p{2cm}p{3cm}|}
 \midrule
 & \multicolumn{2}{c}{Basic tonal pattern} & \multicolumn{2}{c|}{Exceptions: cl.~1 and 9} \\
  \midrule
 Person & Singular & Plural & Singular & Plural \\
 1 & -ã̂ & -ísí (-úsí) & -ã̂ & -{\bfseries ìsí} (-{\bfseries ùsí}) \\
 2 & -ɔ̂ & -ínɛ́ (-únɛ́) & -ɔ̂ & -{\bfseries ìnɛ́} (-{\bfseries ùnɛ́}) \\
 3 & -ɛ́ & -áwɔ́ & -{\bfseries ɛ̂} & -{\bfseries àwɔ́} \\
  \midrule
\end{tabular}
\caption{Tonal patterns of possessor pronouns}
\label{Tab:TonePoss}
\end{table}

\paragraph{Possessee agreement prefixes} Agreement of possessor pronouns is marked by agreement prefixes which are listed in Table \ref{Tab:PossPre} for the various agreement classes.

\begin{table} 
\centering
\begin{tabular}{ll}
 \midrule
AGR class & AGR prefix \\
  \midrule
 1 & w-  \\
 2 & b- \\
 3 & w- \\
4 & mi- \\
5 & l- \\
6 & m- \\
7 & y- \\
8 & bi- \\
9 & ny- \\
  \midrule
\end{tabular}
\caption{Possessee agreement prefixes}
\label{Tab:PossPre}
\end{table}

\noindent Prefixes of classes 4 and 8 ending in a high front vowel are assimilated to the possessor root. If the root starts with a high front vowel /i/ as for the first and second person plural ({\itshape -ísí} and {\itshape -ínɛ́}), the vowel of the prefix is deleted:

\begin{exe} \ex\label{POSS41}
\begin{tabular}{llllll}
\multicolumn{6}{l}{\bfseries class 4:} \\
 {\itshape mi-} & + & {\itshape -ísí} & $\rightarrow$ &  {\itshape mísí} & `our' \\
 {\itshape mi-} & + & {\itshape -ínɛ́} & $\rightarrow$ & {\itshape mínɛ} & `your (PL)' \\
\end{tabular}
\end{exe}

\begin{exe} \ex\label{POSS81}
\begin{tabular}{llllll}
\multicolumn{6}{l}{\bfseries class 8:}  \\
 {\itshape bi-} & + & {\itshape -ísí} & $\rightarrow$ &  {\itshape bísí} & `our' \\
 {\itshape bi-} & + & {\itshape -ínɛ́} & $\rightarrow$ & {\itshape bínɛ} & `your (PL)' \\
\end{tabular}
\end{exe}

\noindent For the other roots starting in different vowels, the prefix vowel is assimilated and becomes a glide:

\begin{exe} \ex\label{POSS42}
\begin{tabular}{llllll}
\multicolumn{6}{l}{\bfseries class 4:} \\
 {\itshape mi-} & + & {\itshape -ã̂} & $\rightarrow$ &  {\itshape myã̂} & `my' \\
 {\itshape mi-} & + & {\itshape -ɔ̂} & $\rightarrow$ & {\itshape myɔ̂} & `your (SG)' \\
 {\itshape mi-} & + & {\itshape -ɛ́} & $\rightarrow$ &  {\itshape myɛ́} & `his/her' \\
 {\itshape mi-} & + & {\itshape -áwɔ́} & $\rightarrow$ & {\itshape myáwɔ́} & `their' \\
\end{tabular}
\end{exe}

\begin{exe} \ex\label{POSS82}
\begin{tabular}{llllll}
\multicolumn{6}{l}{\bfseries class 8:}  \\
 {\itshape bi-} & + & {\itshape -ã̂} & $\rightarrow$ &  {\itshape byã̂} & `my' \\
 {\itshape bi-} & + & {\itshape -ɔ̂} & $\rightarrow$ & {\itshape byɔ̂} & `your (SG)' \\
 {\itshape bi-} & + & {\itshape -ɛ́} & $\rightarrow$ &  {\itshape byɛ́} & `his/her' \\
 {\itshape bi-} & + & {\itshape -áwɔ́} & $\rightarrow$ & {\itshape byáwɔ́} & `their' \\
\end{tabular}
\end{exe}

\noindent I assume that possessee agreement prefixes are underlyingly toneless just as noun class prefixes. As discussed in the phonology chapter in \sectref{sec:toneless}, segments may be deleted, but their tones often survive. Possessee agreement prefixes never surface with a vowel because the vowel is deleted in assimilation with the vowel onset of the possessor root. If the deleted vowel was specified for tone, one would assume that the tone survives and affects the root. As possessor roots that are preceded by a deleted vowel (e.g.~cl.~4 or 8) show the same tonal pattern as those that are just preceded by a consonant  (e.g.~cl.~1 or 3), I suggest that the deleted prefix vowel came without any tone in the first place.

%Below, I give examples for the different possessive pronouns modifying nouns of the various classes. Classes 8a and 8b have the same agreement pattern because they are the same agreement class. Since class 8a is a plural class while 8b contains singular nouns, I provide examples of both sub-classes.



%\begin{longtable} {p{2.4cm}p{3.5cm}p{0.0cm}p{2.8cm}p{3.5cm}}
%\multicolumn{5}{l}{\bfseries class 1:} \\
% {\itshape sɔ́ w-ã̂} & `my friend' &  &  {\itshape sɔ́ w-ùsí} & `our friend' \\
% {\itshape sɔ́ w-ɔ̂} & `your (SG) friend' & & {\itshape sɔ́ w-ùnɛ́} & `your (PL) friend' \\
% {\itshape sɔ́ w-ɛ̂} & `his/her friend' &  &  {\itshape sɔ́ w-àwɔ́} & `their friend' \\
% & & & & \\
%\multicolumn{5}{l}{\bfseries class 2:} \\
% {\itshape ba-sɔ́ b-ã̂} & `my friends' &  &  {\itshape ba-sɔ́ b-ísí} & `our friends' \\
% {\itshape ba-sɔ́ b-ɔ̂} & `your (SG) friends' & & {\itshape ba-sɔ́ b-ínɛ́} & `your (PL) friends' \\
% {\itshape ba-sɔ́ b-ɛ́} & `his/her friends' &  &  {\itshape ba-sɔ́ b-áwɔ́} & `their friends' \\
%& & & & \\
%\multicolumn{5}{l}{\bfseries class 3:} \\
% {\itshape mbɛ̂ w-ã̂} & `my door' &  &  {\itshape mbɛ̂ w-úsí} & `our door' \\
% {\itshape mbɛ̂ w-ɔ̂} & `your (SG) door' & & {\itshape mbɛ̂ w-únɛ́} & `your (PL) door' \\
% {\itshape mbɛ̂ w-ɛ́} & `his/her door' &  &  {\itshape mbɛ̂ w-áwɔ́} & `their door' \\
%& & & & \\
%\multicolumn{5}{l}{\bfseries class 4:} \\
% {\itshape mi-mbɛ̂ my-ã̂} & `my doors' &  &  {\itshape mi-mbɛ̂ m-ísí} & `our doors' \\
% {\itshape mi-mbɛ̂ my-ɔ̂} & `your (SG) doors' & & {\itshape mi-mbɛ̂ m-ínɛ́} & `your (PL) doors' \\
% {\itshape mi-mbɛ̂ my-ɛ́} & `his/her doors' &  &  {\itshape mi-mbɛ̂ my-áwɔ́} & `their doors' \\
%& & & & \\
%\multicolumn{5}{l}{\bfseries class 5:} \\
% {\itshape le-kí l-ã̂} & `my egg' &  &  {\itshape le-kí l-ísí} & `our egg' \\
% {\itshape le-kí l-ɔ̂} & `your (SG) egg' & & {\itshape le-kí l-ínɛ́} & `your (PL) egg' \\
% {\itshape le-kí l-ɛ́} & `his/her egg' &  &  {\itshape le-kí l-áwɔ́} & `their egg' \\
%& & & & \\
%\multicolumn{5}{l}{\bfseries class 6:} \\
% {\itshape ma-kí m-ã̂} & `my eggs' &  &  {\itshape ma-kí m-ísí} & `our eggs' \\
% {\itshape ma-kí m-ɔ̂} & `your (SG) eggs' & & {\itshape ma-kí m-ínɛ́} & `your (PL) eggs' \\
% {\itshape ma-kí m-ɛ́} & `his/her eggs' &  &  {\itshape ma-kí m-áwɔ́} & `their eggs' \\
%& & & & \\
%\multicolumn{5}{l}{\bfseries class 7:} \\
% {\itshape síŋgì y-ã̂} & `my cat' &  &  {\itshape síŋgì y-ísí} & `our cat' \\
% {\itshape síŋgì y-ɔ̂} & `your (SG) cat' & & {\itshape síŋgì y-ínɛ́} & `your (PL) cat' \\
% {\itshape síŋgì y-ɛ́} & `his/her cat' &  &  {\itshape síŋgì y-áwɔ́} & `their cat' \\
%& & & & \\
%\multicolumn{5}{l}{\bfseries class 8a:} \\
% {\itshape be-síŋgì by-ã̂} & `my cats' &  &  {\itshape be-síŋgì b-ísí} & `our cats' \\
% {\itshape be-síŋgì by-ɔ̂} & `your (SG) cats' & & {\itshape be-síŋgì b-ínɛ́} & `your (PL) cats' \\
% {\itshape be-síŋgì by-ɛ́} & `his/her cats' &  &  {\itshape be-síŋgì by-áwɔ́} & `their cats' \\
%& & & & \\
%\multicolumn{5}{l}{\bfseries class 8b:} \\
% {\itshape vìyɔ́ by-ã̂} & `my fire' &  &  {\itshape vìyɔ́ b-ísí} & `our fire' \\
% {\itshape vìyɔ́ by-ɔ̂} & `your (SG) fire' & & {\itshape vìyɔ́ b-ínɛ́} & `your (PL) fire' \\
% {\itshape vìyɔ́ by-ɛ́} & `his/her fire' &  &  {\itshape vìyɔ́ by-áwɔ́} & `their fire' \\
%& & & & \\
%\multicolumn{5}{l}{\bfseries class 9:} \\
% {\itshape ngɔ̀mɔ̀ ny-ã̂} & `my drum' &  &  {\itshape ngɔ̀mɔ̀ ny-ìsí} & `our drum' \\
% {\itshape ngɔ̀mɔ̀ ny-ɔ̂} & `your (SG) drum' & & {\itshape ngɔ̀mɔ̀ ny-ìnɛ́} & `your (PL) drum' \\
% {\itshape ngɔ̀mɔ̀ ny-ɛ̂} & `his/her drum' &  &  {\itshape ngɔ̀mɔ̀ ny-àwɔ́} & `their drum' \\
%\end{longtable}



\subsection{Reflexive pronoun {\itshape mɛ́dɛ́}}
\label{sec:REFL}

The reflexive pronoun {\itshape mɛ́dɛ́} `self' can be used in all kinds of noun phrases: subject, object, and oblique. Its general function is to emphasize a referent or to express identity between subject and object referent. {\itshape mɛ́dɛ́} `self' can take a range of positions. In subject noun phrases, it can follow the subject pronoun, as in (\ref{mede1}) and (\ref{mede2}). 

\begin{exe} 
\ex\label{mede1}
  \glll bímbú lɛ́ mámbòngò mâ wɛ̀ {\bfseries mɛ́dɛ́} dígɛ̂ {\bfseries mɛ́dɛ́} \\
         bímbú lɛ́ ma-mbòngò mâ wɛ mɛ́dɛ́ dígɛ̂ mɛ́dɛ́ \\
       $\emptyset$5.amount 5:ATT ma6-plant 6.DEM.PROX 2\textsc{sg}.SBJ self look.IMP self   \\
    \trans `The amount of these plants, yourself, look yourself,'
\end{exe}

\begin{exe} 
\ex\label{mede2}
  \glll   àà ndáwɔ̀ dé tù nyɛ̀ {\bfseries mɛ́dɛ́} támé   \\
           àà ndáwɔ̀ dé tù nyɛ mɛ́dɛ́ támé      \\
          1.COP $\emptyset$9.house LOC inside 1 self alone \\
    \trans `He is in his house all by himself.'
\end{exe}

The pronoun can also occur just after the finite verb form, as in (\ref{mede3}), where it still refers to the subject. Given that other words, such as the finite verb form in this example, can enter between the subject and reflexive pronoun, I analyze {\itshape mɛ́dɛ́} `self' as a free morpheme.

\begin{exe} 
\ex\label{mede3}
  \glll  à múà {\bfseries mɛ́dɛ́} nyá mùdì \\
          a múà mɛ́dɛ́ nyá m-ùdì   \\
         1 be.almost self real N1-person    \\
    \trans `He was himself a real (old) man.'
\end{exe}


{\itshape mɛ́dɛ́} `self' also follows nouns (instead of pronouns), as in (\ref{mede4}) where it follows the left-dislocated object noun.


\begin{exe} 
\ex\label{mede4}
  \glll sá {\bfseries mɛ́dɛ́} mɛ̀ nzí sâ yî \\
        sá mɛ́dɛ́ mɛ nzí sâ yî \\
        $\emptyset$7.thing self 1\textsc{sg} PROG.PST do 7.DEM.PROX \\
    \trans `The thing itself, I was doing this.'
\end{exe}


In object noun phrases, {\itshape mɛ́dɛ́} `self' directly follows the non-subject pronoun, indicating identity between the subject and the object, as in (\ref{mede}) for all animate person categories.\footnote{The other non-speech act participant categories, namely agreement classes 3 through 9, all adhere to the same pattern.}

\begin{exe} 
\ex\label{mede}
\begin{xlist}
\ex\label{medea}
  \glll  mɛ́ nyɛ́ {\bfseries mɛ̂} {\bfseries mɛ́dɛ́} \\
          mɛ-H nyɛ̂-H mɛ̂ mɛ́dɛ́   \\
          1\textsc{sg}-PRES see-R 1\textsc{sg}.NSBJ self   \\
    \trans `I see myself.'
\ex\label{medeb}
  \glll  wɛ́ nyɛ́ {\bfseries wɛ̂} {\bfseries mɛ́dɛ́} \\
          wɛ-H nyɛ̂-H wɛ̂ mɛ́dɛ́   \\
          2\textsc{sg}-PRES see-R 2\textsc{sg}.NSBJ self   \\
    \trans `You (sg.) see yourself.'
\ex\label{medec}
  \glll  á nyɛ́ {\bfseries nyɛ̂} {\bfseries mɛ́dɛ́} \\
          a-H nyɛ̂-H nyɛ̂ mɛ́dɛ́   \\
          1-PRES see-R 1\textsc{sg}.NSBJ self   \\
    \trans `S/he sees her/himself.'
\ex\label{meded}
  \glll  yá nyɛ́ {\bfseries bî} {\bfseries mɛ́dɛ́} \\
          ya-H nyɛ̂-H bî mɛ́dɛ́   \\
          1\textsc{pl}-PRES see-R 1\textsc{pl}.NSBJ self   \\
    \trans `We see ourselves.'
\ex\label{medee}
  \glll  bwá nyɛ́ {\bfseries bê} {\bfseries mɛ́dɛ́} \\
          bwa-H nyɛ̂-H bê mɛ́dɛ́   \\
          2\textsc{pl}-PRES see-R 2\textsc{pl}.NSBJ self   \\
    \trans `You (pl.) see yourselves.'
\ex\label{medef}
  \glll  bá nyɛ́ {\bfseries bɔ̂} {\bfseries mɛ́dɛ́} \\
          ba-H nyɛ̂-H b-ɔ̂ mɛ́dɛ́   \\
          2-PRES see-R 2-NSBJ self   \\
    \trans `They see themselves.'
\end{xlist}
\end{exe}



















\section{Other pro-forms}
\label{sec:proform}

Other pro-forms substitute other elements than nouns on the phrasal, clausal, or sentential level. In this section, I describe interrogative pro-forms, pro-adverbs, the pro-clausal tag question marker {\itshape ngáà}, and pro-sentence forms.



\subsection{Interrogative pro-forms}
\label{sec:INTERR}

I treat interrogative pro-forms separately from interrogative pronouns {\itshape nzá} `who' and {\itshape gyí} `what' (\sectref{sec:INTERRPRO}) which clearly replace a noun. In contrast, interrogative pro-forms can replace a range of word classes or phrases. For instance, {\itshape líní} `when' might stand {\itshape in lieu} of an adverb {\itshape tɛ̂} `now' or a complex oblique noun phrase {\itshape mbvû lã̂} `last year'. 

Interrogative pro-forms differ in their structural complexity. Simple forms only include the interrogative word. Complex forms require the interrogative form to occur in a special construction, either with the locative preposition {\itshape ɛ́} or in a noun + noun attributive construction.




\subsubsection{Simple interrogative pro-forms}
\label{sec:INTERRS}

Simple interrogative pro-forms are used in questions to replace either a noun phrase or a temporal adverb. They occur independently as free morphemes. Gyeli has three pro-forms, as listed in (\ref{simpinterr}),  that occur in simple interrogative constructions. 

\begin{exe}\ex\label{simpinterr}
\begin{xlist}
\ex líní `when'
\ex vɛ́ `where'
\ex ná `how'
\end{xlist}
\end{exe}

The interrogative pro-form {\itshape líní} `when' exclusively occurs in simple constructions, no matter if it occurs at the beginning or the end of the question phrase, as shown in (\ref{lini}).

\begin{exe}
\ex\label{lini}
\begin{xlist}
\ex \label{lini1}
  \glll    mùdũ̂ à kɛ́ màkítì [nàkùgúù] \\
      m-ùdũ̂ a kɛ̀-H ma-kítì [nà-kùgúù] \\
              N1-man 1.PST1 go-R ma6-market SIM-$\emptyset$7.evening \\
   \trans `The/a man went to the market yesterday.'
\ex\label{lini2}
 \glll    {\bfseries líní} mùdũ̂ à kɛ́ màkítì  \\
           líní m-ùdũ̂ a kɛ̀-H ma-kítì\\
             when N1-man 1.PST1 go-R ma6-market \\
    \trans `When did the man go to the market?'
\ex\label{lini3}
 \glll     mùdũ̂ à kɛ́ màkítì {\bfseries líní} \\
            m-ùdũ̂ a kɛ̀-H ma-kítì líní\\
              N1-man 1.PST1 go-R ma6-market when \\
    \trans `When did the man go to the market?'
\end {xlist}
\end {exe}

 The main use of {\itshape líní} `when' is in temporal adverbial clauses (\sectref{sec:ADVfull}) to express simultaneity. In fact, its use as an interrogative pro-form is rare, even if possible, as shown in (\ref{lini}). When a question asks for a temporal adjunct in the answer, speakers prefer to use complex interrogatives which can be translated as `what day' and `what time', as discussed in the next section. 

In contrast to {\itshape líní} `when', the other two interrogative pro-forms {\itshape vɛ́} `where' and {\itshape ná} `how' only appear in simple constructions if they are used in-situ at the end of the phrase, as illustrated in (\ref{wherevE}) and (\ref{howna}).

\begin{exe} 
\ex\label{wherevE}
  \glll ɛ́ ná mwánɔ̀ nùù {\bfseries vɛ́} \\
       ɛ́ ná m-wánɔ̀ nùù vɛ́ \\
       LOC how N1-child 1.COP where  \\
    \trans `What! Where is the child?'
\end{exe}


\begin{exe} 
\ex\label{howna}
  \glll kó mbúmbù nyɛ̀ nzí lèmbò dyùù bɔ̂ fàmíì bá bùdì {\bfseries ná} \\
       kó mbúmbù nyɛ nzí lèmbo dyùù b-ɔ̂ fàmíì bá b-ùdì ná \\
       EXCL $\emptyset$1.namesake 1.SBJ PROG.PST know kill 2-NSBJ $\emptyset$1.family 2:ATT ba2-person how \\
    \trans `Oh namesake, how could he kill them, the family of people?'
\end{exe}

\noindent If they are used phrase-initially, however, they obligatorily occur in a complex construction with the preposition {\itshape ɛ́}, as discussed in the following.


\subsubsection{Complex interrogative pro-forms}
\label{sec:INTERRC}

Complex interrogative words can be complex in different ways. They can be formed with i) the locative preposition {\itshape ɛ́} (\sectref{sec:PREP}) or ii) a noun + noun attributive construction (\sectref{sec:CONC}). 


\paragraph{Interrogatives with the preposition {\itshape ɛ́}} 
Gyeli has two interrogative pro-forms that are constructed with the locative preposition {\itshape ɛ́} preceding the interrogative form: {\itshape ɛ́ ná} `how'  and {\itshape ɛ́ vɛ́} `where'. Examples of both  interrogatives that require a temporal and a manner adjunct in the answer are given in (\ref{eve}) and (\ref{ena}) , respectively.

\begin{exe}
\ex\label{eve}
\begin{xlist}
\ex \label{eve1}
  \glll mùdũ̂ à kɛ́ [màkítì] \\
    m-ùdũ̂ a kɛ̀-H ma-kítì \\
              N1-man 1.PST1 go-R ma6-market  \\
    \trans `The/a man went to the market.'
\ex\label{eve2}
 \glll    {\bfseries ɛ́} {\bfseries vɛ́} m-ùdũ̂ à kɛ́  \\
            ɛ́ vɛ́ m-ùdũ̂ a kɛ̀-H \\
             LOC where N1-man 1.PST1 go-PST \\
   \trans `Where did the man go?'
\ex\label{eve3}
 \glll    *{\bfseries vɛ́} m-ùdũ̂ à kɛ́  \\
            vɛ́ m-ùdũ̂ a kɛ̀-H \\
             where N1-man 1.PST1 go-PST \\
   \trans `Where did the man go?'
\end {xlist}
\end {exe}

\begin{exe}
\ex\label{ena}
\begin{xlist}
\ex \label{ena1}
  \glll mùdũ̂ à kɛ́ màkítì [nà màtúà] \\
    m-ùdũ̂ a kɛ̀-H ma-kítì nà màtúà \\
             N1-man 1.PST1 go-R ma6-market COM $\emptyset$1.car\\
   \trans `The/a man went to the market by car.'
\ex\label{ena2}
 \glll {\bfseries ɛ́} {\bfseries ná} mùdũ̂ à kɛ́ màkítì  \\
        ɛ́ ná m-ùdũ̂ a kɛ̀-H ma-kítì \\
             LOC how N1-man 1.PRES go-R ma6-market \\
    \trans `How did the man go to the market?'
\ex\label{ena3}
 \glll *{\bfseries ná} mùdũ̂ à kɛ́ màkítì  \\
        ná m-ùdũ̂ a kɛ̀-H ma-kítì \\
            how N1-man 1.PRES go-R ma6-market \\
    \trans `How did the man go to the market?'
\end {xlist}
\end {exe}

\noindent The complex form {\itshape ɛ́ ná} `how' is also used as a greeting in (\ref{enahow})

\begin{exe} 
\ex\label{enahow}
  \glll mbúmbù {\bfseries ɛ́} {\bfseries ná} \\
        mbúmbù ɛ́ ná \\
        $\emptyset$1.namesake LOC how  \\
    \trans `Namesake, how is it?'
\end{exe}





\paragraph{Interrogatives in noun + noun attributive constructions} The second option for complex interrogatives are interrogative pro-forms such as {\itshape vɛ́} `which' and {\itshape níyɛ̀} `how many' which occur as the second constituent in a attributive construction with a noun and an attributive marker, as in (\ref{intercomp1}) and explained in detail in \sectref{sec:NIntPro}.  


\begin{exe}
\ex\label{intercomp1}
\begin{xlist}
\ex\label{intercomp1a}
 \glll    lèfû lé vɛ́   \\
            le-fû lé vɛ́   \\
             le5-day 5:ATT which \\
   \trans `Which day?'
\ex\label{intercomp1b}
 \glll    màfû má níyɛ̀  \\
            ma-fû má níyɛ̀  \\
             ma6-day 6:ATT how.may  \\
   \trans `How many days?'
\end{xlist}
\end{exe}


\noindent Besides asking for nominal entities or their quantities in the answer, these interrogatives systematically combine with temporal nouns such as {\itshape dúwɔ̀} `day' or {\itshape wùlà} `time, hour' in order to form temporal interrogative constructions.










\subsection{Pro-adverbs {\itshape mpù} and {\itshape ndɛ̀náà}}
\label{sec:ProADV}

The pro-adverbs {\itshape mpù} and {\itshape ndɛ̀náà} generally refer to the manner of an event and are translated with `like this'. The semantic difference between the two pro-forms is not clear. They seem to have a very similar distribution in the corpus and speakers state that they can be used interchangeably. {\itshape mpù} is, however, significantly more frequent in the corpus with 24 occurrences in comparison to six occurrences of {\itshape ndɛ̀náà}.

Both pro-adverbs signal a non-verbal gesture or part of the communication that is happening simultaneously to speech time. In (\ref{likethis1}), the speaker is communicating the number of his children by showing two fingers; {\itshape mpù} is signaling this non-verbal gesture.

\begin{exe} 
\ex\label{likethis1} 
  \glll  bwánɔ̀ {\bfseries mpù} [gesture showing 2]. \\
          b-wánɔ̀ mpù \\
         ba2-child like.this   \\
    \trans `that many children [gesture showing 2].'
\end{exe}

\noindent Similarly, in (\ref{likethis2}), {\itshape ndɛ̀náà} indicates that the greeting is ongoing between the speech act participants.

\begin{exe} 
\ex\label{likethis2}
  \glll     mɛ́ sùmɛ́lɛ́ bê {\bfseries ndɛ̀náà} \\
            mɛ-H sùmɛlɛ-H bê ndɛ̀náà \\
              1\textsc{sg}-PRES greet-R 2\textsc{pl}.NSBJ like.that   \\
    \trans `I greet you like this.'
\end{exe}

\noindent {\itshape mpù} often introduces the use of ideophones, as in (\ref{likethis3}).

\begin{exe} 
\ex\label{likethis3}
  \glll yɔ́ɔ̀ Nzàmbí njí {\bfseries mpù} bã̂ã̂ã̂ã̂ njì dígɛ̀ mpù \\
        yɔ́ɔ̀ Nzàmbí njî-H mpù bã̂ã̂ã̂ã̂ njì dígɛ mpù \\
        so $\emptyset$1.PN come-R like.this IDEO:walking far come look like.this \\
    \trans `So Nzambi comes like this [depiction of walking a long distance], comes looking like this.'
\end{exe}

\noindent The deictic reference of pro-adverbs can also be anaphoric rather than signaling an ongoing or immediately following non-verbal communicative event. This is the case in (\ref{likethis4}), for instance, where {\itshape ndɛ̀náà} summarizes the situation that the speaker has elaborated previously.


\begin{exe} 
\ex\label{likethis4}
  \glll  bon pílì yí báàlá nà bɛ̀ {\bfseries ndɛ̀náà} ndɛ̀náà ndáà ná \\
        bon pílì yi-H báàla-H nà bɛ̀ ndɛ̀náà ndɛ̀náà ndáà ná \\
          good[French] when 7-PRES repeat-R COM be like.that like.that also still  \\
    \trans `So, when it continues and is still like this and like that.'
\end{exe}

As (\ref{likethis4}) and (\ref{likethis5}) show, {\itshape mpù} and {\itshape ndɛ̀náà} `like this' can both occur directly after the finite verb, as expected for an adverb. While {\itshape mpù} often is then followed by an object, this is not the case for {\itshape ndɛ̀náà} in the corpus. Speaker state, however, that it would be perfectly grammatical.

\begin{exe} 
\ex\label{likethis5}
  \glll wɛ́ dyúwɔ́ {\bfseries mpù} bàmìntùlɛ̀ bɔ́gá bá tsígɛ̀ tsùk tsùk tsùk \\
         wɛ-H dyúwɔ-H mpù ba-mìntùlɛ̀ bɔ́-gá ba-H tsígɛ tsùk-tsùk-tsùk \\
        2\textsc{sg}-PRES hear-R like.this ba2-mouse 2-other 2-PRES take.off IDEO:rustling \\
    \trans `You hear like this the other mice take off [depiction of noise of mice].'
\end{exe}

{\itshape mpù}, unlike {\itshape ndɛ̀náà}, is often preceded by the preposition {\itshape ɛ́}, as in (\ref{likethis6}).

\begin{exe} 
\ex\label{likethis6}
  \glll yɔ́ɔ̀ Nzàmbí dígɛ́ mísì {\bfseries ɛ́} {\bfseries mpù} \\
      yɔ́ɔ̀ Nzàmbí dígɛ-H m-ísì ɛ́ mpù \\
       so $\emptyset$1.PN look-R ma6-eye LOC like.this  \\
    \trans `So Nzambi looks with the eyes like this.'
\end{exe}

\noindent The specific function of {\itshape ɛ́} in combination with {\itshape mpù} nor its distribution are clear, however. 




\subsection{Pro-clausal {\itshape ngáà}}
\label{sec:ProClause}

The pro-clausal tag question particle {\itshape ngáà} is used to verify the truth value of a clause in leading polar questions (\sectref{sec:Questions}), as in (\ref{Qtag1}).

\begin{exe} 
\ex\label{Qtag1}
  \glll {\bfseries ngáà} wɛ́ nyɛ́ mpù \\
       ngáà wɛ-H nyɛ̂-H mpù \\
       Q(tag) 2\textsc{sg}-PRES see-R like.this  \\
    \trans `Right, you see that?'
\end{exe}

\noindent {\itshape ngáà} appears both at the beginning of the question, as in (\ref{Qtag1}), or at the end of it, as in (\ref{Qtag2}).


\begin{exe} 
\ex\label{Qtag2}
  \glll wɛ́ nyɛ́ mpù {\bfseries ngáà} \\
       wɛ-H nyɛ̂-H mpù ngáà \\
        2\textsc{sg}-PRES see-R like.this Q(tag) \\
    \trans `You see that, don't you?'
\end{exe}

The pro-clausal particle is used in both affirmative and negated questions. An example of the latter is given in (\ref{Qtag3}).

\begin{exe} 
\ex\label{Qtag3}
  \glll wɛ̀ɛ́ nyɛ́lɛ́ mpù {\bfseries ngáà} \\
       wɛ̀ɛ́ nyɛ̂-lɛ mpù ngáà \\
        2\textsc{sg}.NEG.PRES see-NEG like.this Q(tag) \\
    \trans `You don't see that, do you?'
\end{exe}


{\itshape ngáà} is also used independently on its own as a response to a statement, expressing surprise or verifying the truth value of the statement, comparable to English  `really?' or `is that true?'.

In affirmative questions, Gyeli also uses French loan words or code-switching.\footnote{The status of these French words in Gyeli is not clear at the moment.} In (\ref{Qtag4}), {\itshape ɛ̀sɛ́} taken from French {\itshape est-ce} `is it' is used as tag question marker. There seems to be a preference to use it phrase-initially.

\begin{exe} 
\ex\label{Qtag4} 
  \glll  {\bfseries ɛ̀sɛ́} béé ndáà bèyá làwɔ́ fàlà \\
       ɛ̀sɛ́ béé ndáà bèya-H làwɔ-H fàlà \\
        is.it[French] 2\textsc{pl}.SBJ also 2\textsc{pl}[Kwasio]-PRES speak-R $\emptyset$1.French  \\
    \trans `Isn't it, you (pl.) also, you speak French.'
\end{exe}

In contrast, {\itshape nɔ́ɔ̀} from French {\itshape non} `no' is used phrase-finally with the same function, as in (\ref{Qtag5}).

\begin{exe} 
\ex\label{Qtag5} 
  \glll  béé ndáà bèyá làwɔ́ fàlà {\bfseries nɔ́ɔ̀} \\
       ɛ̀sɛ́ béé ndáà bèya-H làwɔ-H fàlà nɔ́ɔ̀ \\
        is.it[French] 2\textsc{pl}.SBJ also 2\textsc{pl}[Kwasio]-PRES speak-R $\emptyset$1.French no[French] \\
    \trans `You (pl.)also, you speak French, isn't it?'
\end{exe}




\subsection{Pro-sentence forms}
\label{sec:ProSent}

Pro-sentence forms replace an entire sentence. They are typically answers to to polar questions (\sectref{sec:Questions}), making a statement about its truth value. They can, however, also occur as response to a statement that the speaker agrees or disagrees with. Gyeli has several pro-sentence forms for each agreement and disagreement signal. (\ref{yes}) provides a list of pro-forms that signal agreement.

\begin{exe} 
\ex\label{yes}
\begin{xlist}
\ex\label{yes1} ɛ́ɛ̀ `yes'
\ex\label{yes2} áà `yes' 
\ex\label{yes3} èè `yes' 
\ex\label{yes4} ɛ́ɛ́ `yes' 
\ex\label{yes5} m̀ḿḿ `yes'
\ex\label{yes6} ɛ̀hɛ́ɛ́  `yes'
\end{xlist}
\end{exe}

When asked for the translation of `yes', speakers would answer with (\ref{yes1}). In natural speech as in the corpus, however, a range of other agreement signaling pro-forms are used. They all have in common that they only consist of a long vowel or nasal. The tonal melody and vowel length is crucial in distinguishing agreement from disagreement, as the segmentally similar but tonally different pairs in, for instance, (\ref{yes4}) and (\ref{no2}) show. Agreement signals have long segments with either a  falling  or L or H tone, as in  (\ref{yes1}) through (\ref{yes4}). (\ref{yes5}) and (\ref{yes6}), which are tonally identical, are used for emphatic agreement, as in English `exactly!'.
Also {\itshape yà}, or its emphatic form {\itshape yáà}, has been observed in the corpus. These forms are likely loan words from German.\footnote{Some German loan words from colonial times (until 1918) are still widespread in the area, for instance also in Mabi. These include, for instance, {\itshape dunkel} `dark' and {\itshape Dummkopf} `idiot', although Cameroonians are not always sure about their meaning.}

There are fewer pro-forms for disagreement than for agreement. The default form is {\itshape tɔ̀sâ} in (\ref{no1}), which is derived from the negative polarity item {\itshape tɔ̀} (\sectref{sec:InvQUANT1}) and the noun {\itshape sâ} `thing'.

\begin{exe} 
\ex\label{no}
\begin{xlist}
\ex\label{no1} tɔ̀sâ `no'
\ex\label{no2} ɛ́'ɛ̂ `no'
\ex\label{no3} ḿ'm̂ `no'
\end{xlist}
\end{exe}

The other two forms in (\ref{no2}) and (\ref{no3}) are identical in their tonal pattern. They also differ from agreement forms in their relative brevity. Disagreement forms are never lengthened, but rather short. In (\ref{no2}) and (\ref{no3}), the medial glottal stop reinforces the impression of short segments.
















\section{Elements of the nominal phrase}
\label{sec:NAdjuncts}

In this section, I describe all the elements that occur in a noun phrase, apart from the noun, which has been discussed in \sectref{sec:N}. As a basic classification criterion, I distinguish nominal modifiers that agree with the head noun and those that do not, i.e.\ which are invariable.  
	
	%(tests: 1. multiple stacking of same category not allowed so Dem and Poss not same category
		  % 2. modification with intensifier bvubvu for adjectives, not determiners)
	

Agreeing elements in the Gyeli noun phrase differ in the form of agreement encoding. For some parts of speech, agreement is achieved through a prefix. This is the case for all elements discussed in \sectref{sec:MODAgrPre} and \sectref{sec:MODAgrPL}. Other elements, such  as demonstratives in \sectref{sec:DEM} and attributive markers in \sectref{sec:ATT}, show agreement through an unbound agreeing morpheme that differs across different agreement classes.	

Invariable modifiers, i.e.\ elements that do not agree with the head noun, differ in their position relative to the noun. Some invariable modifiers precede the head noun (\sectref{sec:InvQUANT1}), some occur post-nominally (\sectref{sec:InvQUANT2}). The structure of the noun phrase and its various types are presented in \chapref{sec:NP} as well as the gender and agreement system.






\subsection{Modifiers with agreement prefix}
\label{sec:MODAgrPre}


Gyeli has five patterns of agreement prefixes, as shown in Table \ref{Tab:DEICMOD}. Agreement prefixes attach to a variety of agreement targets, including numerals and other quantifiers.\footnote{These nominal modifiers could be argued to constitute adjectives on the basis of their agreement prefixes. Adjectives are, however, usually taken to be `lexical' (or content) words, according to \citet[121]{rijkhoff2002}, and describe properties such as ``size, weight, color, age, and value.'' In Gyeli, they do not take agreement prefixes, as described in \sectref{sec:QUAL}.} All modifiers in Table \ref{Tab:DEICMOD} follow the head noun.

\begin{table}[!h]
\centering
\begin{tabular}{l|p{2cm}p{2cm}p{2cm}p{2cm}l}
 \midrule
AGR & 	-vúdũ̂ 	& 	-fúsì  &  	-ɛ́sɛ̀		& -ɔ́(nɛ́)gá	  & numerals  \\
class & 	`one' 	& 	`different'  &  `all'	& `other'	  & `2' through `5'  \\
 \midrule
1 & 	m- 		& 	m-	& 	w-		&      n- 	&  	 \\
2 & 	bà- 		& 	bà-	& 	b-		&      b- 	& bá-	 \\
3 & 	m- 		& 	$\emptyset$- 	& w-		& 	w-  	& 	 \\
4 & 	mì- 		& 	mì-	& 	my-		& 	my- 	& mí- 	\\
5 & 	lè- 		& 	lè-	& 	l-		&      l- 	&  	\\
6 & 	mà-  		& 	mà-	& 	m-		&      m- 	 & má-	\\
7 & 	$\emptyset$- 	& $\emptyset$- & y- &     y- 	 &  	\\
8 & 	bì- 		& 	bì-	& 	by-		&     by- 	& bí- 	 \\
9 & 	m- 		& $\emptyset$-& ny-	&     ny- 	 &  	\\
 \midrule
\end{tabular}
\caption{Agreement prefixes of nominal modifiers}
\label{Tab:DEICMOD}
\end{table}

While agreement prefixes of specific agreement classes are often similar in their shape, there are differences that define distinct agreement patterns. The agreement patterns for -{\itshape vúdũ̂} `one' and -{\itshape fúsì} `different' are only distinguished in agreement classes 3 and 9. The agreement patterns for -{\itshape ɛ́sɛ̀} `all' and -{\itshape ɔ́(nɛ́)gá} `other' differ in their agreement class 1 prefix. For semantic reasons, agreement prefixes for plural numerals only ever allow plural agreement prefixes. They are different from other agreement patterns in that they are the only ones to take a H tone prefix; agreement prefixes of all other patterns that have a tone bearing unit always have a L tone.

Some differences can be explained on a phonological basis, namely vowel deletion or assimilation in the prefix if the following stem starts with a vowel. This is, for instance the case with class 2 {\itshape bà}- before consonants in comparison to class 2 {\itshape b}- before vowels. Differences in prefix shape that are conditioned by phonological rules are not taken as evidence for different agreement patterns. In the following, I present each prefix agreement pattern and the lexical stems that take it.



\subsubsection{{\itshape -vúdũ̂} `one, same'} 
\label{sec:one}

{\itshape -vúdũ̂} can denote both the cardinal numeral `1' and the deictic modifier meaning `same'. It is distinct from the agreement pattern of the other agreeing numerals `2' through `5' in the L tone on CV- prefixes.

As the cardinal numeral `1', {\itshape -vúdũ̂} logically only occurs with singular entities it modifies. If it is used in order to express identity of entities, however, {\itshape -vúdũ̂} also takes an agreement prefix for plural classes, as shown in Table \ref{Tab:same}. 

\begin{table} 
\centering
\begin{tabular}{llll}
 \midrule
cl. 1 & mùdì & {\bfseries m}-vúdũ̂  & `one/same person' \\
cl. 2 & bùdì & {\bfseries bà}-vúdũ̂ & `same people' \\
cl. 3 & nkɛ̌ & {\bfseries m}-vúdũ̂́ & `one/same basket' \\
cl. 4 & mi-nkwɛ̌ & {\bfseries mì}-vúdũ̂ & `same baskets' \\
cl. 5 & le-dùndà & {\bfseries lè}-vúdũ̂ & `one/same sparrow' \\
cl. 6 & ma-dùndà & {\bfseries mà}-vúdũ̂ &  `same sparrows' \\
cl. 7 & síngì & {\bfseries $\emptyset$}-vúdũ̂ & `one/same cat' \\
cl. 8 & be-síngì & {\bfseries bè}-vúdũ̂ & `same cats' \\
cl. 9 & ndáwɔ̀ & {\bfseries m}-vúdũ̂ & `one/same house' \\
 \midrule
\end{tabular}
\caption{AGR-{\itshape vúdũ̂} `one/same' in various agreement classes}
\label{Tab:same}
\end{table}


\subsubsection{{\itshape -fúsì} `different'} 
-{\itshape fúsì} `different' follows the noun it modifies just as the other modifiers that show agreement through a prefix.  Examples for {\itshape fúsì} `different' in different agreement classes are provided in Table \ref{Tab:different}.

\begin{table} 
\centering
\begin{tabular}{llll}
 \midrule
cl. 1 & mùdì & {\bfseries m}-fúsì  & `a different person' \\
cl. 2 & bùdì & {\bfseries bà}-fúsì & `different people' \\
cl. 3 & nkɛ̌ & {\bfseries $\emptyset$}-fúsì & `a different basket' \\
cl. 4 & mi-nkwɛ̌ & {\bfseries mì}-fúsì & `different baskets' \\
cl. 5 & le-dùndà & {\bfseries lè}-fúsì & `a different sparrow' \\
cl. 6 & ma-dùndà & {\bfseries mà}-fúsì &  `different sparrows' \\
cl. 7 & síngì & {\bfseries $\emptyset$}-fúsì & `a different cat' \\
cl. 8 & be-síngì & {\bfseries bè}-fúsì & `different cats' \\
cl. 9 & ndáwɔ̀ & {\bfseries $\emptyset$}-fúsì & `a different house' \\
 \midrule
\end{tabular}
\caption{AGR-{\itshape fúsì} `different' in various agreement classes}
\label{Tab:different}
\end{table}







\subsubsection{{\itshape -ɛ́sɛ̀} `all'} 
\label{sec:ModAll}

%Non-numeral quantifiers can semantically be distinguished as intersective, universal, or proportionality quantifiers in the nominal domain (D- quantifiers). 
% Finally, {\itshape proportionality} quantifiers such as `most', `half of', or `many of' relate a given quantity to a set of entities. \citet[398]{zerbian2008} propose that Bantu languages generally use complex morphosyntactic constructions to express these. 

%Investigating Gyeli quantifiers show, however, that just like in many other languages these semantic distinctions do not map onto distinct construction types. Rather, Gyeli distinguishes four types of quantifiers which do not correspond with the semantic groupings presented above. Gyeli quantifiers can either be expressed as a nominal genitive construction parallel to the English expression `a multitude of x'. This construction type is the most frequent one and discussed in section \ref{sec:NomQUANT}. Then,  Also, there are invariable quantifiers (see section \ref{sec:InvQUANT}) that either precede or follow the noun, but they do not agree with it. Further, the `all' quantifier {\itshape -ɛ́sɛ̀} is a modifier and agrees with its head noun as shown in this section. Finally, reduplication is a means of expressing `each' in a distributive sense as decribed in section \ref{sec:RedQUANT}. 

{\itshape ɛ́sɛ̀}  `all' is a universal quantifier that  agrees with the head noun through an agreement prefix.  Universal quantifiers express totality and contain items such as `all' and `every' \citep[394]{zerbian2008}. Table \ref{Tab:All} provides examples of the quantifier for all agreement classes showing the agreement prefix in bold. The agreement prefix for `all' is the same as the possessee agreement of possessor roots. As most other modifiers, `all' follows the head noun.

\begin{table} 
\centering
\begin{tabular}{llll}
 \midrule
cl. 1 & mùdì & {\bfseries w}-ɛ́sɛ̀ & `all (the parts of) the person' \\
cl. 2 & bùdì & {\bfseries b}-ɛ́sɛ̀ & `all people' \\
cl. 3 & nkwɛ̌ & {\bfseries w}-ɛ́sɛ̀ & `all (the parts of) the basket' \\
cl. 4 & mi-nkwɛ̌ & {\bfseries my}-ɛ́sɛ̀ & `all baskets' \\
cl. 5 & le-dùndá & {\bfseries l}-ɛ́sɛ̀ & `all (the parts of) the sparrow' \\
cl. 6 & ma-dùndà & {\bfseries m}-ɛ́sɛ̀ & `all sparrows' \\
cl. 7 & síngì & {\bfseries y}-ɛ́sɛ̀ & `all (the parts of) the cat' \\
cl. 8 & be-síngì & {\bfseries by}-ɛ́sɛ̀ & `all cats' \\
cl. 9 & ndáwɔ̀ & {\bfseries ny}-ɛ́sɛ̀ & `all the house' \\
 \midrule
\end{tabular}
\caption{AGR-{\itshape ɛ́sɛ̀} `all' in various agreement classes}
\label{Tab:All}
\end{table}

In Gyeli, {\itshape ɛ́sɛ̀} `all' is typically used with plural nouns.  Also singular forms can, however, be modified by -{\itshape ɛ́sɛ̀} `all'  in a specific context, which is also shown in Table \ref{Tab:All}. This special context requires a situation where a typical singular entity consists of or is cut up into several parts. Taking the example of a cat, {\itshape síngì yɛ́sɛ̀} `all the cat' would mean that a cat is cut up into different parts, but then all the parts are used, which is different from meaning `the whole cat' (\sectref{sec:mandjimo}), as shown in (\ref{allcat}).

\begin{exe}
\ex\label{allcat}
\begin{xlist}
\ex\label{allcat1}
 \gll  síngì y-ɛ́sɛ̀  \\
          $\emptyset$7.cat 7-all  \\
    \trans `all (the parts of) the cat'
\ex \label{allcat2}
  \gll    síngì mànjìmɔ̀ \\
              $\emptyset$7.cat whole \\
    \trans `the whole cat (in its entirety)'
\end {xlist}
\end {exe}
















\subsubsection{{\itshape -ɔ́(nɛ́)gá} `(an)other'}
\label{sec:other}

The full form `other' in careful speech is {\itshape -ɔ́nɛ́gá}. 
In fast speech, however, a shortened form AGR-{\itshape ɔ́gá} is used where {\itshape nɛ́} is omitted. The option to omit {\itshape nɛ́} is indicated by the brackets in Table \ref{Tab:Other}. 

\begin{table} 
\centering
\begin{tabular}{llll}
 \midrule
cl. 1 & mùdì & {\bfseries n}-ɔ́(nɛ́)gá & `another person' \\
cl. 2 & bùdì & {\bfseries b}-ɔ́(nɛ́)gá & `other people' \\
cl. 3 & nkɛ̌ & {\bfseries w}-ɔ́(nɛ́)gá & `another basket' \\
cl. 4 & mi-nkwɛ̌ & {\bfseries my}-ɔ́(nɛ́)gá & `other baskets' \\
cl. 5 & le-dùndà & {\bfseries l}-ɔ́(nɛ́)gá & `another sparrow' \\
cl. 6 & ma-dùndà & {\bfseries m}-ɔ́(nɛ́)gá & `other sparrows' \\
cl. 7 & síngì & {\bfseries y}-ɔ́(nɛ́)gá & `another cat' \\
cl. 8 & be-síngì & {\bfseries by}-ɔ́(nɛ́)gá & `other cats' \\
cl. 9 & ndáwɔ̀ & {\bfseries ny}-ɔ́(nɛ́)gá & `another house' \\
 \midrule
\end{tabular}
\caption{AGR-{\itshape ɔ́ (nɛ́) gá} `other' in various agreement classes}
\label{Tab:Other}
\end{table}






\subsubsection{Agreeing plural numerals} 
\label{sec:ModNUM}

Numerals may, depending on the language, form various numeral series such as enumeratives, cardinal, ordinal, or distributive numerals. In Gyeli, only a few cardinal numerals agree with the noun.
Cardinal numerals are used attributively with nouns when counting items.\footnote{Gyeli numerals do not belong to one uniform category. There are monomorphemic (simple) and polymorphemic (complex) numerals.  Even simple numerals do not belong to one category in terms of parts of speech, but can be classified into three types: i) modifiers, ii) invariable modifiers, and iii) nouns. Complex numerals constitute either a coordination construction or a noun + modifier NP or a combination of the two.}

\paragraph{Numerals `2', `3', `4', and `5'} 
The (cardinal) numerals -{\itshape báà} `2', -{\itshape láálɛ̀} `3', -{\itshape nã̂} `4', and -{\itshape tánɛ̀} `5' agree with their head noun. As enumeratives, i.e.\ in general counting without refering to a specific entity, the class 8 prefix {\itshape bí}- is used. The agreement prefixes of agreeing numerals and some examples are listed in Table \ref{tab:NumPre}.\footnote{Since all the numerals that take agreement markers are inherently plural, singular class prefixes are never used.} 

\begin{table} 
\centering
\begin{tabular}{ll|ll}
 \midrule
AGR class & AGR prefix & Example & Gloss \\
  \midrule
2 & {\bfseries bá-} & {\itshape b-ùdì bá-báà} & `two people' \\
4 & {\bfseries mí-} & {\itshape mi-nkwɛ̂ mí-báà} & `two baskets' \\
6 & {\bfseries má-} & {\itshape ma-kí má-báà} & `two eggs' \\
8 & {\bfseries bé-} & {\itshape be-síngì bé-báà} & `two cats' \\
  \midrule
\end{tabular}
\caption{Agreement prefixes of modifying numerals}
\label{tab:NumPre}
\end{table}

\noindent All agreement prefixes on the agreeing numerals come with a H tone, in contrast to noun class prefixes and agreement prefixes of other modifiers (see \sectref{sec:MODAgrPre}).

One could argue that these agreement prefixes should not be analyzed as such, but may rather constitute attributive markers (\sectref{sec:ATT}) which have the same shape and tone as these prefixes. This is unlikely, however, because enumeratives always require a default prefix even though they are not modifying any noun. It is thus more likely to assume that numerals take a default prefix rather than a default attributive marker in a headless construction. Further, also the genitive marker takes H tone prefixes (\sectref{sec:GEN}).

The cardinal numerals from `2' through `5'  invariably follow the head noun, as shown in (\ref{ModNum1}).

\begin {exe}
\ex\label {ModNum1}
\begin{xlist}
\parbox[t]{2in}{\ex \gll b-ùdã̂ {\bfseries bá}-báà\\ 
ba2-woman 2-two \\
 `two women'}
\parbox[t]{2in}{ \ex \gll 
b-ùdã̂ {\bfseries bá}-láálɛ̀ \\
   ba2-woman 2-three \\
 `three women'}

%\vspace{1pc}

\parbox[t]{2in}{ \ex \gll 
b-ùdã̂ {\bfseries bá}-nã̂\\ 
ba2-woman 2-four \\
`four women'}
\parbox[t]{2in}{ \ex \gll 
b-ùdã̂ {\bfseries bá}-tánɛ̀ \\
   ba2-woman 2-five \\
`five women'}
\end{xlist}
\end {exe}

The same noun phrase structure is used in the formation of complex numerals that involve an underlying arithmetic operation of multiplication. In this case, the agreeing numeral will follow a nominal base numeral, as shown in (\ref{ModNum2}), to form multiples of the base.

\begin {exe}
\ex\label {ModNum2}
\begin{xlist}
\parbox[t]{2in}{\ex \glll màwúmɔ̀ mábáà\\ 
ma-wúmɔ̀ má-báà\\
ma6-ten 6-two \\
 `twenty [10 x 2]'}
\parbox[t]{2in}{ \ex \glll màwúmɔ̀ máláálɛ̀ \\ 
ma-wúmɔ̀ má-láálɛ̀ \\
   ma6-ten 6-three \\
 `thirty [10 x 3]'}

%\vspace{1pc}

\parbox[t]{2in}{ \ex \glll bìbwúyà bínã̂ \\
bi-bwúyà {\bfseries bí}-nã̂\\ 
bi8-hundred 2-four \\
`four hundred [100 x 4]'}
\parbox[t]{2in}{ \ex \glll bàtɔ́dyínì bátánɛ̀ \\ 
ba-tɔ́dyínì {\bfseries bá}-tánɛ̀ \\
   ba2-thousand 2-five \\
`five thousand [1000 x 5]'}
\end{xlist}
\end {exe}


Agreeing numerals `2' through `5' can never modify singular nouns for semantic reasons. They therefore lack any singular counterparts. I still distinguish them from modifiers discussed in the next section since those modifiers do occur with singular forms.






\subsection{Modifiers with plural agreement only}
\label{sec:MODAgrPL}


\subsubsection{Genitive marker {\itshape ngá}}
There are two modifiers in Gyeli, the genitive marker {\itshape ngá} and {\itshape nyá} `big', which never take an agreement prefix for singular agreement classes, but require them for plural classes.%\footnote{Numerals `2' through `5' only appear with plural nouns for semantic reasons, but they differ from this group of modifiers in that they never occur with singular nouns.}
Based in this characteristic, I classify them as one modifier group. They differ, however, in many other properties. First, the genitive marker {\itshape ngá} only occurs in noun + noun constructions (\sectref{sec:CONC}), following the head noun it modifies. In contrast, {\itshape nyá} `big' precedes the head noun and is, together with the invariable negative polarity item {\itshape tɔ̀} the only element that can precede the head noun. The genitive marker {\itshape ngá} and {\itshape nyá} `big' also differ in the tonal pattern of their agreement prefixes: {\itshape ngá} takes a H tone CV prefix, while agreement prefixes of {\itshape nyá} surface with a L tone. 
\label{sec:GEN}

Gyeli has a split genitive/attributive system, using different sets of associativity markers depending on the status of the possessor noun. 
The genitive marker {\itshape ngá} is used instead of an attributive marker (\sectref{sec:ATT}) if the second constituent in a noun + noun construction is a proper name, as illustrated in (\ref{house}). This highlights the special status of proper names in contrast to common nouns (\sectref{sec:properN}).

\begin{exe}
\ex\label{house}
\begin{xlist}
\ex\label{house1}
 \gll    ndáwɔ̀ {\bfseries ngá} Àdà \\
                $\emptyset$9.house GEN $\emptyset$1.PN\\
    \trans `Ada's house'
\ex \label{house2}
  \gll     ndáwɔ̀ {\bfseries nyà} m-bvùlè \\
                $\emptyset$9.house 9:ATT N1-Bulu \\
    \trans `the Bulu\footnote{Bulu describes a neighboring ethnic group to the Bagyeli as well as their language which is classified as Bantu A74.} man's house'
\end {xlist}
\end {exe}

\noindent Further, the genitive marker is used in the interrogative pronoun constructions such as {\itshape pú'ù ngá nzá} `for whom' when the answer could potentially be a proper name.  In question words where something else than a proper name is expected as an answer, as in {\itshape pú'ù yá gyí} `for what', the attributive is used.

The genitive marker only takes an agreement marker if the preceding possessee noun occurs in the plural. If {\itshape ndáwɔ̀} `house' in (\ref{house}) was in its plural form, the example would change as in (\ref{housex}) with a plural marker on {\itshape ngá}.

\begin{exe}
\ex\label{housex}
\begin{xlist}
\ex\label{housex1}
 \gll    mà-ndáwɔ̀ {\bfseries má-ngá} Àdà \\
                ma6-$\emptyset$9.house 6-GEN $\emptyset$1.PN\\
    \trans `Ada's houses'
\ex \label{housex2}
  \gll     mà-ndáwɔ̀ {\bfseries má} m-bvùlè \\
                ma6-house 6:ATT N1-Bulu \\
    \trans `the Bulu man's houses'
\end {xlist}
\end {exe}


\noindent If it is singular, however, the genitive marker takes a default form {\itshape ngá}. Table \ref{Tab:nga} shows the agreement pattern of genitive markers with the non-agreeing singular forms in the left and the agreeing plural forms in the right column. 

\begin{table} 
\centering
\begin{tabular}{llp{1cm}lp{2cm}}
 \midrule
\multicolumn{2}{l}{Singular classes} & & \multicolumn{2}{l}{Plural classes} \\
 \midrule
cl. 1 & ngá & & cl. 2 & bá-ngá \\
cl. 3 & ngá & & cl. 4 & mí-ngá \\
cl. 5 & ngá & & cl. 6 & má-ngá \\
cl. 7 & ngá & & cl. 8 & bé-ngá \\
cl. 9 & ngá &    &     &              \\
 \midrule
\end{tabular}
\caption{Agreement marking of genitive markers}
\label{Tab:nga}
\end{table}

The agreement prefix, although it seems to be identical with the attributive marker, belongs prosodically to the genitive word {\itshape ngá}. In contrast, following speakers' intuitions, the attributive marker is prosodically an independent word. I therefore do not view agreeing plural forms of the genitive linker  as constructions containing both attributive and genitive markers. Instead, the H tone agreement prefixes are parallel to those used with agreeing plural numerals.

There is another logical possibility to explain the H tone on the agreement prefix, namely leftwards high tone spreading from the {\itshape -ngá} root. I rule this possibility out for two reasons.
First, high tone spreading from the right to the left does occur in Gyeli, but it seems to be restricted to the verbal domain (as with underlyingly toneless verb extension morphemes which are discussed in \sectref{sec:HTSr}). Therefore, it seems unlikely that the H tone from the {\itshape -ngá} root would spread leftwards onto the prefix.

Second, contrasting cases of L tone CV- agreement prefixes that occur with other modifiers, such as {\itshape -vúdû} `same, one' and {\itshape -fúsì} `different', suggest that the CV- agreement prefixes for the genitive marker (and numerals from `2' through `5') are indeed specified for a H tone. The other modifiers also start with a H tone stem, but they still have CV- agreement prefixes that surface with a L tone. There could be a rule that H tone spreading is restricted to a certain class of agreement targets, but given these two arguments, it seems unlikely. The ultimate proof against H tone spreading, namely checking what happens with the CV- prefixes if the stem starts with a L tone, is not testable because all modifier roots  that take a H tone CV- agreement prefix ({\itshape -ngá} and the numerals `2' through `5') start with a H or HL mora, but never with a L.



\subsubsection{{\itshape nyá} `big'}
\label{sec:nya}


{\itshape -nyá} meaning `big', `important', `luxurious', beautiful'  could qualify as an adjective since it denotes a property of a noun. The semantic difference between {\itshape nyá} `big' and the adjective {\itshape nɛ́nɛ̀} `big' is that the second typically refers to size as in (\ref{nya1}). {\itshape nyá}, however, talks more about the value as demonstrated in (\ref{nya2}). In Cameroonian French, value is often translated with size so that a {\itshape grand panier} `big basket' could, besides referring to the size, also talk about its value.

\begin{exe}
\ex\label{nya}
\begin{xlist}
\ex\label{nya1}
 \gll  nkwě wá nɛ́nɛ̀\\
          $\emptyset$3.basket 3:ATT big \\
    \trans `a/the big basket'
\ex \label{nya2}
  \gll   nyá nkwě \\
             big $\emptyset$3.basket \\
    \trans `a/the important/beautiful/luxurious basket'
\end {xlist}
\end {exe}



Agreement of {\itshape nyá}  is only marked if the head noun comes in a plural form. If the head noun is singular, {\itshape nyá} is invariable as shown in Table \ref{Tab:nya}. This behavior is similar to the genitive marker discussed in \sectref{sec:GEN}. 

\begin{table} 
\centering
\begin{tabular}{llll}
 \midrule
\multicolumn{4}{l}{Singular classes} \\
 \midrule
cl. 1 & nyá & m-ùdì & `important person'  \\
cl. 3 & nyá & nkwě  & `great basket'        \\
cl. 5 & nyá & le-dùndá & `big sparrow'      \\
cl. 7 & nyá & lé       & `great tree'            \\
cl. 8b & nyá & bwálɛ̀ & `beautiful canoe'  \\
cl. 9 & nyá &   ndáwɔ̀ & `luxurious house'   \\
 \midrule
\multicolumn{4}{l}{Plural classes} \\
 \midrule
cl. 2 & ba-nyá & b-ùdì & `important people' \\
cl. 4 & mi-nyá & mì-nkwě & `great baskets'  \\
cl. 6 & ma-nyá & mà-dùndá & `big sparrows' \\
cl. 8a & be-nyá & bè-lé & `great trees' \\
 \midrule
\end{tabular}
\caption{{\itshape nyá} in various agreement classes}
\label{Tab:nya}
\end{table}

\noindent Another particularity is the syntactic position of {\itshape nyá}, preceding the noun whereas basically all other modifiers follow the noun.











\subsection{Modifiers with agreeing free morpheme}
\label{sec:MODAgrForm}

There are two nominal modifiers in Gyeli which do not express agreement with the head noun through a prefix that attaches to a root that is consistent across different agreement classes, but that have free agreeing morphemes which differ across agreement classes. This is the case for demonstratives and for the attributive marker.
		
		

\subsubsection{Demonstratives}
\label{sec:DEM}



Gyeli has two sets of demonstratives distinguishing different degrees of distance between the speaker and the object or person he or she is talking about. One set of demonstratives, the proximal demonstratives, refers to objects or persons close to the speaker. Distal demonstratives are employed when the object or person in question is further away from the speaker (but not necessarily close to the addressee). 

Proximal and distal demonstratives are formally distinguished by different tonal patterns and vowel lengthening of the distal pronouns. Table \ref{Tab:dem} contrasts the two sets of demonstratives. While proximal demonstratives end in a simple vowel with a falling HL tonal pattern, distal demonstratives all have a lengthened vowel with a H tone.

\begin{table} 
\centering
\begin{tabular}{r|p{3cm}|p{3cm}}
 \midrule
 & proximal & distal \\
  \midrule
 1 & \itshape{nû} & \itshape{núú} \\
 2 & \itshape{bâ} & \itshape{báá} \\
 3 & \itshape{wɔ̂} & \itshape{wɔ́ɔ́} \\
4 & \itshape{mî} & \itshape{míí} \\
5 & \itshape{lɛ̂} & \itshape{lɛ́ɛ́} \\
6 & \itshape{mâ} & \itshape{máá} \\
7 & \itshape{yî} & \itshape{yíí} \\
8 & \itshape{bî} & \itshape{bíí} \\
9 & \itshape{nyî} & \itshape{nyíí} \\
  \midrule
\end{tabular}
\caption{Gyeli demonstratives}
\label{Tab:dem}
\end{table}

\noindent Both proximal and distal demonstratives follow the noun they modify in a noun phrase as shown in (\ref{DEM}).

\begin{exe}
\ex\label{DEM}
\begin{xlist}
\ex \label{DEM1}
  \gll     m-ùdì nû \\
                N1-man 1.DEM.PROX \\
    \trans `this man'
\ex\label{DEM2}
 \gll     m-ùdì núú \\
                N1-man 1.DEM.DIST \\
    \trans `that man'
\end {xlist}
\end {exe} 

These demonstratives are also used as presentational or identificational markers in non-verbal predicates of the pattern `This is a house.' Such constructions are discussed in \sectref{sec:nonverbalC}.





\subsubsection{Attributive markers}
\label{sec:ATT}

Attributive markers constitute another class of function words that agree with their head noun. In Bantu studies, they are also called genitive or associative markers. Gyeli has a split genitive system with two different paradigms of genitive, or associativity, indicating constructions. I label one paradigm as `attributive' and the other as `genitive'.  The difference is explained in \sectref{sec:GEN}.


Attributive markers serve as a linking element between a noun and typically another noun, as shown in (\ref{CONM}). %Note that the other noun cannot be a proper name. In that case, a different paradigm of genitive markers would be used. 
Attributive markers also link a noun to a verb, an adjective, or an interrogative. The different constructions and their constituents are discussed in \sectref{sec:GEN} on attributive constructions. At this point, I will only present the agreement target, namely the attributive marker itself.

\begin{exe}
\ex\label{CONM}
\begin{xlist}
\ex \label{CONM1}
  \gll     síngì {\bfseries yá} jìí \\
                $\emptyset$7.cat 7:ATT $\emptyset$7.forest \\
    \trans `forest cat'
\ex\label{CONM2}
 \gll    lè-lɔ̂ {\bfseries lé} síngì \\
                le5-ear 5:ATT $\emptyset$7.cat \\
    \trans `the cat's ear'
\end {xlist}
\end {exe}

Attributive markers are also used in relative clauses, as exemplified in (\ref{RELPro}) and discussed in detail in \sectref{sec:Relativeclauses}.

\begin{exe}
\ex\label{RELPro}
\begin{xlist}
\ex \label{RELPro1}
  \gll     síngì yá yí kwè \\
                $\emptyset$7.cat 7:ATT 7.PRES fall \\
    \trans `the cat that falls'
\ex\label{RELPro2}
 \gll     síngì yá mɛ́ nyɛ̂ \\
                $\emptyset$7.cat 7:ATT 1\textsc{sg}.PRES see \\
    \trans `the cat that I see'
\end {xlist}
\end {exe}

\citet{meeussen67}, and later \citet[219]{velde2013}, posit that the canonical form for Bantu attributives is AGR-{\itshape a}, a root -{\itshape a} which is preceded by an agreement prefix. Many Gyeli attributives follow this canonical form. Exceptions to this tendency are found, however, in classes 4, 5, and 8 which come with high and mid vowel roots rather than with -{\itshape a}, as shown in Table \ref{Tab:AGRCON}.  Attributive markers in Gyeli typically have a H tone, except for those of classes 1 and 9 which both come with a L tone. 

\begin{table} 
\centering
\begin{tabular}{ll}
 \midrule
AGR class & ATT marker \\
  \midrule
 1 & wà  \\
 2 & bá \\
 3 & wá \\
4 & mí \\
5 & lé \\
6 & má \\
7 & yá \\
8 & bé \\
9 & nyà \\
  \midrule
\end{tabular}
\caption{Attributives in the different agreement classes}
\label{Tab:AGRCON}
\end{table}














\subsection{Prenominal invariable modifiers}
\label{sec:InvQUANT1}

Elements that can occur prenominally in Gyeli are restricted in number and distribution. In simple noun phrases, only {\itshape nyá} `big' (\sectref{sec:nya}) and {\itshape tɔ̀} `any' (\sectref{sec:to}) can precede the noun. {\itshape nyá} `big' agrees with the head noun only if the noun is plural, otherwise it is invariable; {\itshape tɔ̀}`any' is always invariable. Other prenominal elements precede second constituents in noun + noun constructions, serving as connectors. They also differ in their agreement behavior ranging from agreeing elements such as the attributive marker (\sectref{sec:ATT}) to those that only agree with plural nouns as the genitive marker (\sectref{sec:GEN}) and the similative marker (\sectref{sec:SIMword}) that is invariable. I discuss the two invariable prenominal elements in the following, namely the negative polarity item {\itshape tɔ̀} and the similative marker {\itshape ná}.


\subsubsection{Negative polarity item {\itshape tɔ̀} `any'}
\label{sec:to}

The negative polarity item {\itshape tɔ̀} `any' does not agree with the head noun, as shown in (\ref{to}). {\itshape tɔ̀} `any' never agrees, no matter if it precedes a singular or plural noun. In that, it differs from the genitive marker {\itshape ngà} (\sectref{sec:GEN}) that agrees with plural nouns.

\begin{exe}
\ex\label{to}
\begin{xlist}
\ex \label{to1}
  \gll    mɛ̀ɛ́ nyɛ́-lɛ́ [{\bfseries tɔ̀} m-ùdì] \\
              1\textsc{sg}.PRES.NEG see-NEG {\bfseries any} N1-person \\
    \trans `I don't see any person.'
\ex \label{to2}
  \gll    mɛ̀ɛ́ nyɛ́-lɛ́ [{\bfseries tɔ̀} b-ùdì] \\
              1\textsc{sg}.PRES.NEG see-NEG {\bfseries any} ba2-person \\
    \trans `I don't see any people.'
\end {xlist}
\end{exe}

The use of {\itshape tɔ̀} in negated sentences is grammatically not obligatory, as shown in (\ref{to3}), where the same sentence as in (\ref{to1}) occurs without {\itshape tɔ̀} `any'. Semantically, however, there is a difference in that no person at all is seen in (\ref{to1}), while (\ref{to2}) negates a specific, known person.

\begin{exe}
\ex\label{to3}
 \gll  mɛ̀ɛ́ nyɛ́-lɛ́ m-ùdì  \\
          1\textsc{sg}.PRES.NEG see-NEG N1-person  \\
    \trans `I don't see the person.'
\end{exe}


%***also as negative coordinator***


\subsubsection{Similative marker {\itshape ná}}
\label{sec:SIMword}

The similative marker {\itshape ná} occurs both as a free morpheme and a prefix. The free morpheme {\itshape ná} functions as a connector in a noun + noun construction that is restricted to the use of the noun {\itshape jínɔ̀} `name' as the first constituent and a proper name in the second constituent, as shown in (\ref{SIMword}).


\begin{exe} 
\ex\label{SIMword} 
  \glll èè mɛ̀ jínɔ̀ {\bfseries ná} Màmà \\
         èè mɛ̀ j-ínɔ̀ ná Màmà \\
       yes 1\textsc{sg} le5-name SIM $\emptyset$1.PN  \\
    \trans `Yes, my name is Mama.'
\end{exe}

The free similative marker is invariable, even if the noun + noun construction has plural constituents. As illustrated in (\ref{SIMword2}), number has to be identical in the first and second constituent, but the connecting similative marker {\itshape ná} does not change.

\begin{exe} 
\ex\label{SIMword2} 
  \glll bà mínɔ̀ {\bfseries ná} Màmà nà Màmbì \\
         bà m-ínɔ̀ ná Màmà nà Màmbì \\
        2 ma6-name SIM $\emptyset$1.PN  CONJ $\emptyset$1.PN \\
    \trans `Their names are Mama and Mambi.'
\end{exe}

\noindent  The use of the similative marker {\itshape na}- as a prefix is more productive than the free connector form and is discussed in \sectref{sec:SIM}.





\subsection{Postnominal invariable modifiers}
\label{sec:InvQUANT2}

Most modifiers in Gyeli occur after the noun. This is also true for most non-agreeing modifiers, such as invariable numerals, some quantifiers, and the free anaphoric marker {\itshape ndɛ̀}.

\subsubsection{Invariable numerals} 
\label{sec:InvNUM}


Gyeli has monomorphemic cardinal numerals which do not agree with the noun, as shown in (\ref{ModNumx2}). As such, they might be thought of belonging to the same category of adjectives (\sectref{sec:QUAL}). In contrast to adjectives, however, they never occur in a construction involving an attributive marker.

\begin {exe} \ex\label {ModNumx2}
\begin{xlist}
\parbox[t]{2in}{ \ex \gll
 b-ùdã̂ ntùɔ́\\ 
ba2-woman six \\
`six women'}
\parbox[t]{2in}{ \ex \gll
b-ùdã̂ mpúɛ̀rɛ́ \\
   ba2-woman seven \\
 `seven women'}

%\vspace{1pc}

\parbox[t]{2in}{ \ex \gll
 b-ùdã̂ lɔ̀mbì\\ 
ba2-woman eight \\
`eight women'}
\parbox[t]{2in}{ \ex \gll
b-ùdã̂ rèbvùá \\
   ba2-woman nine \\
 `nine women'}
\end{xlist}
\end {exe}


\subsubsection{Quantifier {\itshape bvùbvù} `many, much'} 

{\itshape bvùbvù} `many, much'  is an intersective quantifier. \citet[388]{zerbian2008} define {\itshape intersective} quantifiers as ``quantifiers whose truth conditions can be given in terms of the intersection of the noun meaning and the predicate meaning."  Other intersective quantifiers are, for instance, `several', `few', `a certain/other', `some' or `no'. The authors state that most intersective quantifiers in Bantu languages agree with their head noun. This is not true for Gyeli which has a range of non-agreeing quantifiers (or uses attributive constructions (\sectref{sec:NomQUANT}) in order to express quantifiers such as `many' or `few'). 

{\itshape bvùbvù} `many, much' is not sensitive to a mass/count distinction and occurs both with countable and uncountable nouns alike, as shown in (\ref{Invbvu1}) and (\ref{Invbvu2}).


\begin{exe}
\ex\label{Invbvu}
\begin{xlist}
\ex\label{Invbvu1}
 \gll  b-ùdì  bvùbvù \\
          ba2-people many  \\
    \trans `many people'
\ex \label{Invbvu2}
  \gll    mà-jíwɔ́ bvùbvù \\
              ma6-water much \\
    \trans `much water'
\end {xlist}
\end {exe}

This quantifier has a nominal counterpart in agreement class 9 which can be used in a noun + noun attributive construction (\sectref{sec:NomQUANT}). The nominal quantifier has a different tone pattern, however: {\itshape bvúbvù nyà}. It seems to be the more marked form which occurs less frequently than the invariable modifier. Possible meaning differences are subtle; speakers claim that both mean the same and can be used in the same contexts.





\subsubsection{Quantifier {\itshape mànjìmɔ̀} `whole, entire'}
\label{sec:mandjimo}

{\itshape mànjìmɔ̀} `whole, entire' is another invariable quantifier that follows the head noun, as in (\ref{whole}). Despite the similarity to the nominal modifier {\itshape njìmɔ̀ wá} `a certain' plus something that looks like a class 6 prefix, {\itshape mànjìmɔ̀} is not a noun since it lacks noun properties such as the possibility to be modified by, for instance, demonstratives or possessive pronouns, or entering a noun + noun attributive construction as the head.


\begin{exe}
\ex\label{whole}
\begin{xlist}
\ex \label{whole1}
  \gll     púsí mànjìmɔ̀\\
                $\emptyset$7.bottle whole \\
    \trans `the whole bottle'
\ex\label{whole2}
 \gll     ndáwɔ̀ mànjìmɔ  \\
             $\emptyset$9.house whole  \\
    \trans `the entire house'
\ex\label{whole3}
 \gll   bè-síngì mànjìmɔ  \\
             be8-cat whole   \\
    \trans `the entire cats'
\end {xlist}
\end {exe}

{\itshape mànjìmɔ̀} is sensitive to a mass/count distinction in that it does not appear with uncountable nouns, neither liquids nor granular aggregates, as shown in (\ref{wholeno}). Using {\itshape mànjìmɔ̀} with mass nouns requires a specification of the physical entity, for instance a bottle as in (\ref{wholeno3}).

\begin{exe}
\ex\label{wholeno}
\begin{xlist}
\ex \label{wholeno1}
  \gll     *mà-tàngɔ̀ mànjìmɔ̀\\
                ma6-palm.wine whole \\
    \trans `the whole palm wine'
\ex\label{wholeno2}
 \gll     *ndísì mànjìmɔ  \\
             $\emptyset$3.rice whole  \\
    \trans `the entire rice'
\ex \label{wholeno3}
  \gll     púsí (yá) má-vúdɔ̀ mànjìmɔ̀\\
                $\emptyset$7.bottle (7:CON) ma6-oil whole \\
    \trans `a whole bottle of oil'
\end {xlist}
\end {exe}

\noindent In contrast to the singular form of granular aggregate mass nouns which cannot occur with {\itshape mànjìmɔ̀}, their plural counterpart allows for its use as in (\ref{massPL}). In this case, however, it is understood that the noun comes in packaged entities, for instance in sachets or bag, or that different types of the noun are involved.

\begin{exe}
\ex\label{massPL}
  \gll     mì-ndísì mànjìmɔ̀ \\
                mi4-rice whole \\
    \trans `the whole rice (= all its types or packages)'
\end {exe}








\subsubsection{Free anaphoric marker {\itshape ndɛ̀}}
\label{sec:ANAfree}

The anaphoric marker {\itshape ndɛ̀} occurs as a free morpheme after the identificational marker {\itshape wɛ́}, as in (\ref{ANA41t}).


\begin{exe} 
\ex\label{ANA41t}
  \glll kàndá wɛ́ {\bfseries ndɛ̀} \\
        kàndá wɛ́ ndɛ̀ \\
        $\emptyset$7.proverb ID ANA  \\
    \trans `The story is this.'
\end{exe}

\noindent It is also a free morpheme after nouns, as in (\ref{ANA42t}).

\begin{exe} 
\ex\label{ANA42t}  
  \glll  àà kfúmá {\bfseries ndɛ̀} wà Nlúnzɔ̀ \\
        àà kfúmá ndɛ̀ wà Nlúnzɔ̀ \\
        ECXL $\emptyset$1.chief ANA  1:ATT $\emptyset$1.PN  \\
    \trans `Ah, that chief from Nlunzo!'
\end{exe}

As discussed in \sectref{sec:ANASuff}, {\itshape ndɛ̀} occur as a suffix when it attaches to demonstratives.

















\section{Elements of the verbal complex}
\label{sec:VerbAd}

In this section, I describe the elements that occur in a verbal predicate other than the verb, which has been outlined in \sectref{sec:V}. These elements include the subject-tense-aspect-mood-polarity (STAMP) marker and two verbal particles that follow the inflected verb form.



\subsection[The subject-tense-aspect-mood-polarity marker]{The subject-tense-aspect-mood-polarity marker}
\label{sec:SCOP}

The subject-tense-aspect-mood-polarity (STAMP) marker, following terminology coined by Anderson (\citeyear{anderson2011b}, \citeyear{anderson2015}), is a clitic directly preceding the inflected verb form. As a portmanteau morpheme, it encodes subject agreement as well as tense, mood, aspect, and negation. Table \ref{Tab:SCOPAGR1} shows the segmental shape of the STAMP marker in the different agreement classes, omitting the tonal pattern which changes across tense-mood, aspect, and negation categories.
The tonal patterns of the STAMP marker are discussed in \sectref{sec:GramTM}.

\begin{table}[!h]
\centering
\scalebox{0.85}{
\begin{tabular}{|p{1cm}|llllp{.8cm}llllllll|}
 \midrule
TM CAT & 1\textsc{sg} & 2\textsc{sg} & 1\textsc{pl} & 2\textsc{pl} & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\
 \midrule
\textsc{pres} &  mɛ  & wɛ & ya & bwa & a/ nyɛ/ nu & ba & wu & mi & le & ma & yi & be & nyi  \\
 \midrule
\end{tabular}}
\caption{Shape of the STAMP marker in different AGR classes}
\label{Tab:SCOPAGR1}
\end{table}

The tonal pattern and sometimes vowel length of the STAMP marker change across different tense-mood categories, as shown in Table \ref{Tab:SCOPAGR} which lists the STAMP markers' form and surface tones for all agreement classes in all tense-mood categories.  


\begin{table}[!h]
\centering
\scalebox{0.85}{
\begin{tabular}{|p{1cm}|llllp{.8cm}llllllll|}
 \midrule
TM CAT & 1\textsc{sg} & 2\textsc{sg} & 1\textsc{pl} & 2\textsc{pl} & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\
 \midrule
\textsc{pres} &  mɛ́   & wɛ́ & yá & bwá(á) & á/ nyɛ́/ nú & bá & wú & mí & lé & má & yí & bé & nyí  \\
\textsc{inch} &   mɛ̀ɛ́ & wɛ̀ɛ́ & yàá & bwàá & àá & bàá & wùú & mìí & lèé & màá & yìí & bèé & nyìí  \\ 
\textsc{fut} & {\bfseries mɛ̀ɛ̀} &  {\bfseries wɛ̀ɛ̀} & yáà & bwáà & {\bfseries àà/ nyɛ̀ɛ̀/ nùù} & báà & wúù & míì & léè & máà & yíì & béè & nyíì  \\
% \midrule
\textsc{pst1} & mɛ & wɛ & ya & bwa(a) & a/ nyɛ/ nu & ba & wu & mi & le & ma & yi & be & nyi  \\
\textsc{pst2} & mɛ́ɛ̀ & wɛ́ɛ̀ & yáà & bwáà & áà/ nyɛ́ɛ̀/ núù & báà & wúù & míì & léè & máà & yíì & béè & nyíì \\
% \midrule
\textsc{imp} & ---  & ---   & yá & --- & --- & --- & --- & --- & --- & --- & --- & --- & ---  \\ 
\textsc{sbjv} & mɛ́   & wɛ́ & yá & bwá(á) & á/ nyɛ́/ nú & bá & wú & mí & lé & má & yí & bé & nyí  \\
 \midrule
\end{tabular}}
\caption{Patterns of STAMP markers in different AGR classes and TM categories}
\label{Tab:SCOPAGR}
\end{table}

\noindent The \textsc{fut} category has an exceptional tonal pattern for certain agreement classes which are marked in bold. The vowel of the second person plural is either pronounced with a long or a short vowel if the tone is not a contour tone, i.e.\ if it is either H or L. 

Class 1 has {\itshape a} as a basic form and an alternate form {\itshape nyɛ}\footnote{This form could originate from Kwasio.}. At the same time, {\itshape nyɛ} is identical with the non-subject pronoun of agreement class 1. Both forms are equally used and speakers state that both are part of the Gyeli language, although the {\itshape a} form is more frequently found in texts.  Also, agreement class 1 has a third alternate form, namely {\itshape nu} which is identical with the class 1 demonstrative. It can, however, also be used as STAMP marker with the specific tonal pattern for each tense-mood category. In this, the class 1 STAMP marker is exceptional because demonstratives of other agreement classes cannot function as STAMP marker.

\paragraph{Toneless \textsc{past 1} category} I suggest that, underlyingly, the L surface form of the \textsc{pst1} category is tonally not specified and only surfaces phonetically as L. This is comparable to other grammatical morphemes such as noun class prefixes or verbal derivation morphemes as discussed in \sectref{sec:toneless}. I view this phonetically L form as a tonally underspecified default form because it does not only occur in the \textsc{past 1} category, but also serves as general default form in other tense-mood categories when these are combined with true auxiliaries encoding aspect (\sectref{sec:ComplAUX}). It further provides the basic form from which the \textsc{present} category is derived with a H tone. Consequently, in the glossing of examples, the surface L STAMP markers are represented as being toneless in the underlying line. \textsc{pres} STAMP forms are underlyingly represented as toneless STAMP markers which receive a H tone, characterizing this category.


\paragraph{Tone pattern in the \textsc{future} category} 
As shown in Table \ref{Tab:SCOPAGR}, the general pattern for the \textsc{future} is a long vowel with a HL tonal melody. While in other tense-mood categories the tonal and vowel length pattern is the same for each agreement class, in the \textsc{future}, the first and second person singular as well as the class 1 STAMP marker deviate from this pattern, having a long vowel with a L tonal melody, as in (\ref{FUTexept}). 

\begin{exe} 
\ex\label{FUTexept} 
\begin{xlist}
\ex mɛ̀ɛ̀ dè `I will eat'
\ex wɛ̀ɛ̀ dè `you will eat'
\ex àà/nyɛ̀ɛ̀ dè `s/he will eat'
\end{xlist}
\end{exe}




The STAMP marker precedes the finite verb, but is not part of the verb as it can, in fast speech, be assimilated or even omitted in certain tense-mood categories. I outline both cases in turn.

\paragraph{STAMP marker assimilation}
Depending on the morphophonological shape of the STAMP marker, this clitic can undergo assimilation with preceding vocalic material in fast speech. This applies mainly to the agreement class 1 STAMP marker whose segmental material consists of the vowel {\itshape a}. Given that it is not preceded by a consonant, unlike the STAMP markers of all other agreement classes, it can assimilate with the final vowel of a preceding verb or noun.

An example of STAMP marker assimilation with both preceding verbs and nouns is provided in (\ref{SCOPAs1}). In the first instance, the STAMP marker assimilates to the verb {\itshape njì} `come' of the preceding phrase. Thus, STAMP marker assimilation in fast speech is not restricted to in-phrase assimilation, but can also cross phrase boundaries. 

\begin{exe} 
\ex\label{SCOPAs1}
  \glll  à nj{\bfseries â} dyùmɔ́ bùd{\bfseries àà} dyùmɔ́ bùd{\bfseries àà} dyùmɔ́ bùd{\bfseries àà} dyùmɔ́ bùdì  \\
           a nji-H a dyùmɔ-H b-ùdì a dyùmɔ-H  b-ùdì a dyùmɔ-H b-ùdì   \\
          1.PST1 come-R 1.PST1 heal-R ba2-person 1.PST1 heal-R ba2-person 1.PST1 heal-R ba2-person  \\
    \trans `He came, he was healing people.'
\end{exe}

\noindent In the other assimilation instances in (\ref{SCOPAs1}), the STAMP marker assimilates to the nominal object {\itshape bùdì} `people', also of the previous phrase. In both cases, the final vowel of the noun is elided while the vowel of the STAMP marker surfaces. At the same time, the tone of the omitted vowel survives, as seen with the contour tone on [njí + à]  $\rightarrow$ /njâ/. In the second instance, while vowel quality is assimilated to the STAMP marker, both tone and vowel length survive, surfacing in a long vowel: [bùdì + à] $\rightarrow$ /bùdàà/.

\paragraph{STAMP marker assimilation with proper names}
As seen in the previous example, in STAMP marker assimilation it is usually the preceding vocalic material of a noun or verb that is deleted. This is different for STAMP marker assimilation with proper names. Proper names do not change their vowel quality, but assimilate tonally to the class 1 STAMP marker whose vocalic material is being elided, as shown in (\ref{SCOPAs2}). 

\begin{exe}
\ex\label{SCOPAs2}
\begin{xlist}
\ex\label{SCOPAs2a}
\glll Màmbì á kwè   $\rightarrow$ /Màmb{\bfseries í} kwè/ \\ 
      Màmbì a-H kwè \\
	$\emptyset$1.PN 1-PRES fall\\
	\trans `Mambi falls.'
\ex \label{SCOPAs2b}
  \glll Màmbì àá kwè  $\rightarrow$ /Màmb{\bfseries ìí} kwè/ \\
        Màmbì àá kwè \\
          $\emptyset$1.PN 1.INCH fall    \\
    \trans `Mambi is at the beginning of falling.'
\end{xlist}
\end{exe}

Tonal changes on the proper name do not depend on tonal or phonological patterns of the name, but are controlled by the noun's feature of being a proper name (\sectref{sec:properN}).  The fact that proper names receive special morphosyntactic treatment in Gyeli is also seen in the split genitive system (\sectref{sec:GEN}). 

If the proper name's final tone and the STAMP marker's tone are identical, there is no tonal or vocalic surface change, but the STAMP marker simply is elided, as shown in (\ref{SCOPAs3a}) for  the proper name {\itshape Màmbì} ending in a L tone and a following L STAMP marker and, in (\ref{SCOPAs3b}), the proper name {\itshape Bìyã́} ending in a H tone in combination with a \textsc{pres} H tone STAMP marker.


\begin{exe}
\ex\label{SCOPAs3}
\begin{xlist}
\ex \label{SCOPAs3a}
   \glll   Màmbì à kwé   $\rightarrow$ /Màmbì kwé/ \\
          Màmbì a kwè-H \\
            $\emptyset$1.PN 1.PST1 fall-PST  \\
    \trans `Mambi fell.'
\ex\label{SCOPAs3b}
\glll  Bìyã́ á sàgà $\rightarrow$ /Bìyã́ sàgà/ \\
       Bìyã́ a-H sàga \\
           $\emptyset$1.PN 1-PRES frighten    \\
    \trans `Biyang is frightened.'
 \end{xlist}
\end{exe}

\noindent These cases are thus rather instances of STAMP marker omission than STAMP marker assimilation, which leads to the next section on STAMP marker omisson.



\paragraph{STAMP marker omission}
Under certain circumstances, the STAMP marker can be elided rather than assimilated. STAMP marker omission requires some conditions. First, the clause has to be either in the \textsc{present} or the \textsc{recent past}, as in (\ref{SCOPOMtense})\footnote{In this example, the class 1 STAMP marker takes the alternate shape of the demonstrative rather than the default shape {\itshape a}. The shape of the class 1 STAMP marker does not, however, influence the possibility of its omission.}, while the other tense-mood categories (\sectref{sec:GramTM}) exclude STAMP marker omission.  The parentheses indicate that the use of the STAMP marker is optional while a lack of parentheses indicates that the STAMP marker has to be used obligatorily. 

\begin{exe}
\ex\label{SCOPOMtense}
\begin{xlist}
\ex \label{SCOPOMtense1}
  \glll  kálɛ́ (nú) kwè $\rightarrow$ /kálɛ́ kwè/ \\
         kálɛ́ nu-H kwè \\
              $\emptyset$1.sister 1-PRES fall \\
    \trans `The sister falls.'
\ex\label{SCOPOMtense2}
  \glll  kálɛ́ (nù) kwé $\rightarrow$ /kálɛ́ kwé/ \\
          kálɛ́ nu kwé  \\
              $\emptyset$1.sister 1.PST1 fall \\
    \trans `The sister fell (recently).'
\ex \label{SCOPOMtense3}
  \glll  kálɛ́ núù kwé $\rightarrow$ */kálɛ́ kwé/ \\
       kálɛ́ núù kwè-H  \\
              $\emptyset$1.sister 1.PST2 fall-PST \\
    \trans `The sister fell (a long time ago).'
\ex\label{SCOPOMtense4}
  \glll  kálɛ́ nùù kwè $\rightarrow$ */kálɛ́ kwè/ \\
          kálɛ́ nùù kwè \\
              $\emptyset$1.sister 1.FUT fall \\
    \trans `The sister will fall.'
\ex\label{SCOPOMtense5}
  \glll  kálɛ́ nùú kwè $\rightarrow$ */kálɛ́ kwè/ \\
        kálɛ́ nùú kwè \\
              $\emptyset$1.sister 1.INCH fall \\
    \trans `The sister starts to fall.'
\end{xlist}
\end{exe}

Second, a nominal subject has to surface, excluding, however, all nouns with a CV noun class prefix, as in (\ref{NCOM1}), and plural subject nominals, as in (\ref{NCOM2}).\footnote{Potential STAMP marker omission was checked for a range of nouns, controlling for different tonal and phonological patterns, noun class affiliation, number, animacy, and different verbs. For simplicity, I only contrast two nouns in their singular and plural form, both belonging to gender 5/6. Most nouns in this gender have a CV prefix in both class 5 and class 6, but preceding a vowel initial stem, the prefix only consists of a consonant, providing a good testing ground for different phonological environments.} (This is parallel to the potential omission of the attributive marker discussed in \sectref{sec:CONOM}, which has similar conditioning factors.)


%\noindent The \textsc{inch} category is somewhere in between. Mostly, omission is not allowed, as for example in (\ref{INCHOM1}), but there are some exceptions as in (\ref{INCHOM2}). 

%\begin{exe}
%\ex\label{INCHOM}
%\begin{xlist}
%\ex \label{INCHOM1}
 % \gll  dàá kwè\\
 %             5.crab.INCH fall \\
  %  \trans `The crab started to fall.'
%\ex\label{INCHOM2}
 %\gll  sɔ̀ yìí kwè \\
 %        7.saw 7.INCH fall \\
  %  \trans `The saw started to fall.'
%\end{xlist}
%\end{exe}



\begin{exe}
\ex\label{NCOM1}
\begin{xlist}
\ex \label{NCOM1a}
   \glll   lèndzólɛ̀ lé kwè   $\rightarrow$ */lèndzólɛ̀ kwè/ \\
          le-ndzólɛ̀ le-H kwè   \\
            le5-tear 5-PRES fall  \\
    \trans `The tear falls.'
\ex\label{NCOM1b}
\glll   màndzólɛ̀ má kwè   $\rightarrow$ */màndzólɛ̀ kwè/ \\
          ma-ndzólɛ̀ ma-H kwè   \\
            ma6-tear 6-PRES fall  \\
    \trans `The tears fall.' 
\end{xlist}
\end{exe}

In (\ref{NCOM1}), both examples are excluded from STAMP marker omission, based on the CV noun class prefix. In contrast, in (\ref{NCOM2}) with consonantal noun class prefixes,  only the plural noun in (\ref{NCOM2b}) does not allow STAMP marker omission, but its singular counterpart in (\ref{NCOM2a}) does allow it.

\begin{exe}
\ex\label{NCOM2}
\begin{xlist}
\ex \label{NCOM2a}
   \glll    jáwɛ̀ (lé) kwè  $\rightarrow$ / jáwɛ̀ kwè/ \\
             j-áwɛ̀ le-H kwè  \\
            le5-goliath.frog 5-PRES fall  \\
    \trans `The goliath frog falls.'
\ex\label{NCOM2b}
\glll   máwɛ̀ má kwè   $\rightarrow$ */máwɛ̀ kwè/ \\
          m-áwɛ̀ ma-H kwè   \\
            ma6-goliath.frog 6-PRES fall  \\
    \trans `The goliath frogs fall.' 
\end{xlist}
\end{exe}

\noindent At the same time, these two examples also illustrate that animacy does not play a role, neither does general noun class affiliation since both examples belong to gender 5/6.

The STAMP marker can also be elided with more complex noun phrases such as noun + possessive constructions, as in (\ref{OmNPoss}). The tense reading comes from the \textsc{absolute completive} marker {\itshape mɔ̀} (\sectref{sec:mo}, which is restricted to \textsc{recent past}.

\begin{exe} 
\ex\label{OmNPoss}
  \glll nyɛ̀ nâ [sɔ́ wɔ́ɔ̀]\textsubscript{NP} nɔ̀ɔ́ mɔ̀ mwánɔ̀ \\
        nyɛ nâ sɔ́ w-ɔ́ɔ̀ nɔ̀ɔ̀-H mɔ̀ m-wánɔ̀ \\
      1 COMP $\emptyset$1.friend 1-POSS.2\textsc{sg} take-R PRF 1-child   \\
    \trans `She [says] `Your friend has taken the child.''
\end{exe}

There are also examples in the corpus showing that noun + noun attributive constructions may occur without a STAMP marker, as in (\ref{OmNN}). The tense reading of this utterance is ambiguous. As STAMP marker omission only occurs in \textsc{present} and \textsc{recent past}, this narrows possible interpretations down. In (\ref{OmNN}), formal marking allows both tense interpretations. Through common ground, however, it is clear that it has to be the \textsc{present} since all participants know that the road has not been built yet.

\begin{exe} 
\ex\label{OmNN} 
  \glll  mɛ́ dyúwɔ́ nâ [mpàgó wá pɔ́dɛ̀]\textsubscript{NP} lã́ vâ \\
        mɛ-H dyúwɔ-H nâ mpàgó wá pɔ́dɛ̀ lã̀-H vâ \\
            1\textsc{sg}-PRES hear-R COMP $\emptyset$3.street 3:ATT $\emptyset$1.port pass-R here \\
    \trans `I hear that the road to the port passes [= will pass] here. '
\end{exe}

Third, the STAMP marker can also be elided when the subject noun phrase is expressed by a pronoun, as in (\ref{OmInterr}) with the interrogative pronoun {\itshape nzá} `who'. The tense reading in this example comes from the shape of the \textsc{progressive} auxiliary which has a different form for the \textsc{past} (\sectref{sec:PROG}).


\begin{exe} 
\ex\label{OmInterr}
  \glll nzá nzíí mɛ̂ nyɛ̂ \\
    nzá nzíí mɛ̂ nyɛ̂ \\
         who PROG.PRES 1\textsc{sg}.NSBJ see\\
    \trans `Who is seeing me?'
\end{exe}



In a quotative index, signaling reported discourse, both the nominal subject and the STAMP marker can be elided, as the STAMP marker that would normally precede {\itshape kì} `say' in (\ref{SOm}).


\begin{exe} 
\exN\label{SOm} 
  \glll à kɛ̃́ɛ̃̀ nyî pɛ̀ dyúwɔ̀ à dígɛ́ɛ̀ à díg-â dígɛ́ɛ̀. kì nâ nzá nyɛ́ mɛ̂ \\
        a kɛ̃́ɛ̃̀ nyî pɛ̀ dyúwɔ̀ a dígɛ́ɛ̀ a dígɛ́ɛ̀ a dígɛ́ɛ̀ kì nâ nzá nyɛ̂-H mɛ̂ \\
       1.PST1 go.PRF enter there on.top 1.PST1 watch.PRF 1.PST1 watch.PRF 1.PST1 watch.PRF say COMP who see-R 1\textsc{sg}.NSBJ  \\
    \trans `He went inside there on top and watched and watched and watched. [He] says: `Who sees me?'
\end{exe}

\noindent Following \citet[105]{guldemann2008}, encoding the speaker in a quotative index is permissible in some languages since the speaker ``is normally the central character in a given discourse context'' so that ``such a participant tends to assume the minimum force of reference, and in some languages this is zero expressed''.

%The Gyeli text corpus shows clearly that the most common case involves the occurrence of the SCOP rather than its omission. Table \ref{Tab:SCOPdistr} summarizes the variation in the presence of the SCOP. It shows that in 412 instances, out of a total of 472,\footnote{Identificational markers agreeing with the subject and constituting the predicate at the same time were not counted since they do not classify as SCOPs. Also special constructions of quotative indexes where the SCOP is present, but no verb, were not taken into consideration. A third case that was excluded concerns imperatives since imperatives never take a SCOP and thus do not show any variation.}
% the SCOP is present. This number includes both cases where the SCOP is the subject of the clause and cases where the SCOP is preceded by more noun phrase material such as nouns, noun + possessive constructions, and so forth.



%\begin{table}[!h]
%\centering
%\begin{tabular}{l|rr}
% \midrule
%SCOP present         & 412 & (87.3\%) \\
%Assimilation             & 6     & (1.3\%)    \\
%Omission with NP    & 47   & (9.9\%) \\
%Subject ellipsis       &  7      & (1.5\%) \\
%(Imperatives)            & 18     & (3.7\%)      \\
% \midrule
%Total                    & 472   \\
% \midrule
%\end{tabular}
%\caption{Variation of SCOP presence in the text corpus}
%\label{Tab:SCOPdistr}
%\end{table}

%While SCOP assimilation as discussed in \sectref{sec:SCOPAs} and complete subject ellipsis are rather rare, the omission of the SCOP when a noun phrase is present occurs in 9.9\% of the cases. Imperatives are listed in the table as well because their category is characterized by the absence of a SCOP.










\subsection{Verbal particles}
\label{sec:VParticle}

There are two other elements that appear in the Gyeli verbal complex, namely particles that follow the finite verb form. This includes the \textsc{absolute completive} marker {\itshape mɔ̀} and the verbal plural marker {\itshape nga}.

\subsubsection{\textsc{Absolute completive} {\itshape mɔ̀}}
\label{sec:mo}

The \textsc{absolute completive} marker, glossed as COMPL, has two variants, namely a postverbal particle {\itshape mɔ̀}, as in (\ref{absmo1a}), and a long nasalized vowel with a falling HL tone (\ref{absmo1b}). The latter is said to be more typical Gyeli, but {\itshape mɔ̀} is also productively used.\footnote{It can be excluded that {\itshape mɔ̀} is a loan form from Mabi since the cognate form in Mabi is {\itshape mà}.}


\begin{exe} 
\ex\label{absmo1}
\begin{xlist} 
\ex\label{absmo1a}
  \glll    bà kwɛ̀lɔ́ mɔ̀ yɔ̂  \\
           ba kwɛ̀lɔ-H mɔ̀  y-ɔ̂ \\
             2.PST1  cut-R COMPL 7-NSBJ  \\
    \trans `They have (already) cut it.'
\ex\label{absmo1b}
  \glll    bà kwɛ̀lɔ̃́ɔ̃̀ yɔ̂ \\
          ba kwɛ̀lɔ̃́ɔ̃̀ y-ɔ̂   \\
             2.PST1 cut:COMPL.R 7-NSBJ    \\
    \trans `They have (already) cut it.'
\end{xlist}
\end{exe}

\noindent The variant with the final lengthened and nasalized vowel is the contracted form of {\itshape mɔ̀}. The segmental nasal has been deleted, but nasality survived on the lengthened vowel. Also, the tonal pattern of the realis marking H plus the L tone {\itshape mɔ̀} is maintained.

While there are some verbs, as in (\ref{absmo1}), which can take both the {\itshape mɔ̀} form and the contracted form, other verbs can only take one or the other. {\itshape lámbɔ} `trap', for instance, can only take the contracted form as in (\ref{absmo2a}), while the non-contracted form in (\ref{absmo2b}) is judged as ungrammatical. It seems to be lexically determined whether a verb takes one or the other or both forms. 

\begin{exe} 
\ex\label{absmo2}
\begin{xlist} 
\ex\label{absmo2a}
  \glll    mɛ̀ lámbɔ̃́ɔ̃̀ kù  \\
           mɛ lámbɔ̃́ɔ̃̀ kù  \\
             1\textsc{sg}  trap.R.COMPL $\emptyset$1.rat  \\
    \trans `I have (already) trapped the rat.'
\ex\label{absmo2b}
  \glll    *mɛ̀ lámbɔ́ mɔ̀ kù \\
          mɛ̀ lámbɔ-H mɔ̀ kù \\
             1\textsc{sg} trap-R COMPL $\emptyset$1.rat    \\
    \trans `I have (already) trapped the rat.'
\end{xlist}
\end{exe}

\noindent There is a tendency for semi-auxiliaries, such as {\itshape kɛ̀} `go' and {\itshape sílɛ} `finish', to only occur with  the contracted \textsc{absolute completive} form, while {\itshape dyúwɔ} `hear' has only been observed to occur with the full form {\itshape mɔ̀}.

I consider {\itshape mɔ̀} a free morpheme rather than a verbal suffix since  tonal inflection of past tense and/or realis mood (\sectref{sec:SimpPred}) through the grammatical H tone happens on the preceding verb. If {\itshape mɔ̀} was a suffix, it would be the suffix (and the preceding toneless verbal derivation morphemes) that would take the grammatical H tone in non final position. This, however, is not the case, as (\ref{absmo1c}) shows.

\begin{exe}
\ex\label{absmo1c}
  \glll    mɛ̀ lùng{\bfseries á} {\bfseries mɔ̀} bvùbvù\\
           mɛ lùnga-H mɔ̀  bvùbvù \\
             1\textsc{sg}  grow-R COMPL  lots \\
    \trans `I have (already) grown lots.'
\end{exe}

The is no other element that can occur between the verb and the verbal particle.  With verbs that require the comitative marker {\itshape nà} (\sectref{sec:MainVerbs}), for instance, {\itshape nà} follows the verbal particle, as in (\ref{absmo7}).

\begin{exe}
\ex\label{absmo7}
  \glll     nà wɛ̀  làdɔ́ {\bfseries mɔ̀} nà ngã̀ \\
          nà wɛ  làdtɔ-H mɔ̀ nà ngã̀ \\
           Q 2\textsc{sg}.PST1  meet-R COMPL COM $\emptyset$1.healer  \\
    \trans `Have you ever/already met the healer?'
\end{exe}









\subsubsection{Verbal plural particle {\itshape (n)ga}}
\label{sec:nga}

The verbal plural particle {\itshape ga} or its variant {\itshape nga} only occurs in the \textsc{imperative} with the first and second person plural (\sectref{sec:imp}). Just like the \textsc{absolute completive} marker {\itshape mɔ̀}, the verbal plural particle directly follows the finite verb. The two particles never co-occur since they are restricted to different tense-mood categories. (\ref{gaimp1}) shows examples of the second person plural with the variant {\itshape ga}; (\ref{ngaimp1}) includes examples with the variant {\itshape nga}.\footnote{The first person plural, which also involves the use of the verbal plural particle, has the same structure as the second person plural, just with the addition of the first person plural STAMP marker {\itshape yá}, as outlined in \sectref{sec:imp}.}

\begin{exe}
\ex\label{gaimp1}
\begin{xlist}
\ex dê {\bfseries gà} `don't (pl.) eat'
\ex kɛ̂ {\bfseries gà} `don't (pl.) go'
\end{xlist}
\end{exe}


\begin{exe}
\ex\label{ngaimp1}
\begin{xlist}
\ex\label{ngaimp1a} lã̂ {\bfseries ngà} `don't (pl.) count'
\ex\label{ngaimp1b} gyàgâ {\bfseries ngà} `don't (pl.) buy'
\ex sílɛ̂ {\bfseries ngà} `don't (pl.) finish'
\end{xlist}
\end{exe}


The choice and distribution of {\itshape ga} versus {\itshape nga} is not yet fully understood. It seems that {\itshape ga} is the default case that is used with most verbs. {\itshape nga}, in contrast, appears definitely when a monosyllabic verb ends in a nasal vowel as it is the case with {\itshape lã̂}  `read, count', as in (\ref{ngaimp1a}). Nasal vowels are, however, not the only factor that triggers the plural particle to surface with a nasal since {\itshape nga} is also found with bi- and trisyllabic verbs which do not end in a nasal vowel, as shown in (\ref{ngaimp1b}).



There also seems to be a certain degree of free variation since both {\itshape ga} and {\itshape nga} can occur with the same verb form, as in (\ref{IMPga}).

\begin{exe}
\ex\label{IMPga}
\begin{xlist}
\ex\label{IMPgaa} dê {\bfseries gà}! `don't (pl.) eat'
\ex dê {\bfseries ngá} wámíyɛ̀! `don't (pl.) eat fast'
\end{xlist}
\end{exe}

{\itshape ga/nga} always follows the finite verb, as can be seen in the contrast between the positive and the negated cohorative forms in (\ref{IMPga2}). In (\ref{IMPga2a}), the finite verb is {\itshape gyàgâ} `buy' in a simple predicate. In contrast, (\ref{IMPga2b}) shows a complex predicate where the finite verb is the negation auxiliary {\itshape tí}. The verbal plural particle follows the auxiliary, preceding the lexical verb.

\begin{exe} 
\ex\label{IMPga2}
\begin{xlist}
\ex\label{IMPga2a}
  \glll  yá {\bfseries gyàgâ} {\bfseries ngá} bèkálàdè \\
        yá gyàgâ nga bèkálàdè \\
           1\textsc{pl}-PRES buy.IMP PL be8-book  \\
    \trans `Let's buy book!'
\ex\label{IMPga2b}
  \glll  yá {\bfseries tí} {\bfseries ngá} gyàgà békálàdè \\
        ya-H tí nga gyàga H-be-kálàdè \\
           1\textsc{pl}-PRES NEG.R PL buy OBJ.LINk-be8-book \\
    \trans `Let's not buy books!'
\end{xlist}
\end{exe}

I consider {\itshape ga/nga} as a particle rather than a suffix that is bound to the finite verb in an imperative construction. If the particle was a suffix, one would expect it to take the HL tone that is characteristic of the \textsc{imperative} category. Instead, the particle is underlyingly toneless, behaving like toneless CV- noun class prefixes. Phrase finally, {\itshape ga/nga} surfaces L, as shown in (\ref{gaimp1}) and (\ref{ngaimp1}). If a nominal object follows, however,  {\itshape nga} `steals' the object linking H tone which would otherwise surface on the noun class prefix in (\ref{IMPga2a}). In that case, {\itshape be-kálàdè} surfaces with a L tone on the prefix. 
The same is true when the particle is followed by {\itshape wámíyɛ̀} `quickly', as in (\ref{ngapos1}).

\begin{exe} 
\ex\label{ngapos1}
  \glll tí {\bfseries ngá} dè wámíyɛ̀ \\
       tí nga-H dè wámíyɛ̀ \\
        NEG PL-OBJ.LINK eat fast \\
    \trans `Don't eat fast.'
\end{exe}

\noindent If {\itshape nga} is preceding a non-finite verb in a complex predicate, however, no H tone attaches, as shown in (\ref{ngapos2}).

\begin{exe} 
\ex\label{ngapos2}
  \glll sílɛ̂ {\bfseries ngà} nyî vâ \\
       sílɛ̂ nga nyî vâ \\
        finish.IMP PL enter here \\
    \trans `Enter all here.'
\end{exe}

\noindent The H tone on {\itshape nga} in (\ref{IMPga2b}) is therefore not from the object linking H tone, but spread from the H tone on the preceding auxiliary {\itshape tí}. The object linking H tone in this case attaches to the prefix of the nominal object.



%***tone pattern: underlyingly toneless, steals OBJ.LINK H tone from following object:
%The tonal difference in (\ref{ti1}) might seem to be explicable as either high tone spreading from the preceding {\itshape tí}, or lack thereof in (\ref{ti1b}) with {\itshape dê}, or as a phenomenon tied to the phrase-final or non-phrase final position of the plural marker. (\ref{ti2}) shows, however, that this is not the case  













\section{Adpositions} 
\label{sec:Adpos}

\citet[81-82]{dryer2007b} explains that, ``[p]rototypical instances of adpositions are words that combine with noun phrases and that indicate the semantic relationship of that noun phrase to the verb''. In Gyeli, they can be formally distinguished from elements of the noun phrase (\sectref{sec:NAdjuncts}) in that the use of adpositions is generally restricted to oblique noun phrases, excluding argument noun phrases from hosting adpositions.
Gyeli has both prepositions (\sectref{sec:PREP}) and postpositions (\sectref{sec:POST}).   






\subsection{Prepositions} 
\label{sec:PREP}

Gyeli has a limited set of prepositions, including only one locative preposition {\itshape ɛ́}. Also the comitative marker {\itshape nà}, {\itshape tí} `without', and the associative marker {\itshape bà} fall into this category.


\subsubsection{Locative marker {\itshape ɛ́}}
\label{sec:LOCe}

The preposition {\itshape ɛ́}\footnote{The corresponding preposition in Mabi is {\itshape ɔ́}.} is most frequently used to accompany a locative adverb as discussed in \sectref{sec:G1ADV} and listed in (\ref{eADV}).

\begin{exe}
\ex\label{eADV}
\begin{xlist}
\ex ɛ́ vâ `here'
\ex ɛ́ wû `there (MID)'
%\ex ɛ́ wɛ̂ `there (MID)'
\ex ɛ́ pɛ̀ `there (DIST)'
\ex ɛ́ bà `to, at'
\end{xlist}
\end{exe}

\noindent Further, the preposition {\itshape ɛ́} can precede a noun in a locative context as in (\ref{eN}).

\begin{exe}
\ex\label{eN}
\begin{xlist}
\ex ɛ́ tísɔ̀nì `in town'
\ex ɛ́ nkɔ̀lɛ́ `on the line'
\end{xlist}
\end{exe}

Semantically, {\itshape ɛ́} is used as a locative preposition when the described location is about any spatial relation except containment. Spatial containment relations are expressed by the postposition {\itshape dé} as discussed in \sectref{sec:LOCde}.

It is possible that {\itshape ɛ́} is also used as a directional, as in (\ref{totown}), which shows two lexical options of saying `I go to town', differing in the noun used for the landmark. As the final vowel of the verb and the locative preposition are identical, it could be the case that the locative is deleted in its surface form.

\begin{exe}
\ex\label{totown}
\begin{xlist}
\ex\label{totown1}
  \glll    mɛ́ kɛ́ mã̂\\
          mɛ́ kɛ̀-H-ɛ́? m-ã̂\\
              1\textsc{sg}.PRES go-R-LOC? ma6.sea \\
    \trans `I go to town\footnote{From the perspective of the village Ngolo, the town Kribi is located towards the sea line. Therefore, speakers most frequently refer to the direction of the sea when they talk about the town.}.'
\ex\label{totown2}
  \glll    mɛ́ kɛ́ tísɔ̀nì\\
         mɛ́ kɛ̀-H-ɛ́ tísɔ̀nì\\
              1\textsc{sg}.PRES go-R-LOC? $\emptyset$7.town \\
    \trans `I go to town.'
\end{xlist}
\end {exe}

In a case such as in (\ref{efields}), it is thus not clear if the H tone on the noun class prefix comes from an underlying locative marker {\itshape ɛ́} or if the noun is treated as an object receiving an object linking H tone (see \sectref{sec:GR} for a discussion of grammatical relations.)

\begin{exe}
\ex\label{efields}
  \glll    mɛ́ kɛ́ mánkɛ̃̂\\
         mɛ́ kɛ̀-H-ɛ́ ma-nkɛ̃̂\\
              1\textsc{sg}.PRES go-R-LOC? ma6-field \\
    \trans `I go to the fields.'
\end {exe}

  %[EXTEND]

%Finally, the locative preposition {\itshape ɛ́} is also used as part of complex interrogative constructions as in {\itshape ɛ́ ná} `how' and {\itshape ɛ́ vɛ́} `where' which are discussed in detail in \sectref{sec:INTERRC}.



\subsubsection{Comitative marker {\itshape nà}}
\label{sec:COM}

The comitative marker {\itshape nà} broadly expresses association.  As such, it is often translated as `with', as in (\ref{COM1}).

\begin{exe} 
\ex\label{COM1}
  \glll mùdã̂ kɛ̂ {\bfseries nà} nyɛ̀ mánkɛ̃̂ \\
       m-ùdã̂ kɛ̀-H nà nyɛ̀ H-ma-nkɛ̃̂ \\
         n1-woman go-R COM 1 OBJ.LINK-ma6-field  \\
    \trans `Woman [his wife], go with her to the fields,'
\end{exe}

The comitative marker is used in conjunction with the verb {\itshape bɛ̀} `be' to express possession, as in (\ref{COM2}).

\begin{exe} 
\ex\label{COM2}
  \glll mɛ̀ kí bɛ̀ {\bfseries nà} tsídí \\
       mɛ kí bɛ̀ nà tsídí \\
       1\textsc{sg}.PST1 NEG[Kwasio] be COM $\emptyset$1.meat  \\
    \trans `I didn't have any meat.'
\end{exe}

{\itshape nà} is used in an instrumental sense, as in (\ref{COM3a}).

\begin{exe} 
\ex\label{COM3a}
  \glll mɛ̀ vùlɔ́ pɛ́mbɔ́ {\bfseries nà} ntfúmò \\
       mɛ vùlɔ-H pɛ́mbɔ́ nà ntfúmò  \\
       1\textsc{sg}.PST1 cut-R $\emptyset$7.bread COM $\emptyset$3.knife  \\
    \trans `I cut the bread with a knife.'
\end{exe}

\noindent Extended uses of the instrumental sense are given in (\ref{COM3}) through (\ref{COM5}).

\begin{exe} 
\ex\label{COM3}
  \glll nyɛ̀gà váà nyɛ̀gá tsíyɛ́ sâ {\bfseries nà} màlɛ́ndí \\
         nyɛ-gà váà nyɛ-gá tsíyɛ́ sâ nà ma-lɛ́ndí \\
          1.SBJ-CONTR here 1.SBJ-CONTR live-R only COM 6-palm.tree \\
    \trans `Him here, he lives only from palm trees.'
\end{exe}

\begin{exe} 
\ex\label{COM4}
  \glll  mɛ̀ múà wɛ̀ {\bfseries nà} nzà \\
        mɛ múà wɛ̀ nà nzà \\
          1\textsc{sg} be.almost die COM $\emptyset$9.hunger  \\
    \trans `I'm about to die from hunger.'
\end{exe}

\begin{exe} 
\ex\label{COM5}
  \glll  à múà á kɛ́ jìí dé tù {\bfseries nà} ndzǐ gyâ    \\
          a múà a-H kɛ̀-H jìí dé tù nà ndzǐ gyâ      \\
       1.PST1 be.almost 1-PRES go-R $\emptyset$7.forest LOC inside COM $\emptyset$9.path $\emptyset$7.length \\
    \trans `He was about to go into the forest, on the long path'
\end{exe}

Some verbs lexically the use of {\itshape nà} (\sectref{sec:MainVerbs}), as in (\ref{COM6}) where the combination of {\itshape njì} `come' and the comitative yields the meaning `bring'.

\begin{exe} 
\ex\label{COM6} 
  \glll   ɛ́ pɛ̀ nâ á njíyɛ̀ mɛ̂ {\bfseries nà} yɔ̂ \\
         ɛ́ pɛ̀ nâ a-H njíyɛ mɛ̀ nà y-ɔ̂ \\
         LOC there COMP 1-PRES come.SBJV 1\textsc{sg}.NSBJ COM 7-NSBJ  \\
    \trans `so that she bring me that [food].'
\end{exe}


The comitative also coordinates noun phrases, as in (\ref{COM7}), where {\itshape nà} links a subject pronoun and a noun + possessive pronoun construction.

\begin{exe} 
\ex\label{COM7}
  \glll  bá {\bfseries nà} mùdã̂ wɛ̂ \\
         bá nà m-ùdã̂ w-ɛ̂\\
           2.SBJ COM 1-woman 1-POSS.3\textsc{sg}  \\
    \trans `they and his wife'
\end{exe}

Finally, the comitative marker {\itshape nà} is frequently used in temporal or spatial adjuncts using the non-finite form {\itshape pámɔ} `arrive'.\footnote{Since {\itshape pámɔ} is uninflected and does not carry any person marking, and it seems to be used as a fixed expression, I consider {\itshape nà} as a comitative rather than a verb phrase coordinating conjunction.}

\begin{exe} 
\ex\label{COM8} 
  \glll èhè báà bù mpàgó {\bfseries nà} pámò pɛ̀ Kyíɛ̀ngɛ̀ \\
        èhè báà bù mpàgó nà pámo pɛ̀ Kyíɛ̀ngɛ̀ \\
        EXCL 2.FUT break $\emptyset$3.road COM arrive over.there $\emptyset$7.PN  \\
    \trans `Yes, they will build a road up to Kienge [= river and name for Kribi].'
\end{exe}







\subsubsection {{\itshape tí} `without'}

The preposition {\itshape tí} `without' is the negative counterpart to the comitative {\itshape nà}. It is used, for example, in (\ref{withouttown}).

\begin{exe}
\ex\label{withouttown}
  \glll    mɛ́ kɛ́ tísɔ̀nì tí wɛ̂\\
         mɛ́ kɛ̀-H tísɔ̀nì tí wɛ̂ \\
              1\textsc{sg}.PRES go-R $\emptyset$7.town without 2\textsc{sg}.NSBJ \\
    \trans `I go to town without you.'
\end {exe}


The use of {\itshape tí} as a preposition is a derived function from its primary status as a negative auxiliary (\sectref{sec:NEGti}).


\subsubsection{Associative plural marker {\itshape bà}}
\label{sec:AP}

The associative plural marker {\itshape bà}, which is form identical with the agreement class 2 subject pronoun, indicates a collection of related people, when used with a noun, as in (\ref{AP1}) and (\ref{AP2}).

\begin{exe} 
\ex\label{AP1}
  \glll  mɛ́ kɛ́ jìyɔ̀ vé yá {\bfseries bà} fàmí wã̂ \\
          mɛ-H kɛ̀-H jìyɔ vé ya-H bà fàmí w-ã̂ \\
            1\textsc{sg}-PRES go-R stay where 1\textsc{pl}-PRES AP $\emptyset$1.family 1-POSS.1\textsc{sg} \\
    \trans `Where will I live, we with my family?'
\end{exe}

\noindent (\ref{AP2}) is similar to (\ref{COM7}), but differs in that no comitative marker is used. The tonal pattern of {\itshape ba} also differs: as the associative plural, {\itshape bà} has a L tone, as a subject pronoun, it has a H tone.

\begin{exe} 
\ex\label{AP2}
  \glll bà mùdã̂ wà nû \\
       bà m-ùdã̂ wà nû \\
          AP N1-woman 1:ATT 1.DEM.PROX  \\
    \trans `the people/family of this woman'
\end{exe}

When preceding a pronoun, instead of a noun, the associative plural marker {\itshape bà} expresses directionality towards human entities, as in (\ref{AP3}).\footnote{This is similar to the French use of {\itshape chez} `to' that is used for human goals.}

\begin{exe} 
\ex\label{AP3}
  \glll mɛ́ lɔ́ njì bàgyɛ̃̂ {\bfseries bà} wɛ̂ \\
       mɛ-H lɔ́ njì ba-gyɛ̃̂ bà wɛ̂ \\
       1\textsc{sg}-PRES RETRO come ba2-stranger AP 2\textsc{sg}  \\
    \trans `I just came as a guest to you.'
\end{exe}


Other directionals that  typically require by a preposition  in English, such as `go up', `go down', or `go around', are expressed by verbs in Gyeli, as illustrated in (\ref{directionals}). Therefore, they do not include further adpositions. 

\begin{exe}
\ex\label{directionals}
\begin{xlist}
\ex\label{directionals1}
  \gll    mɛ́ bédégá nkùlɛ́\\
              1\textsc{sg}.PRES ascend $\emptyset$3.hill \\
    \trans `I go up the hill.'
\ex\label{directionals2}
  \gll    mɛ́ sìlégá nkùlɛ́\\
              1\textsc{sg}.PRES descend $\emptyset$3.hill \\
    \trans `I go down the hill.'
\ex\label{directionals3}
  \gll    mɛ́ kɛ́ vyàmbɛ̀lɛ̀ nkùlɛ́\\
              1\textsc{sg}.PRES go surround $\emptyset$3.hill \\
    \trans `I go aound the hill.'
\end{xlist}
\end {exe}



%todo benefactive (*** BEN bà 1.27)


%*** todo for mpá'à 1.152





\subsection{Postpositions} 
\label{sec:POST}

Gyeli has a few postpositions which mostly express location. I distinguish three groups. The first and most frequent category includes {\itshape dé} `in/on' and {\itshape tù} `inside' which can co-occur. The second category comprises simple locative postpositions that cannot combine with any other postposition and that are clearly derived from location nouns. The third group consists of only one temporal postposition {\itshape wɛ̂} which cannot combine with other adpositions either, but that differs from group two postpositions in that it is not derived from nouns.




\subsubsection[Combinable postpositions {\itshape dé} and {\itshape tù}]{Combinable postpositions {\itshape dé} `in/on' and {\itshape tù} `inside'}
\label{sec:LOCde}

The locative postpositions {\itshape dé} `in/on' and {\itshape tù} `inside' generally encode a spatial relation of \textsc{containment}. Most commonly, both postpositions co-occur where {\itshape dé} directly follows the noun and {\itshape tù} follows {\itshape dé}, as shown in (\ref{deN}).\footnote{It is possible that {\itshape dè} was diachronically a preposition to {\itshape tù} `inside', which may have been a noun originally.}

\begin{exe}
\ex\label{deN}
\begin{xlist}
\ex \gll ndáwɔ̀ dé tù  \\
      $\emptyset$9.house LOC inside \\
\trans`in the house'
\ex  \gll mìnkĩ́ dé tù  \\
 $\emptyset$1.pot LOC inside \\
\trans `in the pot'
\end{xlist}
\end{exe}

Examples of the co-occurrence of both postpositions from natural text are provided in (\ref{detu1}) and (\ref{detu2}).

\begin{exe} 
\ex\label{detu1}
  \glll  bɔ́nɛ́gá báà ná jìí {\bfseries dé} {\bfseries tù} \\
        b-ɔ́nɛ́gá báà ná jìí dé tù \\
          2-other 2.COP still $\emptyset$7.forest LOC inside  \\
    \trans `The others are still in the forest.'
\end{exe}

\begin{exe} 
\ex\label{detu2}
  \glll   àà ndáwɔ̀ {\bfseries dé} {\bfseries tù} nyɛ̀ mɛ́dɛ́ támé   \\
           àà ndáwɔ̀ dé tù nyɛ mɛ́dɛ́ támé      \\
          1.COP $\emptyset$9.house LOC inside 1 self alone \\
    \trans `He is in the house all by himself.'
\end{exe}

Both postpositions can, however, occur without the other one while maintaining their meaning of spatial \textsc{containment}, as in (\ref{deNde}) and (\ref{deNtu}). The exact semantic difference between constructions with both postpositions, only {\itshape dé}, or only {\itshape tù} is not clear at this point and likely requires a systematic study of postposition combinations with a large set of different nouns as spatial reference points.  Generally, it seems, however, that the component of \textsc{containment} is stronger with {\itshape tù} `inside'.

%\vspace{.3cm}

\begin{minipage}{0.5\textwidth}
\begin{exe}
\ex\label{deNde}
\begin{xlist}
\ex ndáwɔ̀ dé \\ `in the house'
\ex mìnkĩ́ dé \\ `in the pot'
\end{xlist}
\end{exe}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{exe}
\ex\label{deNtu}
\begin{xlist}
\ex ndáwɔ̀ tù \\ `inside the house'
\ex mìnkĩ́ tù  \\ `inside the pot'
\end{xlist}
\end{exe}
\end{minipage}

%\vspace{.3cm}

In contrast to {\itshape tù} `inside', {\itshape dé} can also describe spatial relations of \textsc{contact} as in (\ref{decontact}).

\begin{exe}
\ex\label{decontact}
  \gll   nsɔ̃́ wúù wɛ̀ nyúlɛ̀ dé \\
           $\emptyset$3.worm 3.\textsc{cop} 2\textsc{s}   $\emptyset$9.body \textsc{loc}  \\
    \trans `The worm is on your body.'
\end {exe}

\noindent I therefore gloss {\itshape dé} more generally as\textsc{loc} while {\itshape tù} has the more specific meaning `inside'. {\itshape dé} as a locative postposition is not only formally but also semantically distinct from the locative preposition {\itshape ɛ́} which I also gloss as LOC, but which lacks the connotation of \textsc{containment}. Cases of {\itshape dé} as encoding\textsc{contact}T rather than\textsc{containment} may have some semantic similarity with the locative preposition {\itshape ɛ́} in \sectref{sec:LOCe}, although {\itshape ɛ́} seems to mark close proximity rather than contact.

Examples of the locative postposition {\itshape dé} only that come from natural text are given in (\ref{deonly1}) through (\ref{deonly3}).

\begin{exe} 
\ex\label{deonly1} 
  \glll mbúmbù lèbvúú léè nlémò {\bfseries dé} \\
         mbúmbù le-bvúú léè nlémò dé \\
        N1.namesake le5-anger 5.COP $\emptyset$3.heart LOC  \\
    \trans `The namesake has anger in his heart (he is angry).'
\end{exe}

\noindent As (\ref{deonly2}) shows, {\itshape dé} can also be used to indicate directionality rather than location.

\begin{exe} 
\ex\label{deonly2}
  \glll Nzàmbí màbɔ́ɔ̀ nkwɛ́ɛ̀ {\bfseries dé} nâ vɔ́sì \\
        Nzàmbí ma-bɔ́ɔ̀ nkwɛ́ɛ̀ dé nâ vɔ́sì \\
          $\emptyset$1.PN ma6-bread.fruit $\emptyset$3.basket LOC COMP IDEO:pouring\\
    \trans `Nzambi pours the bread fruits into the basket.'
\end{exe}

\noindent The same is true for figurative directionality with the verb {\itshape vìdɛga dé} `turn into' in (\ref{deonly3}).

\begin{exe} 
\ex\label{deonly3}
  \glll mìntángánɛ́ mí múà vìdɛ̀gà {\bfseries dé} \\
       mi-ntángánɛ́ mi-H múà vìdɛga dé \\
       mi4-white.person 4-PRES be.almost turn LOC  \\
    \trans `They are about to turn into white people.'
\end{exe}


Examples of the sole use of {\itshape tù} `inside' as postposition in natural text is less frequent, but attested as in (\ref{tuonly}). 

\begin{exe} 
\ex\label{tuonly}
  \glll bùdì bà sílɛ̃́ɛ̃̀ mɛ̂ wɛ̀ ndáwɔ̀ {\bfseries tù} vâ \\
        b-ùdì ba sílɛ̃́ɛ̃̀ mɛ̂ wɛ̀ ndáwɔ̀ tù vâ \\
       ba2-person 2.PST1 finish.COMPL 1\textsc{sg}.NSBJ die $\emptyset$9.house inside here  \\
    \trans `The people have all died here inside the house.'
\end{exe} 




\subsubsection{Simple locative postpositions}
\label{sec:LOCgen}

Some of the locative nouns described in \sectref{sec:NomLOC} can also be used as locative postpositions. They behave like the postposition {\itshape dé} as explained in  \sectref{sec:LOCde}, but differ in their degree of grammaticalization.  In contrast to the locative postposition {\itshape dè}, these other locative postpositions are clearly used as nouns and as such their meaning is obvious. (\ref{Post}) lists the various nouns that can be also used as postpositions.

\begin{exe}
\ex\label{Post}
\begin{xlist}
\ex ndáwɔ̀ {\bfseries dyúwɔ̀} `on top/over the house' <  dyúwɔ̀ `top'
\ex ndáwɔ̀ {\bfseries sí} `under the house' <  sí `ground'
\ex ndáwɔ̀ {\bfseries písɛ̀} `behind the basket' <  písɛ̀ `back'
\ex ndáwɔ̀ {\bfseries sɔ̀} `in front of the house' <  sɔ́ `front' 
\ex ma-ndáwɔ̀ {\bfseries tɛ́mɔ́} `between the houses' < tɛ́mɔ́ `middle'
\end{xlist}
\end {exe}





\subsubsection{Temporal postposition {\itshape wɛ̂ }}
\label{sec:POSTwe}

Gyeli has one temporal postposition {\itshape wɛ̂} that follows time denoting nouns as in (\ref{postwe}).

\begin{exe}
\ex\label{postwe}
\begin{xlist}
\ex mɛ́nɔ́ wɛ̂  `in the morning'
\ex dùwɔ̀ wɛ̂  `in the day'
\ex kùgúù wɛ̂  `in the evening'
\ex bvùlɛ́ wɛ̂  `at night'
\end{xlist}
\end{exe}


\noindent A natural text example is given in (\ref{temppost}). 

\begin{exe} 
\ex\label{temppost} 
  \glll   yá sàgà mɛ́nɔ́ {\bfseries wɛ̂} nyɛ́ɛ̀ mápà má njìbù má bwámɔ́ ndáwɔ̀ dé tù \\
          ya-H sàga mɛ́nɔ́ wɛ̂ nyɛ́ɛ̀ H-ma-pà má njìbù ma-H bwámɔ-H ndáwɔ̀ dé tù     \\
        1\textsc{pl}-PRES be.surprised 7$\emptyset$.morning in see.SBJV OBJ.LINK-ma6-paw 6:ATT $\emptyset$1. antelope 6-PRES come.out-R $\emptyset$9.house LOC inside  \\
    \trans `We are surprised in the morning to see traces of an antelope which come out of the house.'
\end{exe}





\section{Conjunctions}
\label{sec:CONJ}

Conjunctions are used in complex clauses (\chapref{sec:CC}) and link phrases and clauses, resulting in coordination (\sectref{sec:Coord}) or subordination (\sectref{sec:Sub}). Conjunctions that link the same type of clause or phrase are referred to as `coordinatiors'. Subordinating conjunctions include complementizers and adverbializers.

\subsection{Coordinators}
\label{sec:COORD}

Gyeli has three coordinators, as shown in (\ref{Coordi}).

\begin{exe}
\ex\label{Coordi}
\begin{xlist}
\ex\label{Coordi} nà `and' (\sectref{sec:Conjunction})
\ex\label{Coordi} kânà/nânà `or' (\sectref{sec:Disjunction})
\ex\label{Coordi} ndí `but' (\sectref{sec:AdverseCoord})
\end{xlist}
\end{exe}

\noindent More details and examples are given in the respective chapters.




\subsection{Subordinators}
\label{sec:COMP}

The most frequent subordinator in Gyeli is the complementizer {\itshape nâ} that links a complement clause to the main clause, as discussed in \sectref{sec:CompC}. The subordinator {\itshape ká} `if' introduces conditional clauses, which are more free with respect to their position before or after the main clause, as discussed in \sectref{sec:Cond}.


\section{Minor word classes}
\label{sec:Others}

This last section includes all minor parts of speech, ranging from connectors in non-verbal sentences--copulas and the identificational marker {\itshape wɛ́}--to question markers, and extrasentential elements.

\subsection{Copulas}
\label{sec:Copulas}

A copula links two elements, namely the subject and the predicate, in a non-verbal clause (\sectref{sec:COP}). In Gyeli, the copula agrees with the head noun. The agreeing copula is form identical to the STAMP marker (\sectref{sec:SCOP}) and takes a long vowel with a HL default tonal pattern. Exceptional person categories, including the first and second person singular and the agreement class 1 copula, have a long vowel with a L tone.

\begin{table} 
\centering
\begin{tabular}{l|l|l}
 \midrule
 & Singular & Plural \\

Speech Act Participants & 1\textsc{sg} {\itshape mɛ̀ɛ̀} & 1\textsc{pl} {\itshape yáà} \\
 & 2\textsc{sg} {\itshape wɛ̀ɛ̀} & 2\textsc{pl} {\itshape bwáà} \\
 \midrule
Non-Speech Act Participants & cl.1 {\itshape àà/nùù} & cl.2 {\itshape báà} \\
(3\textsuperscript{rd} person) & cl.3 {\itshape wúù} & cl.4  {\itshape míì} \\
& cl.5 {\itshape léè} & cl.6 {\itshape máà} \\
 & cl.7 {\itshape yíì} & cl.8 {\itshape béè} \\
&  cl.9 {\itshape nyì} & \\
 \midrule
\end{tabular}
\caption{Copula forms across agreement classes}
\label{Tab:COP}
\end{table}




\subsection{Identificational marker {\itshape wɛ́}}
\label{sec:IDwe}

Another element used in non-verbal clauses is the identificational marker {\itshape wɛ́} which links a subject to a demonstrative or deictic adverb. Unlike the copula, however, a non-verbal clause with {\itshape wɛ́} can also just consist of a subject and the identificational marker. Both constructions are described with examples in \sectref{sec:ID}.


\subsection{Question markers}
\label{sec:Qna}

Gyeli has two question markers: {\itshape nà} and {\itshape nànâ}. The first generally signals a question, the second is emphatic and is thus pragmatically marked. Examples and a more detailed discussion are provided in \sectref{sec:Questions}.


\subsection{Sentential modifiers}
\label{sec:SentMod}

Sentential modifiers include {\itshape ná} `again, still', {\itshape lìí} `not yet', and {\itshape ndáà} `also'. They are distinguished from adverbs (\sectref{sec:ADV}) in that sentential modifiers usually occur immediately after the finite verb form, which is not possible for adverbs in complex predicate constructions, as adverbs rather appear at both the left and the right edge of a sentence. \sectref{sec:SentMod} gives more information about the function of sentential modifiers within a clause.


\subsection{Extrasentential elements}
\label{sec:Extrasent}

The Gyeli corpus contains a number of extrasentential elements. I roughly distinguish interjections from exclamations. Interjections are words that do not relate to the rest of the sentence in a grammatical way. They are, however, lexical words. Exclamations, in contrast, are not considered as lexical words, but rather sounds that convey attitudes and emotions.

\subsubsection{Interjections}
\label{sec:Interjections}

Most (recognizable) interjections used in Gyeli are actually loan words from French.\footnote{It is possible that I classify some local interjections with exclamations since their meaning is generally hard to describe for speakers and the difference between a lexical word and an emotion encoding sound is possibly not always very clear.} Interjections have a discourse structuring function and often appear at the beginning of an intonation phrase, as in (\ref{INTERJ1}).

\begin{exe} 
\ex\label{INTERJ1} 
  \glll {\bfseries donc} sí nyã̂ nyíì búùlɛ̀ yá Ngɔ̀lɔ́ \\
        donc sí ny-ã̂ nyíì búùlɛ̀ yá Ngɔ̀lɔ́ \\
        so[French] $\emptyset$9.ground 9-POSS.1\textsc{sg} 9:COP $\emptyset$7.old.camp 7:ATT $\emptyset$3.PN  \\
    \trans `So, my land is the old settlement of Ngolo.'
\end{exe}

Pragmatically, interjections are also used to reinforce common ground, as in (\ref{INTERJ2}) where the speaker acknowledges that he and the addressee are on the same page. 

\begin{exe} 
\ex\label{INTERJ2}
  \glll  {\bfseries voilà} wɛ̀ɛ̀ njǐ nà njǐ wɛ̀ɛ̀ njǐ nà njǐ\\
        voilà wɛ̀ɛ̀ njǐ nà njǐ wɛ̀ɛ̀ njǐ nà njǐ \\
          ok[French] 2\textsc{sg}.COP $\emptyset$9.path CONJ $\emptyset$9.path 2\textsc{sg}.COP $\emptyset$9.path CONJ $\emptyset$9.path  \\
    \trans `Exactly, you are on the right track.'
\end{exe}

Even though the Bagyeli of Ngolo report that their French is, if at all, very limited, they are all be able to use these French interjections, as well as {\itshape allez} `come on'  and {\itshape alors} `so, then'.

\subsubsection{Exclamations}
\label{sec:EXCL}

Exclamations reveal the speaker's attitude and emotion towards a situation, usually encoding agreement, disagreement, surprise, or getting the addressee's attention. All exclamations can be manipulated in terms of their length. A longer sound (and often increased volume) correlates with higher emotional intensity.

A widely used exclamation in the area (not only in Gyeli) is {\itshape  ɛ́ɛ́kɛ̀} which signals general surprise about either a positive or negative event. In (\ref{eeke1}), {\itshape ɛ́ɛ́kɛ̀} is a reaction to a character in a story who wants to eat a child. The exclamation refers potentially to both the narrator's attitude the reaction of the woman in the story whose child will be eaten.

\begin{exe} 
\ex\label{eeke1}
  \glll  {\bfseries ɛ́ɛ́kɛ̀} mùdã̂ à gyɛ̃́ɛ̃̀ \\
        ɛ́ɛ́kɛ̀ m-ùdã̂ a gyɛ̃́ɛ̃̀ \\
          EXCL N1-woman 1.PST1 cry.COMPL \\
    \trans `Oh, the woman cried'
\end{exe}


Exclamations are also frequently used in reported discourse, as in (\ref{eeke}).

\begin{exe} 
\ex\label{eeke}
  \glll  yɔ́ɔ̀ bá kí nâ {\bfseries ɛ́ɛ́kɛ̀} \\
            yɔ́ɔ̀ ba-H ki-H nâ ɛ́ɛ́kɛ̀ \\
         so 2-PRES say-R COMP EXCL \\
    \trans `So they say that [EXCL of surprise]!'
\end{exe}

Another frequent exclamation is {\itshape áá} or {\itshape áà} or {\itshape àà}. The tonal pattern seems to depend, at least partially,  on the distance between speaker and addressee, with a H tone indicating distance and a L tone proximity. {\itshape áá} has been observed to occur often to introduce a question, as in (\ref{aa1}) and (\ref{aa2}). It seems comparable to the English exclamation `oh!' expressing surprise or desire.

\begin{exe} 
\ex\label{aa1}
  \glll  {\bfseries áá} gyí wɛ́ lɔ́ njì gyɛ́sɔ̀ \\
        áá gyí wɛ-H lɔ́ njì gyɛ́sɔ \\
           EXCL what 2\textsc{sg}-PRES RETRO come look.for  \\
    \trans `Ah, what have you just come to look for?'
\end{exe}

\begin{exe} 
\ex\label{aa2}
  \glll     {\bfseries áá} bíì màndáwɔ̀ má zì, yáà mɔ̂ fúàlà bwɛ̂ lèwùlà lé vɛ́\\
          áá bíì ma-ndáwɔ̀ má zì yáà m-ɔ́ fúala bwɛ̂ le-wùlà lé vɛ́\\
              EXCL 1\textsc{pl}.NSBJ ma6-house 6:ATT $\emptyset$7.tin[Bulu] 1\textsc{pl}.FUT 6-NSBJ end receive le5-hour 5:ATT which \\
    \trans `Ah, us, tin houses, when will we receive them?'
\end{exe}

{\itshape àà} is also used in addressing someone and getting the addressee's attention.

\begin{exe} 
\ex\label{8aa3}
  \glll   mɛ̀ bìyɛ́ làwɔ̀ nâ {\bfseries àà} bwánɔ̀ bã̂ \\
           mɛ bìyɛ-H làwɔ nâ àà b-wánɔ̀ b-ã̂ \\
          1\textsc{sg} in.vain? speak COMP EXCL ba2-child 2-POSS.1\textsc{sg} \\
    \trans `I say in vain: `ah, my children...''
\end{exe}

\noindent The H tone on {áá} in (\ref{aa4}) indicates that mother and father are far away from the speaker.

\begin{exe} 
\ex\label{aa4}
  \glll áá nyáò áá táò \\
       áá nyá-ò áá tá-ò \\
       EXCL N1-mother-VOC EXCL N1-father-VOC  \\
    \trans `Oh mother, oh father!'
\end{exe}

A similar function of attention seeking and address is found with {\itshape ɔ́ɔ́ɔ́} in (\ref{ooo}) and {\itshape ɛ́} in (\ref{eee}), comparable to English `hey!'.

\begin{exe} 
\ex\label{ooo}
  \glll  nyɛ̀ nâ {\bfseries ɔ́ɔ́ɔ́} mùdã̂ \\
          nyɛ nâ ɔ́ɔ́ɔ́ m-ùdã̂ \\
            1 COMP EXCL N1-woman  \\
    \trans `He: `Oh, wife,''
\end{exe}

\begin{exe} 
\ex\label{eee}
  \glll {\bfseries ɛ́} mwánɔ̀ wã̂ dyúwɔ̀ \\
       ɛ́ m-wánɔ̀ w-ã̂ dyúwɔ̀ \\
        EXCL N1-child 1-POSS.1\textsc{sg} on \\
    \trans `Hey, about my child!'
\end{exe}

Exclamations with a clear negative connotation are {\itshape yééé} as a sound of disappreciation and {\itshape kɛ́ɛ́ɛ́} (with varying length). The latter expresses outrage and strong disapproval, as in (\ref{keee1}) where the speaker expresses his indignation after learning that his child had been eaten by his friend.

\begin{exe} 
\ex\label{keee1}
  \glll nyɛ̀ nâ {\bfseries kɛ́ɛ́ɛ́ɛ́} \\
        nyɛ nâ kɛ́ɛ́ɛ́ɛ́ \\
       1 COMP EXCL  \\
    \trans `He [says]: What!'
\end{exe}

This exclamation can also be used less strongly in a pejorative way, as in (\ref{keee2}). Here, {\itshape kɛ́ɛ́ɛ́} shows the belittling attitude of the speaker towards his children.

\begin{exe} 
\ex\label{keee2}
  \glll  nyɛ̀ nâ {\bfseries kɛ́ɛ́ɛ́} bwánɔ̀ bã̂ mɛ̀ sílɛ̃́ɛ̃̀ bɔ̂ dyùù \\
        nyɛ nâ kɛ́ɛ̀ b-wánɔ̀ b-ã̊ mɛ sílɛ̃ɛ̃̀ b-ɔ̂ dyùù \\
        1 COMP EXCL ba2-child 2-POSS.1\textsc{sg} 1\textsc{sg} finish.COMPL 2-NSBJ kill \\
    \trans `He [says]: Ha, my children, I have already killed them all.'
\end{exe}



%wɛ́ɛ̀


